### 2020-07-26 paper.tex
i hate git

\documentclass[11pt]{amsart}
\usepackage{geometry}                % See geometry.pdf to learn the layout options. There are lots.
\geometry{letterpaper}                   % ... or a4paper or a5paper or ... 
%\geometry{landscape}                % Activate for for rotated page geometry
%\usepackage[parfill]{parskip}    % Activate to begin paragraphs with an empty line rather than an indent
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{epstopdf}
\usepackage{amsmath}
\usepackage{tikz}

\DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
\usetikzlibrary{arrows,positioning}
% LB things
\newenvironment{packed_item}{
\begin{itemize}
 \setlength{\itemsep}{0pt}
  \setlength{\parskip}{0pt}
  \setlength{\parsep}{0pt}
}{\end{itemize}}

\newcommand{\Unif}{\text{Unif}}
\newcommand{\Beta}{\text{Beta}}
\newcommand{\Normal}{\text{Normal}}
\newcommand{\Binomial}{\text{Binomial}}
\newcommand{\E}{\mathbb{E}}

\title{Mechanistic Bayesian Forecasts of COVID19}
\author{Graham Gibson}
%\date{}                                           % Activate to display a given date or no date

\begin{document}
\section*{Abstract}
\begin{itemize}
\item Covid has affected XX people
\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
\begin{itemize}
\item Complexities of interventions 
\item Issues with testing and reporting
\end{itemize}
\item We propose a novel forecasting algorithm to overcome 
\begin{itemize}
\item Under-reporting of cases
\item Time-varying interventions
\end{itemize}
\item Bayesian end to end estimation using both cases and deaths in numpyro
\end{itemize}

\maketitle

%\section{}
%\subsection{}

\section{Introduction}

\begin{itemize}
\item Emergence and spread of covid
\item Forecasts useful for public health
\begin{itemize}
\item Resource allocation
\item Intervention Planning
\item Disease burden
\end{itemize}
\item Cite Flu Forecasting
\item Describe COVID-HUB
\begin{itemize}
\item Number of participating teams as evidence of importance
\item Soliciting forecasts for incident and cumulative deaths
\end{itemize}
\item Mechanistic models have been around since K\&K
\item Demonstrated success in modeling infectious disease 
\item Since they were developed around the last pandemic their usefulness as applied forecasting models has gone untested
\item Extensions to basic model needed 
\begin{itemize}
\item Observations on cases and deaths
\item Time-varying detection probability for varied testing
\item Non-parametric model of interventions
\item Full Bayesian fitting due to unidentifiability of parameters given only a time series of cases and deaths.
\end{itemize}
\item MechBayes outperforms in two settings
\begin{itemize}
\item COVID-HUB relative to baseline model 
\item Ablation test relative to basic SEIR model
\end{itemize}
\item We show improvements in MAE and WIS relative to baseline model in both experimental setups

\end{itemize}

\section{Data}

\begin{itemize}
\item We use cases/deaths from JHU
\item Under-reporting issues
\item Batch reporting issues
\item Revision Issues
\item Weekly cycle issues
\item High variability in incident data
\item Ref Figure 1
\end{itemize}

 \begin{figure}
     \centering
     \includegraphics[scale=.1]{data_plot.png}
     \caption{Deaths by state. }
     \label{fig:my_label}
 \end{figure}

 \section{Compartmental Model}
 \begin{itemize}
 \item Basic SEIR model description
 \item Definition of compartmental model (description of diff eq as flow between compartments)
 \item Ref Figure 2
 \item Interpretation of parameters in basic SEIR model 
 \begin{itemize}
 \item Sigma
 \item Gamma
 \item beta
 \end{itemize}

 \end{itemize}

 \begin{figure}

 \begin{center}
\begin{tikzpicture}[node distance=1cm,auto,>=latex',every node/.append style={align=center},int/.style={draw, minimum size=1cm}]
    \node [int] (S)             {$S$};
      \node [int,right=of S] (E)  {$E$};
    \node [int, right=of E] (I) {$I$};
    \node [int, right=of I] (D_1) {$D_1$};
    \node [int, right=of D_1] (D_2) {$D_2$};
     \node [int, below=of I] (R) {$R$};

   % \coordinate[right=of ] (out); %
    \path[->, auto=false] (S) edge node[yshift=12pt] {$\beta(t) I$ \\[.2em]} (E)
                          (E) edge node[yshift=12pt] {$\sigma$       \\[.2em] } (I) 
                           (I) edge node[yshift=12pt] {$\rho\gamma$       \\[.2em] } (D_1)
                           (I) edge[bend right=90] node[xshift=-20pt,yshift=-20pt] {$(1-\rho)\gamma$       \\[.4em] } (R)
                            (D_1) edge node[yshift=12pt] {$\lambda$       \\[.2em] } (D_2)  ;
                          \end{tikzpicture}
\end{center}
      \caption{Comparmental model parameters }
     \label{fig:seird}
 \end{figure}
 \subsection{Observations on cases and deaths}
 \begin{itemize}
 \item Bayesian model has observations on both cases and deaths
 \item We can weight likelihood of one over the other
 \item Explain how cases inform deaths on average 10 days later
 \end{itemize}

 \subsection{Time-varying beta parameter}
 \begin{itemize}
 \item Non-parametrically models time-varying transmissibility through a random walk
 \item Makes forecasts conditional on current level of interventions
 \item Requires no external intervention data to make forecasts 
 \item 
 \end{itemize}

 \subsection{Time-varying detection probability}
  \begin{itemize}
 \item Non-parametrically models time varying testing and overall detection of case issues
 \item Allows for "data dumps"
 \item Logistic random walk 
 \item Fixed detection probability on deaths

  \end{itemize}

\subsection{Seeding Epidemic}

 \begin{itemize}
 \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
  \end{itemize}

\end{document}

### 2020-07-26 analysis.R

@@ -1,0 +1,77 @@
+library(reticulate)
+library(tidyverse) #for all data wrangling
+library(cowplot) #for manuscript ready figures
+library(lme4) #for lmer & glmer models
+library(sjPlot) #for plotting lmer and glmer mods
+library(sjmisc) 
+#library(effects)
+library(sjstats) #use for r2 functions
+
+np <- import("numpy")
+files = list.files("/Users/gcgibson/covid-intervention-analysis/samples/")
+files <- c("US.npz","FL.npz","NY.npz","CA.npz")
+beta <- matrix(NA,ncol=length(files),nrow=137 )
+col_idx <- 1
+states = c()
+for (file in files){
+  if (file !="DC.npz"){
+    print (file)
+    npz1 <- np$load(paste0("/Users/gcgibson/covid-intervention-analysis/samples/",file),allow_pickle = T)
+    beta[,col_idx] = colMeans(npz1$f$mcmc_samples[[1]]$beta)
+    states <-c(states,substr(file,1,2))
+    col_idx <- col_idx + 1
+  }
+}
+
+X <- read.csv(url("https://raw.githubusercontent.com/CEIDatUGA/COVID-19-DATA/master/US/us-state-intervention-data/US_state_intervention_time_series.csv"))
+
+data_for_lmer  = data.frame(R0 = c((beta)), t= rep(1:137,length(states)), state=rep(states,each=137))
+data_for_lmer$state <- as.factor(data_for_lmer$state)
+data_for_lmer$date <- rep(seq(as.Date("2020-03-05"),as.Date("2020-07-19"),"days"),length(states))
+library(openintro)
+
+library(stringr)
+
+bet_t_plot <- ggplot(data_for_lmer, aes(x=date,y=R0,group=state,col=state)) + geom_line() + facet_grid(~state) + theme_bw() + ylab("Beta(t)") + xlab("Date")
+ggsave("/Users/gcgibson/mech_bayes_paper/beta_t_plot.png",bet_t_plot,device="png",width=8,height=4)
+
+det_probs <- matrix(NA,nrow=137,ncol=4)
+counter <- 1
+for (file in c("US.txt","CA.txt","NY.txt","FL.txt")){
+  npz1 <- read.csv(paste0("/Users/gcgibson/covid-intervention-analysis/summary/",file))
+  
+  rw  <-c()
+  for (line in 10:146){
+     row <- (str_split(npz1$mean.......std....median......5.0......95.0......n_eff.....r_hat[line]," "))[[1]]
[…]

### 2020-07-26 data.R

@@ -1,0 +1,13 @@
+library(stringr)
+library(dplyr)
+deaths <- read.csv("/Users/gcgibson/Desktop/truth-Incident Deaths.txt")  # Number of students in bed
+
+
+
+state_deaths <- deaths %>%
+  filter(!str_detect(location_name, "County"))
+
+library(ggplot2)
+
+data_plot <- ggplot(state_deaths,aes(x=date,y=value)) + geom_point() + facet_wrap(~location_name,scales="free")
+ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=20,height=10)

### 2020-07-26 score.R

@@ -1,0 +1,55 @@
+library(zoltr) ## devtools::install_github("reichlab/zoltr")
+library(tidyverse)
+library(lubridate)
+zoltar_connection <- new_connection()
+zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
+## read in fips
+
+## access scores for the COVID-19 Forecasts project
+tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
+
+  mutate(wis = (abs_error+.01*interval_2+.025*interval_5+.05*interval_10+.1*interval_20+.15*interval_30+.2*interval_40+.25*interval_50+.3*interval_60+.35*interval_70+.40*interval_80+.45*interval_90)/12,
+         mae=abs_error)
+tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
+tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
+#tmp_mech_bayes <- tmp_mech_baye
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,state,model) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
+
+
+#tmp_subset$timezero <- as.factor(tmp_subset$timezero)
+tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-19",]
+tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
+library(ggplot2)
+ggplot(tmp_subset,aes(x=timezero,y=wis,col=model)) + geom_point() + facet_wrap(~state) + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+  
+
+
+#by timezero
+
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,model) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
+tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-19",]
+tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
+library(ggplot2)
[…]

### 2020-07-26 paper.tex
i hate git

@@
 \usepackage{tikz}
+\usepackage{subcaption}
 
@@
  \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
+ \item Put priors on initial seed
+  \end{itemize}
+
+  \subsection{Priors}
+
+ \begin{itemize}
+ \item We choose tight priors based on literature
+  \item Fundamental unidentifiability due to renewal equation expression where $I(t)$ is a convolution between $R_t$ and Seiral interval
+  \item Cannot estimate both simultaneously 
+  \item Most flexibility comes in with beta and detection random walk
+  \item Ideal for forecasting, maybe not for inference
+  \end{itemize}
+
+  \subsection{Fitting}
+
+ \begin{itemize}
+ \item Fully Bayesian HMC on parameters
+ \item Why we choose deterministic compartmental model with only uncertainty on parameters and observations
+ \item Estimation in numpyro very fast
+   \end{itemize}
+
+ \section{Experimental Setup}
+
+ \subsection{COVID-HUB}
+ \begin{itemize}
+ \item Real-time forecasting evaluation for 14 weeks starting April 20th 2020. 
+ \item Forecasts submitted every Monday using incident data up until Sunday
+ \item Cumulative forecasts generated by aggregation of incident forecasts
+ \item 1-4 week ahead targets generated from 28 day ahead predictions 
+ \end{itemize}
+
+ \subsection{Ablation Test}
+ \begin{itemize}
+ \item Evaluate a set of nested models
+ \begin{itemize}
+ \item Basic SEIR model with observations only on deaths
+ \item SEIR with joint observations
+ \item SEIR with joint observations and random walk detection probability
+ \end{itemize}
+ \item We omit the test that involves no random walk on beta since a model must take into account interventions.
+
+ \end{itemize}
+
+\section{Results}
+
+\begin{figure}
+  \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.5]{US.png}
+    \caption{US}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.5]{NY.png}
+    \caption{New York}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.5]{FL.png}
+    \caption{Florida}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.5]{CA.png}
+    \caption{California}
+\end{subfigure}%
+
+\caption{Example fit and forecast for four states.}
+\label{fig:fit_and_forecast_results}
+\end{figure}
+
+ \begin{figure}
+     \centering
+     \includegraphics[scale=.1]{beta_t_plot.png}
+     \caption{Deaths by state. }
+     \label{fig:my_label}
+ \end{figure}
+  \begin{figure}
+     \centering
+     \includegraphics[scale=.1]{detection_plot.png}
+     \caption{Deaths by state. }
+     \label{fig:my_label}
+ \end{figure}
+
+\begin{itemize}
+\item MechBayes is uniformly better than baseline model when broken down by region and timezero.
 \end{itemize}

### 2020-07-26 paper.tex
i hate git

@@
 
-\section{Results}
+\section{COVID-Hub Results}
 
@@
 
+\begin{figure}
+  \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_time_zero.png}
+    \caption{MAE by timezero}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_region.png}
+    \caption{MAE by region}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
+    \caption{WIS by timezero}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_region.png}
+    \caption{WIS by region}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_target.png}
+    \caption{WIS by timezero}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_target.png}
+    \caption{WIS by region}
+\end{subfigure}%
+
+\caption{Scores from covid-hub.}
+\label{fig:covidhub}
+\end{figure}
+
 \begin{itemize}
-\item MechBayes is uniformly better than baseline model when broken down by region and timezero.
+\item MechBayes is almost always better than baseline model when broken down by region, target, and timezero after June 1st 2020.
+\item MechBayes has improved over time.
+\begin{itemize}
+\item First two submissions did not have detection probability random walk
+\item Observations were Normally distributed on cumulative deaths
+\item Switched to current model
+\end{itemize}
+\item Naive baseline is hard to beat
+\item MechBayes is biased high. This comes from uncertainty in the random walk leading to potentially huge growth rates. 
+\end{itemize}
+
+   \section{Ablation Results}
+
+   \begin{itemize}
+   \item Model ranking is as follows
+   \begin{itemize}
+   \item MechBayes
+   \item MechBayes-Case Observations
+   \item MechBayes - no detection random walk
+   \end{itemize}
+   \item This ordering is inu
+   \end{itemize}
+
+\section{Discussion}
+\begin{itemize}
+\item Mech Bayes is a fast fully bayesian compartmental model capable of accounting for real-world modeling challenges during a pandemic.
+\item Demonstrated success across regions, targets, and timezeros
+\item Real-time model results show the practice of modeling during an epidemic. Results are improving.
+\item Ablation studies show the results are grounded in real model improvements using historical validation.
+\item Talk about how overall MAE obscures the huge geographic variability. 
+\end{itemize}
+
+\section{Conclusion}
+
+\begin{itemize}
+\item Summarize mech bayes as bayesian compartmental model
+\item Further work: most of the model is about $\beta(t)$. Better methods for modeling it? Spline Etc.
 \end{itemize}

### 2020-07-26 score.R

@@ -13,43 +13,13 @@ tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %
 tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
 tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
 #tmp_mech_bayes <- tmp_mech_baye
-tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,state,model) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
 
 
 #tmp_subset$timezero <- as.factor(tmp_subset$timezero)
-tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-19",]
+tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-04-20",]
 tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
 library(ggplot2)
-ggplot(tmp_subset,aes(x=timezero,y=wis,col=model)) + geom_point() + facet_wrap(~state) + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
-  
+mae_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+ggsave("/Users/ mech_bayes_paper/mae_results_by_time_zero.png",mae_results_by_time_zero,device="png",width=8,height=4)
 
-
-#by timezero
-
-tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,model) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
-tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-19",]
-tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
-library(ggplot2)
[…]

### 2020-07-27 paper.tex
i hate git

@@
 \usepackage{subcaption}
+\usepackage{bm}
 
@@
 
-\begin{itemize}
-\item Emergence and spread of covid
-\item Forecasts useful for public health
-\begin{itemize}
-\item Resource allocation
-\item Intervention Planning
-\item Disease burden
-\end{itemize}
-\item Cite Flu Forecasting
-\item Describe COVID-HUB
-\begin{itemize}
-\item Number of participating teams as evidence of importance
-\item Soliciting forecasts for incident and cumulative deaths
-\end{itemize}
-\item Mechanistic models have been around since K\&K
-\item Demonstrated success in modeling infectious disease 
-\item Since they were developed around the last pandemic their usefulness as applied forecasting models has gone untested
-\item Extensions to basic model needed 
-\begin{itemize}
-\item Observations on cases and deaths
-\item Time-varying detection probability for varied testing
-\item Non-parametric model of interventions
-\item Full Bayesian fitting due to unidentifiability of parameters given only a time series of cases and deaths.
-\end{itemize}
-\item MechBayes outperforms in two settings
-\begin{itemize}
-\item COVID-HUB relative to baseline model 
-\item Ablation test relative to basic SEIR model
-\end{itemize}
-\item We show improvements in MAE and WIS relative to baseline model in both experimental setups
+The emergence of COVID-19 in early 2020 in the United States developed into the largest pandemic the country has seen in over a century.
+As of July 2020, there are over 4 million confirmed infections and over 150,000 deaths due to COVID-19 \cite{dong2020interactive}.
+Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic recession, and political stability.
+Forecasts of incident and cumulative deaths due to COVID may help in resource allocation, vaccine clinical trial planning, and re-opening strategies.
+Along with non-pharamacutical interventions, forecasts are one of the few public health tools available to help fight the pandemic.
+Infectious disease forecasting has been demonstrated to benefit public health decision makers during annual influenza outbreaks \cite{lutz2019applying}.
+However, many forecasts of seasonal disease, such as influenza, often rely on ample historical data to look for patterns that can be projected forward into the future.
+In an emerging pandemic situation, models must be able to fit to limited data.
+The COVID-19 pandemic has seen a resurgence in the use of differential equation models to explain the underlying transmission of a disease through a population.
+First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically the susceptible, exposed, infected, and recovered compartment.
+The model is specified by setting the rates of flow between compartments.
+While these models have been used since their inception in 1918, the COVID pandemic represents a unique opportunity to explore their properties in real-time.
+Emerging pandemics create a unique set of challenges for accurately predicting future deaths.
+These include, but are not limited to, severe under reporting of cases due to asymptomatic transmission, time-varying testing rates, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
+In this work, we introduce a set of extensions to the classic compartmental model that are able to account for the real-world and real-time complexities of infectious disease forecasting during a pandemic.
+We demonstrate the success of the model in both real-time forecast submissions  as well as an ablation test to demonstrate the additional forecast accuracy of our extensions.
+In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
 
-\end{itemize}
+%\begin{itemize}
+%
+%\item Resource allocation
+%\item Intervention Planning
+%\item Disease burden
+%\end{itemize}
+%\item Cite Flu Forecasting
+%\item Describe COVID-HUB
+%\begin{itemize}
+%\item Number of participating teams as evidence of importance
+%\item Soliciting forecasts for incident and cumulative deaths
+%\end{itemize}
+%\item Mechanistic models have been around since K\&K
+%\item Demonstrated success in modeling infectious disease 
+%\item Since they were developed around the last pandemic their usefulness as applied forecasting models has gone untested
+%\item Extensions to basic model needed 
+%\begin{itemize}
+%\item Observations on cases and deaths
+%\item Time-varying detection probability for varied testing
+%\item Non-parametric model of interventions
+%\item Full Bayesian fitting due to unidentifiability of parameters given only a time series of cases and deaths.
+%\end{itemize}
+%\item MechBayes outperforms in two settings
+%\begin{itemize}
+%\item COVID-HUB relative to baseline model 
+%\item Ablation test relative to basic SEIR model
+%\end{itemize}
+%\item We show improvements in MAE and WIS relative to baseline model in both experimental setups
+%
+%\end{itemize}
 
 \section{Data}
+	In this analysis we use confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}.
+This a time series dataset which we truncate to begin March 1st 2020 to April 27th 2020 and captures all 50 states, as well as Guam, Puerto Rico, and American Samoa.
+As noted in \cite{krantz2020level}, COVID-19 cases are often dramatically under-reported, with reporting rates for the U.S.
+estimated at 20-30\% \cite{russel2020using}.
+In addition, there has also been severe temporal variation in the percent of symptomatic cases  \cite{russel2020using}.
+We can also see relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends.
+Example incident death data is show in \ref{fig:data}.
 
-\begin{itemize}
-\item We use cases/deaths from JHU
-\item Under-reporting issues
-\item Batch reporting issues
-\item Revision Issues
-\item Weekly cycle issues
-\item High variability in incident data
-\item Ref Figure 1
-\end{itemize}
+In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the COVID-HUB forecast repository and visualization tool \cite{covid-hub}.
+The COVID-HUB team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths as represented by a set of quantiles defined as
+
+\begin{equation}
+\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}
+\end{equation}
+%\begin{itemize}
+%\item We use cases/deaths from JHU
+%\item Under-reporting issues
+%\item Batch reporting issues
+%\item Revision Issues
+%\item Weekly cycle issues
+%\item High variability in incident data
+%\item Ref Figure 1
+%\end{itemize}
 
@@
      \caption{Deaths by state. }
-     \label{fig:my_label}
+     \label{fig:data}
  \end{figure}
@@
  \section{Compartmental Model}
- \begin{itemize}
- \item Basic SEIR model description
- \item Definition of compartmental model (description of diff eq as flow between compartments)
- \item Ref Figure 2
- \item Interpretation of parameters in basic SEIR model 
- \begin{itemize}
- \item Sigma
- \item Gamma
- \item beta
- \end{itemize}
+% \begin{itemize}
+% \item Basic SEIR model description
+% \item Definition of compartmental model (description of diff eq as flow between compartments)
+% \item Ref Figure 2
+% \item Interpretation of parameters in basic SEIR model 
+% \begin{itemize}
+% \item Sigma
+% \item Gamma
+% \item beta
+% \end{itemize}
+% 
+% \end{itemize}
+% 
+% \begin{figure}
+In a given time-step (e.g. one day), each member of the population of  interest belongs is assumed to be in one of the mutually-and-exhaustive compartments:
+ Susceptible $S$, Exposed but not yet infectious $E$, Infectious $I$, Recovered $R$, hospitalized before death $D_1$ , and finally deceased $D_2$.
+Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter.
+For simplicity, we assume a closed population of size $N$.
+The following parameters govern how members of the population move between compartments:  
+\begin{packed_item}
+\item $\beta(t)$: transmission rate, which we allow to vary by time $t$
+\item $\sigma$: rate of transition from the exposed state $E$ to infectious state $I$; i.e., $1/\sigma$ is the expected duration of the latent period
+\item $\gamma$: rate of transition from  the infectious state $I$ to no longer being infectious; i.e., $1/\gamma$ is the expected duration of the infectious period
+\item $\rho$: fatality rate %(i.e., probability of transitioning from I to $D_1$ instead of I to R)
+\item $\lambda$: rate of transition from $D_1$ to $D_2$ 
+(i.e., the inverse of expected number of days in $D_1$ compartment before death) 
+\end{packed_item}
+For a given time-step $t$, the following differential equations describe the changes in each compartment:
+$$\begin{aligned} 
+\frac{dS}{dt} &= - \beta(t) \frac{SI}{N} \\
+\frac{dE}{dt} &= \beta(t) \cdot \frac{SI}{N} - \sigma E \\
+\frac{dI}{dt} &= \sigma E - \gamma I\\ 
+\frac{dR}{dt} &= (1-\rho)\gamma I \\
+\frac{dD_1}{dt} &= \rho \gamma I - \lambda D_1\\
+\frac{dD_2}{dt} &= \lambda D_2\\
+\frac{dC}{d} &= \sigma E
+\end{aligned} $$
 
- \end{itemize}
+We can write this in a state space representation as follows: \[
+X(t) = (S(t), E(t), I(t), R(t), D_1(t), D_2(t))
+\]
+The update from time $t$ to time $t+1$ can be solved numerically as
+$$\bm{X}(t+1)= \text{RK4}\left(\bm{X}(t),\frac{dX}{dt},\beta(t)\right)$$,
+where RK4 is the Runge-Katta 4th order approximation (see numpyro ode docs) \cite{numpyro}.
 
+\begin{align*}
+ E_0 &\sim \Unif(0, 0.02 N) \\
+I_0 &\sim \Unif(0, 0.02 N) \\
+  D_{1_0} &\sim \Unif(0, 0.02 N) \\
+   D_{2_0} &\sim \Unif(0, 0.02 N) \\   
+   R_{0} &\sim \Unif(0, 0.02 N) \\   
+\end{align*}
+
+This allows us to initialize the process model:
+ $$ X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0) $$
+
+We also place the following priors on the transition parameters: 
+\begin{align*}
+\sigma &\sim \Gamma(5, 5 \hat{d}_E)\\
+\gamma & \sim \Gamma(7, 7 \hat{d}_I) \\
+\beta(0) &\sim \Gamma(1, \hat{d}_I/\hat{R}) \\
+    \rho &\sim \Beta(10, 90)\\ 
+\lambda &\sim \Gamma(10, 100)
+\end{align*}
+ Our prior on rate for leaving the exposed compartment $\sigma$ satisfies $\E[\sigma] = 1/\hat{d}_E$, where $\hat{d}_E$ is an initial guess of the duration of the latent period.
+Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) [cite].
+ %
+ Our prior on the rate for leaving the infectious compartment $\gamma$ satisfies $\E[\gamma] = 1/\hat{d}_I$, where $\hat{d}_I$ is an initial guess for the duration of infectiousness.
+The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset (cite).
+Our prior on the initial transmission rate is derived from the relationship between the basic reproductive number $R(0)$ and the length of the infectious period: $R(0) = \beta(0)/\gamma = \beta(0)\times \hat{d}_I$.
+Therefore, we set our prior on the initial transmission rate to satisfy $\E[\beta(0)] = \hat{R}/\hat{d}_I$ where $\hat{R} = 3.0$ is an initial guess for $R(0)$ and $\hat{d}_I = 2.0$, as described above.
+%%
+Our prior on the fatality rate $\rho$ satisfies $\E[\rho] = 0.1$ with concentration of $100$.
+Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
+
+Informative priors are useful on at least some of the internal parameters, since observed cases alone essentially place only one constraint on $(\sigma, \gamma, \beta)$--- we put more informative priors on $(\sigma, \gamma)$ and allow the model more freedom to estimate $\beta$.
+We have made some but not extensive efforts to keep priors on other parameters less informative.
+Relaxing priors can lead to numerical issues during model estimation for locations with fewer cases or noisier data.
 \begin{figure}
-
  \begin{center}
@@
  \end{figure}
+
  \subsection{Observations on cases and deaths}
@@
   \centering
-    \includegraphics[scale=.1]{mae_results_by_region.png}
-    \caption{MAE by region}
+    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
+    \caption{WIS by timezero}
 \end{subfigure}
@@
   \centering
-    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
-    \caption{WIS by timezero}
+    \includegraphics[scale=.1]{mae_results_by_region.png}
+    \caption{MAE by region}
 \end{subfigure}%
@@
     \caption{WIS by region}
-\end{subfigure}%
+\end{subfigure}
 \begin{subfigure}{.5\textwidth}
@@
     \includegraphics[scale=.1]{mae_results_by_target.png}
-    \caption{WIS by timezero}
+    \caption{WIS by target}
 \end{subfigure}%
@@
     \includegraphics[scale=.1]{wis_results_by_target.png}
-    \caption{WIS by region}
+    \caption{WIS by target}
 \end{subfigure}%
 
-\caption{Scores from covid-hub.}
+\caption{Scores from covid-hub broken down by region, target and timezero.
+Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+}
 \label{fig:covidhub}
@@
 
+\bibliographystyle{unsrt}
+\bibliography{sir}
+
 \end{document}

### 2020-07-28 paper.tex
i hate git

@@
      \centering
-     \includegraphics[scale=.1]{data_plot.png}
-     \caption{Deaths by state. }
+     \includegraphics[scale=.175]{data_plot.png}
+     \caption{Deaths by state for all 50 states and territories.
+Notice the large variability in incident reporting.
+There appears to be a weekly cycle where deaths are under-reported on the weekend.
+We can also see that there are some negative incident deaths, where data are revised to account for deaths that were incorrectly attributed to COVID.
+}
      \label{fig:data}
@@
 
-\begin{align*}
- E_0 &\sim \Unif(0, 0.02 N) \\
-I_0 &\sim \Unif(0, 0.02 N) \\
-  D_{1_0} &\sim \Unif(0, 0.02 N) \\
-   D_{2_0} &\sim \Unif(0, 0.02 N) \\   
-   R_{0} &\sim \Unif(0, 0.02 N) \\   
-\end{align*}
-
-This allows us to initialize the process model:
- $$ X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0) $$
-
-We also place the following priors on the transition parameters: 
-\begin{align*}
-\sigma &\sim \Gamma(5, 5 \hat{d}_E)\\
-\gamma & \sim \Gamma(7, 7 \hat{d}_I) \\
-\beta(0) &\sim \Gamma(1, \hat{d}_I/\hat{R}) \\
-    \rho &\sim \Beta(10, 90)\\ 
-\lambda &\sim \Gamma(10, 100)
-\end{align*}
- Our prior on rate for leaving the exposed compartment $\sigma$ satisfies $\E[\sigma] = 1/\hat{d}_E$, where $\hat{d}_E$ is an initial guess of the duration of the latent period.
-Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) [cite].
- %
- Our prior on the rate for leaving the infectious compartment $\gamma$ satisfies $\E[\gamma] = 1/\hat{d}_I$, where $\hat{d}_I$ is an initial guess for the duration of infectiousness.
-The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset (cite).
-Our prior on the initial transmission rate is derived from the relationship between the basic reproductive number $R(0)$ and the length of the infectious period: $R(0) = \beta(0)/\gamma = \beta(0)\times \hat{d}_I$.
-Therefore, we set our prior on the initial transmission rate to satisfy $\E[\beta(0)] = \hat{R}/\hat{d}_I$ where $\hat{R} = 3.0$ is an initial guess for $R(0)$ and $\hat{d}_I = 2.0$, as described above.
-%%
-Our prior on the fatality rate $\rho$ satisfies $\E[\rho] = 0.1$ with concentration of $100$.
-Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
-
-Informative priors are useful on at least some of the internal parameters, since observed cases alone essentially place only one constraint on $(\sigma, \gamma, \beta)$--- we put more informative priors on $(\sigma, \gamma)$ and allow the model more freedom to estimate $\beta$.
-We have made some but not extensive efforts to keep priors on other parameters less informative.
-Relaxing priors can lead to numerical issues during model estimation for locations with fewer cases or noisier data.
 \begin{figure}
@@
 
- \subsection{Observations on cases and deaths}
- \begin{itemize}
- \item Bayesian model has observations on both cases and deaths
- \item We can weight likelihood of one over the other
- \item Explain how cases inform deaths on average 10 days later
- \end{itemize}
+ \subsection{Time-varying transmission parameter}
 
- \subsection{Time-varying beta parameter}
- \begin{itemize}
- \item Non-parametrically models time-varying transmissibility through a random walk
- \item Makes forecasts conditional on current level of interventions
- \item Requires no external intervention data to make forecasts 
- \item 
- \end{itemize}
+We have seen significant efforts to control the spread of COVID through non-pharmaceutical interventions.
+These include social distancing, lock-downs, and mask wearing.
+To add to the complexity, these interventions have been implemented and repealed at different time points.
+They also face compliance issues.
+In order to capture the aggregate effect of the interventions non-parametrically we choose a flexible model for the time-varying transmission parameter.
+We allow $\beta(t)$ to vary as follows,
 
- \subsection{Time-varying detection probability}
-  \begin{itemize}
- \item Non-parametrically models time varying testing and overall detection of case issues
- \item Allows for "data dumps"
- \item Logistic random walk 
- \item Fixed detection probability on deaths
+$$log(\beta(t)) \sim N(log(\beta(t-1), \sigma_{\beta}^2)$$
 
-  \end{itemize}
+This model assumes that forecasts are made on the current level of interventions.
+
+% \begin{itemize}
+% \item Non-parametrically models time-varying transmissibility through a random walk
+% \item Makes forecasts conditional on current level of interventions
+% \item Requires no external intervention data to make forecasts 
+% \end{itemize}
+% 
+ \subsection{Observation Model}
+
+ The observed data used to fit the model is based on time-series data of confirmed cases $Cases_{t}$ and recorded deaths $Deaths_{t}$. 
+For a given state and day, the change in the  confirmed cases and reported deaths are subset of the  number of infections $I(t)$ and underlying number of deaths $D2(t)$, respectively.
+Therefore, we introduce two additional parameters for the detection probability of cases $p_c$ and the detection probability of deaths $p_d$.
+For both, we set fairly flat priors to reflect these parameters are poorly determined from observed data.
+
+In more detail, $p_c$ is the probability that an infectious person receives a positive test result and is confirmed as case.
+We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$ with concentration 50.
+This means that we expect 30\% of cases to be detected initially.
+However, we also allow this to vary by time.
+
+\begin{equation}
+logit(p_{c,t}) \sim N(logit(p_{c,t-1}), \sigma^2)
+\end{equation}
+
+We also assume the probability that a COVID-19 death is reported $p_d$ has a prior distribution given by  $p_d \sim \Beta(90, 10)$. This prior satisfies $\E[p_d] = 0.9$ with concentration 100.
+
+Using the above SEIR model and these detection probabilities, we can then express the observed numbers of confirmed cases and deaths as follows.
+\begin{equation}
+\text{Cases}_{t} \sim NB(p_{c,t}*I_{t},\sigma_{c}^2)
+\end{equation}
+
+\begin{equation}
+\text{Deaths}_{t} \sim NB(p_d*D_{2_{t}}, \sigma_d^2)
+\end{equation}
+
+%  \begin{itemize}
+% \item Non-parametrically models time varying testing and overall detection of case issues
+% \item Allows for "data dumps"
+% \item Logistic random walk 
+% \item Fixed detection probability on deaths
+%
+%  \end{itemize}
 
 \subsection{Seeding Epidemic}
+Due to the under-reporting of cases, we cannot use the observed data to seed the epidemic.
+We instead allow the model to find the initial state values for all compartments except the number of susceptible people, which we take as the population size of the geographic region minus the sum of the initial values for the other compartments to enforce the constraint that the entire system size sums to the population size.
+We do this by assigning uniform probability to all initial states where the number of people in any given compartment at time zero does not exceed 2\% of the total population.
+This is a highly conservative estimate for the number of infected and exposed people at the start of the epidemic.
 
- \begin{itemize}
- \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
- \item Put priors on initial seed
-  \end{itemize}
+\begin{align*}
+ E_0 &\sim \Unif(0, 0.02 N) \\
+I_0 &\sim \Unif(0, 0.02 N) \\
+  D_{1_0} &\sim \Unif(0, 0.02 N) \\
+   D_{2_0} &\sim \Unif(0, 0.02 N) \\   
+   R_{0} &\sim \Unif(0, 0.02 N) \\   
+\end{align*}
+
+This allows us to initialize the process model:
+ $$ X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0) $$ 
+%
+% \begin{itemize}
+% \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
+% \item Put priors on initial seed
+%  \end{itemize}
 
   \subsection{Priors}
+We also place the following priors on the transition parameters: 
+\begin{align*}
+\sigma &\sim \Gamma(5, 5 \hat{d}_E)\\
+\gamma & \sim \Gamma(7, 7 \hat{d}_I) \\
+\beta(0) &\sim \Gamma(1, \hat{d}_I/\hat{R}) \\
+    \rho &\sim \Beta(10, 90)\\ 
+\lambda &\sim \Gamma(10, 100)
+\end{align*}
+ Our prior on rate for leaving the exposed compartment $\sigma$ satisfies $\E[\sigma] = 1/\hat{d}_E$, where $\hat{d}_E$ is an initial guess of the duration of the latent period.
+Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) [cite].
+ %
+ Our prior on the rate for leaving the infectious compartment $\gamma$ satisfies $\E[\gamma] = 1/\hat{d}_I$, where $\hat{d}_I$ is an initial guess for the duration of infectiousness.
+The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset (cite).
+Our prior on the initial transmission rate is derived from the relationship between the basic reproductive number $R(0)$ and the length of the infectious period: $R(0) = \beta(0)/\gamma = \beta(0)\times \hat{d}_I$.
+Therefore, we set our prior on the initial transmission rate to satisfy $\E[\beta(0)] = \hat{R}/\hat{d}_I$ where $\hat{R} = 3.0$ is an initial guess for $R(0)$ and $\hat{d}_I = 2.0$, as described above.
+%%
+Our prior on the fatality rate $\rho$ satisfies $\E[\rho] = 0.1$ with concentration of $100$.
+Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
 
- \begin{itemize}
- \item We choose tight priors based on literature
-  \item Fundamental unidentifiability due to renewal equation expression where $I(t)$ is a convolution between $R_t$ and Seiral interval
-  \item Cannot estimate both simultaneously 
-  \item Most flexibility comes in with beta and detection random walk
-  \item Ideal for forecasting, maybe not for inference
-  \end{itemize}
+The identifiability of model parameters in compartmental models where the data consists of only a time series of incident cases and deaths presents a problem for uninformative priors.
+Using the renewal style equations, it can be shown that the number of newly infected at time $t$ is a function of the time-varying reproductive number, serial interval and previously reported new infections \cite{wallinga2007generation}.
+This means that a single time series does not contain enough information to separately estimate both the serial interval and the time-varying reproduction number.
+In an SEIR model, the serial interval is distributed exponential with rate parameter $\sigma + \gamma$ \cite{wallinga2007generation}.
+Additionally, the time varying reproduction number is $R_t = \frac{\beta(t)*S(t)}{\gamma}$.
+Therefore, the time series of incident cases is not enough to uniquely identify $\gamma,\sigma,\beta(t)$.
+In order to make the model identifiable, we impose tight priors on the parameters $\sigma$ and $\gamma$ as estimated by the literature, in essence fixing the serial interval and we let $\beta(t)$ vary freely.
+This reflects the underlying biology of the system, since the reciprocal of the sum of $\sigma$ and $\gamma$ may be interpreted as the average time from when an individual becomes infected to when they infect someone else, given that they infect someone else.
+This is a biological property of the disease, rather than $\beta(t)$ which contains both the biological transmissibility as well as the aggregate effects of human behavior through intervention.
+This highlights a fundamental philosophical difference between using compartmental models for forecasting rather than interpreting parameters for epidemiological purposes.
+\textbf{I want to say a little more}
 
+% \begin{itemize}
+% \item We choose tight priors based on literature
+%  \item Fundamental unidentifiability due to renewal equation expression where $I(t)$ is a convolution between $R_t$ and Seiral interval
+%  \item Cannot estimate both simultaneously 
+%  \item Most flexibility comes in with beta and detection random walk
+%  \item Ideal for forecasting, maybe not for inference
+%  \end{itemize}
+%  
   \subsection{Fitting}
+We use the hamiltonian monte carlo algorithm implemented in numpyro (some citation here) to fit the model to data.
+That is, given a time series of confirmed cases ($\text{Cases}_{1:t,j}$) and confirmed deaths ($\text{Deaths}_{1:t,j}$) we use Bayesian inference (via HMC) to obtain
 
- \begin{itemize}
- \item Fully Bayesian HMC on parameters
- \item Why we choose deterministic compartmental model with only uncertainty on parameters and observations
- \item Estimation in numpyro very fast
-   \end{itemize}
+$$P(\bm{\theta} | \text{Cases}_{1:t,j}, \text{Deaths}_{1:t,j}) = \frac{P(\text{Cases}_{1:t,j},\text{Deaths}_{1:t,j}|\bm{\theta}) P(\bm{\theta})}{\int_{\bm{\theta}} P(\text{Cases}_{1:t,j}, \text{Deaths}_{1:t,j} |\bm{\theta})P(\bm{\theta})}}$$
 
+where $\bm{\theta}$ is a vector containing all model parameters.
+
+$$\bm{\theta} = \begin{pmatrix} 
+\beta_{t,j} \\
+\sigma_j \\
+\gamma_j \\
+\rho_j \\
+\lambda_j \\
+p_j \\
+p_{d_j} \\ 
+\text{case detection probability}_{t,j} \\
+\text{death detection probability}_{j} \\
+\text{case over-dispersion}_{j} \\
+\text{death over-dispersion}_{j} 
+\end{pmatrix}
+
+%
+% \begin{itemize}
+% \item Fully Bayesian HMC on parameters
+% \item Why we choose deterministic compartmental model with only uncertainty on parameters and observations
+% \item Estimation in numpyro very fast
+%   \end{itemize}
+%   
+%   
+%   
  \section{Experimental Setup}
@@
   \centering
-    \includegraphics[scale=.5]{US.png}
+    \includegraphics[scale=.3]{US.png}
     \caption{US}
@@
   \centering
-    \includegraphics[scale=.5]{NY.png}
+    \includegraphics[scale=.3]{NY.png}
     \caption{New York}
@@
   \centering
-    \includegraphics[scale=.5]{FL.png}
+    \includegraphics[scale=.3]{FL.png}
     \caption{Florida}
@@
   \centering
-    \includegraphics[scale=.5]{CA.png}
+    \includegraphics[scale=.3]{CA.png}
     \caption{California}
@@
      \includegraphics[scale=.1]{beta_t_plot.png}
-     \caption{Deaths by state. }
+     \caption{Time varying transmissibilty par }
      \label{fig:my_label}
@@
      \includegraphics[scale=.1]{detection_plot.png}
-     \caption{Deaths by state. }
+     \caption{Time-varying detection probability. }
      \label{fig:my_label}
@@
 
+\begin{figure}
+  \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.05]{ablation_1.png}
+    \caption{Ablation results for 1 week ahead.}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.05]{ablation_2.png}
+    \caption{Ablation results for 2 week ahead.}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.05]{ablation_3.png}
+    \caption{Ablation results 3 week ahead.}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.05]{ablation_4.png}
+    \caption{Ablation results 4 week ahead. }
+\end{subfigure}
+
+\caption{Scores from covid-hub broken down by region, target and timezero.
+Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+}
+\label{fig:covidhub}
+\end{figure}
+
    \begin{itemize}

### 2020-07-28 ablation.R

@@ -1,0 +1,7 @@
+ablation_res <- read.csv("/Users/gcgibson/mech_bayes_paper/summary-states.csv")
+ablation_res$mae <- ablation_res$MAE
+library(ggplot2)
+ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
+
+ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
+         summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point() + theme_bw()

### 2020-07-28 data.R

@@ -9,5 +9,5 @@ state_deaths <- deaths %>%
 
 library(ggplot2)
 
-data_plot <- ggplot(state_deaths,aes(x=date,y=value)) + geom_point() + facet_wrap(~location_name,scales="free")
-ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=20,height=10)
+data_plot <- ggplot(state_deaths,aes(x=date,y=value)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free")
+ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=10,height=10)

### 2020-07-28 score.R

@@ -13,7 +13,7 @@ tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %
 tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
 tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
 #tmp_mech_bayes <- tmp_mech_baye
-tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
 
 
 #tmp_subset$timezero <- as.factor(tmp_subset$timezero)
@@ -21,5 +21,25 @@ tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-04-20",]
 tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
 library(ggplot2)
 mae_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
-ggsave("/Users/ mech_bayes_paper/mae_results_by_time_zero.png",mae_results_by_time_zero,device="png",width=8,height=4)
+ggsave("/Users/gcgibson/mech_bayes_paper/mae_results_by_time_zero.png",mae_results_by_time_zero,device="png",width=8,height=4)
+
+
+mae_results_by_region <- ggplot(tmp_subset %>% group_by(unit,model) %>% summarize(mae=mean(mae)),aes(x=unit,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+ggsave("/Users/gcgibson/mech_bayes_paper/mae_results_by_region.png",mae_results_by_region,device="png",width=8,height=4)
+
+mae_results_by_target <- ggplot(tmp_subset %>% group_by(target,model) %>% summarize(mae=mean(mae)),aes(x=target,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
[…]

### 2020-07-28 paper.tex
i hate git

@@
 We use the hamiltonian monte carlo algorithm implemented in numpyro (some citation here) to fit the model to data.
-That is, given a time series of confirmed cases ($\text{Cases}_{1:t,j}$) and confirmed deaths ($\text{Deaths}_{1:t,j}$) we use Bayesian inference (via HMC) to obtain
+That is, given a time series of confirmed cases ($C_{1:t}$) and confirmed deaths ($D_{1:t}$) we use Bayesian inference (via HMC) to obtain
 
-$$P(\bm{\theta} | \text{Cases}_{1:t,j}, \text{Deaths}_{1:t,j}) = \frac{P(\text{Cases}_{1:t,j},\text{Deaths}_{1:t,j}|\bm{\theta}) P(\bm{\theta})}{\int_{\bm{\theta}} P(\text{Cases}_{1:t,j}, \text{Deaths}_{1:t,j} |\bm{\theta})P(\bm{\theta})}}$$
+$$f(\bm{\theta} | C_{1:t},D_{1:t}) \propto f(C_{1:t},D_{1:t} | \bm{\theta})f(\bm{\theta})$$
 
@@
 
-$$\bm{\theta} = \begin{pmatrix} 
-\beta_{t,j} \\
-\sigma_j \\
-\gamma_j \\
-\rho_j \\
-\lambda_j \\
-p_j \\
-p_{d_j} \\ 
-\text{case detection probability}_{t,j} \\
-\text{death detection probability}_{j} \\
-\text{case over-dispersion}_{j} \\
-\text{death over-dispersion}_{j} 
-\end{pmatrix}
+$$\bm{\theta} =[\beta_{t} ,
+\sigma ,
+\gamma ,
+\rho ,
+\lambda ,
+p_{c,t} ,
+p_{d} ,
+\sigma_c^2 ,
+\sigma_d^2 ,
+I_0 ,
+E_0 ,
+D_{1_0} ,
+D_{2_0} ,
+R_0 ]$$
+
+We use the numpyro probabilistic framework to fit the model.
 
@@
  \section{Experimental Setup}
+ To evaluate our model, we examine two different scenarios.
+First, we describe the submission process and infrastructure used for the real time evaluation as part of the COVID-HUB consortium.
+Second, we describe the internal evaluation used to demonstrate our model enhancements improve accuracy over a naive compartmental model.
 
  \subsection{COVID-HUB}
+ The COVID-HUB began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
+We began submitting the first version on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
+The forecasts use daily data up to and including the Sunday before submission the next Monday.
+The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
+Our model went through three distinct iterations as we evaluated performance in real-time.
+
  \begin{itemize}
- \item Real-time forecasting evaluation for 14 weeks starting April 20th 2020. 
- \item Forecasts submitted every Monday using incident data up until Sunday
- \item Cumulative forecasts generated by aggregation of incident forecasts
- \item 1-4 week ahead targets generated from 28 day ahead predictions 
+
+ \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
+Model had normal observation noise (instead of negative binomial) and a non-time varying case detection probability.
+Model was fit to cumulative deaths.
+  \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
+Model had normal observation noise (instead of negative binomial) and time varying case detection probability.
+Model was fit to cumulative deaths.
+    \item \textbf{Version 2: May 24th 2020 through August 1st 2020}. Model had negative binomial observation noise  and time varying case detection probability. Model was fit to incident deaths. 
  \end{itemize}
+
+ These three versions highlight the complexities of forecasting during a real-time pandemic. 
+% \begin{itemize}
+% \item Real-time forecasting evaluation for 14 weeks starting April 20th 2020. 
+% \item Forecasts submitted every Monday using incident data up until Sunday
+% \item Cumulative forecasts generated by aggregation of incident forecasts
+% \item 1-4 week ahead targets generated from 28 day ahead predictions 
+% \end{itemize}
 

### 2020-07-28 paper.tex
update

@@
 \usepackage{bm}
-
+\usepackage{setspace} 
+\doublespacing
 \DeclareGraphicsRule{.tif}{png}{.png}{`convert #1 `dirname #1`/`basename #1 .tif`.png}
@@
 \title{Mechanistic Bayesian Forecasts of COVID19}
-\author{Graham Gibson}
+\author{Graham Gibson, Nicholas Reich, Laura Balzer, Dan Sheldon}
 %\date{}                                           % Activate to display a given date or no date
@@
 \section*{Abstract}
-\begin{itemize}
-\item Covid has affected XX people
-\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
-\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
-\begin{itemize}
-\item Complexities of interventions 
-\item Issues with testing and reporting
-\end{itemize}
-\item We propose a novel forecasting algorithm to overcome 
-\begin{itemize}
-\item Under-reporting of cases
-\item Time-varying interventions
-\end{itemize}
-\item Bayesian end to end estimation using both cases and deaths in numpyro
-\end{itemize}
+
+As of August 1st 2020 the Covid-19 pandemic has caused over 4 million infections and over 150,000 deaths in the United States alone.
+Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials.
+However, data quality during a pandemic suffers from under-reporting, backlog reporting, and limited testing.
+Classical infectious disease differential equation models, such as the susceptible-exposed-infected-recovered model, do not take these data issues into account.
+We propose a novel Bayesian compartmental model that allows for non-parametric modeling of non-pharamacuetical interventions, time-varying testing, and joint observations on case counts and deaths.
+The model has been used to submit forecasts to the Covid-hub repository organized by the Reich Lab.
+We examine the performance relative to the baseline model in addition to performing a nested model comparison of our extensions to the classic models.
+We demonstrate a significant gain in both MAE and probabilistic scoring measures when taking into account data quality issues during a real-time pandemic.
+%\begin{itemize}
+%\item Covid has affected XX people
+%\item Forecasts useful for public health resource planning, intervention planning (vaccine trials), and disease burden
+%\item Basic Mechanistic Models unable to capture complexities of real world disease epidemics due to 
+%\begin{itemize}
+%\item Complexities of interventions 
+%\item Issues with testing and reporting
+%\end{itemize}
+%\item We propose a novel forecasting algorithm to overcome 
+%\begin{itemize}
+%\item Under-reporting of cases
+%\item Time-varying interventions
+%\end{itemize}
+%\item Bayesian end to end estimation using both cases and deaths in numpyro
+%\end{itemize}
 
@@
 Model was fit to cumulative deaths.
-    \item \textbf{Version 2: May 24th 2020 through August 1st 2020}. Model had negative binomial observation noise  and time varying case detection probability. Model was fit to incident deaths. 
+    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}. Model had negative binomial observation noise  and time varying case detection probability. Model was fit to incident deaths. 
  \end{itemize}
@@
  These three versions highlight the complexities of forecasting during a real-time pandemic.
+Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
+We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for the weeks of XXXX.
 % \begin{itemize}
@@
  \subsection{Ablation Test}
+
+ While real-time model evaluation is valuable for understanding evolving model performance, we also perform a retrospective evaluation using three model variants.
+We define the following variations on MechBayes,
+
  \begin{itemize}
- \item Evaluate a set of nested models
- \begin{itemize}
- \item Basic SEIR model with observations only on deaths
- \item SEIR with joint observations
- \item SEIR with joint observations and random walk detection probability
+ \item \textbf{MechBayes Full} Mech Bayes as of model version 3.
+That is, a model using negative binomial observation noise as well as a time-varying random walk, using a joint likelihood over cases and deaths.
+
+ \item \textbf{MechBayes Fixed Case Detection Probability} MechBayes Full with $p_{c,t}$ fixed to $p_c$, that is, removing the time-varying detection probability.
+
+ \item \textbf{MechBayes Death Only} MechBayes Full with observations on cases removed.
  \end{itemize}
- \item We omit the test that involves no random walk on beta since a model must take into account interventions.
 
- \end{itemize}
+ Note that these are nested models, with MechBayes Death Only contained in MechBayes Fixed Case Detection Probability contained in MechBayes Full.
+This allows to perform a nested model comparison using MAE as the scoring metric.
+\textbf{Include something with beta}
 
@@
   \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
     \includegraphics[scale=.1]{beta_t_plot.png}
-     \caption{Time varying transmissibilty par }
-     \label{fig:my_label}
- \end{figure}
-  \begin{figure}
+    \caption{Time varying-beta.}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
   \centering
     \includegraphics[scale=.1]{detection_plot.png}
-     \caption{Time-varying detection probability. }
-     \label{fig:my_label}
+    \caption{Time-varying detection probability}
+\end{subfigure} 
+ \caption{  A) Time varying beta parameter for four example states.
+We can clearly see that the random walk is able to non-parametrically account for the non-pharamceutical interventions.
+B) Time varying detection probability for four example states.
+The time varying detection random walk is able account for the increase in testing.
+}
+\label{fig:model_details}
 \end{figure}
@@
 
-\begin{itemize}
-\item MechBayes is almost always better than baseline model when broken down by region, target, and timezero after June 1st 2020.
-\item MechBayes has improved over time.
-\begin{itemize}
-\item First two submissions did not have detection probability random walk
-\item Observations were Normally distributed on cumulative deaths
-\item Switched to current model
-\end{itemize}
-\item Naive baseline is hard to beat
-\item MechBayes is biased high. This comes from uncertainty in the random walk leading to potentially huge growth rates. 
-\end{itemize}
+We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes version 3 is able to accurately model the observed data in the three example states (NY,CA,FL) as well as the US.
+The model is able to adapt to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves.
+The model is also able capture the uncertainty of the differential equation parameters well enough to produce well calibrated prediction intervals.
+Figure \ref{fig:fit_and_forecast_results}  shows intervals at the 50\%,80\% and 95\% levels.
+However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below.
+Figure \ref{fig:fit_and_forecast_results}  also shows 4 weeks of daily forecasts, along with the daily observed incidence for 1 week out.
+We can see that the predictions are tracking the data even under the weekly reporting cycles.
+
+We can see from Figure \ref{fig:model_details} that MechBayes is able to learn to adapt to the evolving pandemic situation.
+Panel A shows the time-varying transmission $\beta(t)$ for three example states, CA, FL, and NY as well as the U.S.
+We can see that $\beta(0)$ is centered on our prior but as data comes in, the estimate increases.
+This is especially true in NY, where the epidemic took off quickly in March.
+However, the model is then able to adjust to the varying levels of non-pharamceutical interventions present in each of the three states as well as across the U.S.
+This radically reduces the transmissibility parameter by the first week of April 2020.
+This is consistent with a peak in overall deaths two weeks later in mid-April 2020.
+There seems to be some estimation issues at the boundary of $\beta(t)$ where the number of cases does not match the number of deaths, since the cases have not yet converted to deaths.
+This results in the model underestimating transmissibility to reduce the flow through the compartments.
+\textbf{Need to think on this a bit}
+
+We can also see from Figure \ref{fig:model_details}  that MechBayes is able to account for the drastic increase in testing that has occurred across the U.S.
+since March 2020.
+Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in detection probability before climbing again.
+Note that interpreting this strictly as time-varying detection is obscuring the fact that this parameter $p_{t,d}$ can soak-up any excess variation beyond the ability of cases and the case-fatility ratio to explain the number of deaths.
+That is, the time-varying detection probability is able to "de-couple" cases and deaths beyond the case-fataility ratio regardless of the underlying reason (whether that be an increase in testing or shifting age distribution of cases).
+Thus, interpretation of this parameter as a strict mapping to testing is incorrect.
+As a forecasting model, we only need the ability to non-parametrically model deviations and reporting issues in cases.
+
+We next turn to the comparison of MechBayes against the COVID-HUB baseline model.
+This baseline model uses the previous daily incident as the mean forecast for the current daily incidence, along with  bootstrapped prediction intervals from historical changes in daily incidence.
+See covid-hub for more details.
+\cite{covid-hub}.
+
+We begin by breaking down the results by week the forecast was made (timezero) and averaging over both region and target.
+As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 did not outperform the COVID-HUB baseline on MAE or WIS when broken down by timezero.
+However, MechBayes V2 (the introduction of the time-varying detection probability), outperformed the COVID-HUB baseline model on both MAE and WIS when broken down by timezero.
+The same is true for MechBayes V3 with the exception of June 26th, where the COVID-HUB baseline model slightly outperformed MechBayes V3.
+However, as seen in Figure \ref{fig:data}, overall incident deaths were at their lowest across the U.S.
+in late June and remarkably stable, meaning this week was the hardest of the weeks to beat the baseline model.
+
+We also break down the results by geographical region, as seen in Figure \ref{fig:covidhub} (C,D).
+Note that this break-down is averaged over target, but also timezero, meaning that we average over each of the model versions.
+We feel this is important, as it reflects the real-time accuracy of our evolving model efforts, instead of \textbf{cherry-picking} the best model.
+However, we still consistent improvements in MAE under MechBayes when broken down by region, with the exception of New York.
+This is mainly due to MechBayes V1 large MAE for New York in March, where most of the cases (and therefore contribution to MAE) occurred.
+The WIS results are similar.
+
+Finally, we break down the results by target by averaging over region and timezero  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS.
+We can also see that the MAE increase as horizon increase, which is to be expected.
+We can also see that incident MAE is lower than cumulative MAE, which is again to be expected due to the lower absolute numbers of incident deaths.
+Finally, we see the model is slightly better calibrated on incident deaths than cumulative.
+%
+%\begin{itemize}
+%\item MechBayes is almost always better than baseline model when broken down by region, target, and timezero after June 1st 2020.
+%\item MechBayes has improved over time.
+%\begin{itemize}
+%\item First two submissions did not have detection probability random walk
+%\item Observations were Normally distributed on cumulative deaths
+%\item Switched to current model
+%\end{itemize}
+%\item Naive baseline is hard to beat
+%\item MechBayes is biased high. This comes from uncertainty in the random walk leading to potentially huge growth rates. 
+%\end{itemize}   
+%
 

### 2020-07-28 paper.tex
update

@@
 
+We can see from \ref{fig:ablation} that MechBayes Full is consistently better than MechBayes Death Only or MechBayes Fixed Case Detection Probability.
+The only exception seems to be June 22nd 2020 for the two week ahead target.
+Late June 2020 is when many states in the U.S.
+began to see an uptick in cases again, with Texas, Florida, and California seeing their largest total case counts since the beginning of the pandemic.
+Since MechBayes is conditional on the current level of interventions, forecasts made from June 22nd assumed the same level of intervention present on June 22nd.
+During this time, each state was re-opening various establishments, while cases were rising.
+MechBayes Full forecasted an exponential growth in cases, which was reflected in a large over prediction two weeks later on July 4th.
+
+We can also see that the MechBayes Death Only is consistently better than MechBayes Fixed Case Detection Probability.
+This may be evidence that naively including case data, without adjusting for time-varying testing rates, may be worse than not including it at all.
+
 \begin{figure}
@@
   \centering
-    \includegraphics[scale=.05]{ablation_1.png}
+    \includegraphics[scale=.15]{ablation_1.png}
     \caption{Ablation results for 1 week ahead.}
@@
   \centering
-    \includegraphics[scale=.05]{ablation_2.png}
+    \includegraphics[scale=.15]{ablation_2.png}
     \caption{Ablation results for 2 week ahead.}
@@
   \centering
-    \includegraphics[scale=.05]{ablation_3.png}
+    \includegraphics[scale=.15]{ablation_3.png}
     \caption{Ablation results 3 week ahead.}
@@
   \centering
-    \includegraphics[scale=.05]{ablation_4.png}
+    \includegraphics[scale=.15]{ablation_4.png}
     \caption{Ablation results 4 week ahead. }
@@
 }
-\label{fig:covidhub}
+\label{fig:ablation}
 \end{figure}
 
+\section{Discussion}
+
+MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world modeling challenges during a pandemic. Our experiments led us to the following conclusions.
+
 \begin{itemize}
-   \item Model ranking is as follows
-   \begin{itemize}
-   \item MechBayes
-   \item MechBayes-Case Observations
-   \item MechBayes - no detection random walk
-   \end{itemize}
-   \item This ordering is inu
-   \end{itemize}
+\item \textbf{Adding case data when predicting deaths is helpful but only when accounting for data quality issues}.
+Our ablation test (Figure \ref{fig:ablation}) clearly shows that time-varying detection probability is a key feature in the model for reducing MAE of forecasts.
+As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the detection probability at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
+There is remarkable similarity between the increase across geographical locations as well, suggesting (as testing data reflects) an overall increase in the number of detected cases across the U.S.
 
-\section{Discussion}
-\begin{itemize}
-\item Mech Bayes is a fast fully bayesian compartmental model capable of accounting for real-world modeling challenges during a pandemic.
-\item Demonstrated success across regions, targets, and timezeros
-\item Real-time model results show the practice of modeling during an epidemic. Results are improving.
-\item Ablation studies show the results are grounded in real model improvements using historical validation.
-\item Talk about how overall MAE obscures the huge geographic variability. 
+\item \textbf{Allowing for time-varying transmissibility is necessary to non-parametrically capture the effect of non-pharamceutical interventions}.
+Our ablation test explicitly did not include a model that fixed $\beta$ across time.
+This is because the model would not converge without the flexibility to capture interventions.
+While non-parametrically modeling interventions is appealing from a forecasting perspective, it does modify the philosophy behind compartmental modeling.
+By including such a flexible parameter, we may view MechBayes as simply a random-walk model, with a set of epidemiological parameters transforming that random-walk in an almost deterministic way to match both cases and deaths.
+For instance, if the variance of the random walk $\sigma_{\beta}^2$ was allowed to be arbitrarily large, then $\beta(t)$ could vary enough to match the data exactly.
+This would clearly attribute reporting issues as true changes in transmissibility.
+Bypassing the epidemiological interpretation that compartmental models provide.
+
+\item \textbf{Models must evolve in order to be successful in real-time pandemic forecasting}.
+As Figure \ref{fig:covidhub} shows, MechBayes significantly improved over time relative to the baseline model.
+Forecasting during a real-time pandemic is subject to a wide variety of data reporting issues.
+Continuously responding to challenges is key to being able to forecast well.
+
 \end{itemize}
@@
 
-\begin{itemize}
-\item Summarize mech bayes as bayesian compartmental model
-\item Further work: most of the model is about $\beta(t)$. Better methods for modeling it? Spline Etc.
-\end{itemize}
+We have seen that MechBayes is a powerful Bayesian compartmental model that can capture the real-world complexities of forecasting during a pandemic.
+Through internal and external evaluation, we have demonstrated the success of MechBayes in forecasting.
+The model is able to improve over a naive baseline model as well as a naive compartmental model.
+Allowing for time-varying interventions and detection probability is a necessary model component during a real-time pandemic forecasting effort.
+
+While we chose an exponential random walk on $\beta(t)$ there are many other choices for flexible non-parametric modeling of transmissibility.
+Further work might consider a spline model, or a Gaussian process, or semi-parametric models capable of taking intervention dates as covariates.
 

### 2020-07-28 ablation.R

@@ -3,5 +3,35 @@ ablation_res$mae <- ablation_res$MAE
 library(ggplot2)
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
 
-ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-         summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point() + theme_bw()
+ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
+         summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()
+ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png",width=6,height=6)
+
+
+ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(13),]
+
+ablation_2 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
+                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()
+ggsave("/Users/gcgibson/mech_bayes_paper/ablation_2.png",ablation_2,device="png",width=6,height=6)
+
+
+
+
+ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(20),]
+
+ablation_3 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
+                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()
+ggsave("/Users/gcgibson/mech_bayes_paper/ablation_3.png",ablation_3,device="png",width=6,height=6)
+
+
+
+
+ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(27),]
+
+ablation_4 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
+                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()
+ggsave("/Users/gcgibson/mech_bayes_paper/ablation_4.png",ablation_4,device="png",width=6,height=6)
+
+
+
+

### 2020-07-28 score.R

@@ -9,7 +9,7 @@ zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
 tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
 
   mutate(wis = (abs_error+.01*interval_2+.025*interval_5+.05*interval_10+.1*interval_20+.15*interval_30+.2*interval_40+.25*interval_50+.3*interval_60+.35*interval_70+.40*interval_80+.45*interval_90)/12,
-         mae=abs_error)
+         mae=abs_error,error=error)
 tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
 tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
 #tmp_mech_bayes <- tmp_mech_baye

### 2020-07-29 paper.tex
update

@@
 The COVID-19 pandemic has seen a resurgence in the use of differential equation models to explain the underlying transmission of a disease through a population.
-First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically the susceptible, exposed, infected, and recovered compartment.
+First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically the susceptible, exposed, infected, and recovered compartment \cite{kermack1927contribution}.
 The model is specified by setting the rates of flow between compartments.
-While these models have been used since their inception in 1918, the COVID pandemic represents a unique opportunity to explore their properties in real-time.
+While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time.
+Compartmental models have been used to effectively model and forecast disease in non-pandemic situations.
+These include complex compartmental models for Influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, including a retrospective model evaluation of the 1918 Influenza pandemic \cite{hall2007real}.
+Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, Measles \cite{bokler1993chaos}, Dengue \cite{syafruddin2012seir} and a wide variety of other disease.
+
+Compartmental models have also been adopted into a Bayesian framework before, including both stochastic disease dynamics and deterministic dynamics \cite{hotta2010bayesian}\cite{dukic2012tracking}.
+In fact, non-parametric transmissibility was included in a Bayesian SEIR model to study Ebola by Frasso and Lambert \cite{frasso2016bayesian}.
+Time-varying transmissibility has also been studied in the frequentist setting \cite{smirnova2019forecasting}.
+With the outbreak of COVID-19, accounting for testing has become a critical element in using an SEIR model.
+Lopez et al.
+used a fixed detection probability to model undiagnosed individuals in Spain and Italy \cite{lopez2020modified}.
+Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling}\cite{yang2020modified} \cite{bertozzi2020challenges}\cite{prem2020effect}.
+
 Emerging pandemics create a unique set of challenges for accurately predicting future deaths.
@@
 
-In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the COVID-HUB forecast repository and visualization tool \cite{covid-hub}.
-The COVID-HUB team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths as represented by a set of quantiles defined as
+In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the COVID-HUB forecast repository and visualization tool \cite{COVID-HUB}.
+The COVID-HUB team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths since the beginning of the April 2020.
+A COVID-HUB forecast is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
 
@@
 
+Forecasts are due on Monday evenings and therefore use the incident data up until the Sunday before.
+A one-week ahead target corresponds to the following Saturday.
+A two-week ahead corresponds two the next Saturday and so on.
+Forecasts are stored in the generic repository Zoltar, which allows teams to access both their model and a baseline model \cite{reich2020zoltar}.
+Zoltar also scores models according to mean-absolute-error (MAE) and the weighted-interval score (WIS) \cite{bracher2020evaluating}.
+This allows forecasts to be evaluated both point-wise and probabilistically.
+
  \begin{figure}
@@
      \includegraphics[scale=.175]{data_plot.png}
-     \caption{Deaths by state for all 50 states and territories.
+     \caption{Deaths by state for any state with over 50 incident deaths for a given day.
 Notice the large variability in incident reporting.
@@
 
-\caption{Scores from covid-hub broken down by region, target and timezero.
+\caption{Scores from COVID-HUB broken down by region, target and timezero.
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
@@
 This baseline model uses the previous daily incident as the mean forecast for the current daily incidence, along with  bootstrapped prediction intervals from historical changes in daily incidence.
-See covid-hub for more details.
-\cite{covid-hub}.
+See COVID-HUB for more details.
+\cite{COVID-HUB}.
 
@@
 
+\newpage
+
 \bibliographystyle{unsrt}
-\bibliography{sir}
+\bibliography{bib}
 

### 2020-07-29 score.R

@@ -13,7 +13,7 @@ tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %
 tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
 tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
 #tmp_mech_bayes <- tmp_mech_baye
-tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis))
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis),error=mean(error))
 
 
 #tmp_subset$timezero <- as.factor(tmp_subset$timezero)
@@ -43,3 +43,7 @@ wis_results_by_target <- ggplot(tmp_subset %>% group_by(target,model) %>% summar
 ggsave("/Users/gcgibson/mech_bayes_paper/wis_results_by_target.png",wis_results_by_target,device="png",width=8,height=4)
 
 
+
+error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(unit,model) %>% summarize(error=mean(error)),aes(x=unit,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+
+error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero) %>% summarize(error=mean(error)),aes(x=timezero,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))

### 2020-07-29 data.R
update

@@ -6,8 +6,56 @@ deaths <- read.csv("/Users/gcgibson/Desktop/truth-Incident Deaths.txt")  # Numbe
 
 state_deaths <- deaths %>%
   filter(!str_detect(location_name, "County"))
+state_deaths$date <- as.Date(state_deaths$adate)
 
+state_deaths_subset_idx  <- unique(state_deaths[state_deaths$value >= 50,]$location_name)
+
+state_deaths_subset <- state_deaths[state_deaths$location_name %in% state_deaths_subset_idx,]
 library(ggplot2)
 
-data_plot <- ggplot(state_deaths,aes(x=date,y=value)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free")
+data_plot <- ggplot(state_deaths_subset,aes(x=date,y=value)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free") + theme_bw()
 ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=10,height=10)
+
+
+
+library(reticulate)
+np <- import("numpy")
+files = list.files("/Users/gcgibson/mech_bayes_paper/samples/samples")
+fit_plot_df <- array(NA,dim=c(length(files),3,144) )
+col_idx <- 1
+states = c()
+for (file in files){
+    print (file)
+    npz1 <- np$load(paste0("/Users/gcgibson/mech_bayes_paper/samples/samples/",file),allow_pickle = T)
+    tmp <- npz1$f$post_pred_samples[[1]]$dz
+    tmp_med <- apply(tmp,2,median)
+    tmp_f <- function(x)return(quantile(x,probs = .95))
+    tmp_f2 <- function(x)return(quantile(x,probs = .05))
+    
+    tmp_95 <- apply(tmp,2,tmp_f)
+    tmp_05 <- apply(tmp,2,tmp_f2)
+    
+    fit_plot_df[col_idx,1 ,] <- tmp_med
+    fit_plot_df[col_idx,2 ,] <- tmp_95
+    fit_plot_df[col_idx,3 ,] <- tmp_05
+    
+    
+    col_idx <- col_idx + 1
+    states <-c(states,substr(file,1,2))
+    
+  
+}
+
+plot_df <- data.frame(time=rep(seq(as.Date("2020-03-05"),as.Date("2020-07-26"),by="day"),length(states)),
+                      median=c(t(fit_plot_df[,1,])),upper=c(t(fit_plot_df[,2,])),
[…]

### 2020-07-29 score.R

@@ -46,4 +46,4 @@ ggsave("/Users/gcgibson/mech_bayes_paper/wis_results_by_target.png",wis_results_
 
 error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(unit,model) %>% summarize(error=mean(error)),aes(x=unit,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
 
-error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero) %>% summarize(error=mean(error)),aes(x=timezero,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(error=mean(error)),aes(x=timezero,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))

### 2020-07-29 paper.tex
update

@@
   \centering
-     \begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.3]{US.png}
-    \caption{US}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.3]{NY.png}
-    \caption{New York}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.3]{FL.png}
-    \caption{Florida}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.3]{CA.png}
-    \caption{California}
-\end{subfigure}%
 
-\caption{Example fit and forecast for four states.}
+    \includegraphics[scale=.2]{fit_and_forecast_results.png}
+
+\caption{Example fit and forecast for states with over 50 incident deaths.}
 \label{fig:fit_and_forecast_results}
@@
 
+\item \textbf{MechBayes is biased high}.
+We can see from Figure \ref{fig:bias} that MechBayes was biased high with respect to the median forecast.
+We suspect this is due to the inherent exponential growth of an SEIR model when the number of susceptibles is small.
+Since overall prevalence in any state is well below the number needed to reach herd immunity, when the model sees an increase in cases it treats this as exponential growth, which translates into exponential growth in deaths.
+However, in the recent forecasts the bias is consistently lower than the COVID-HUB baseline model.
+Note the extreme low bias in the first submission is due to the omission of the U.S.
+in the submission.
+\textbf{is this true} \textbf{something is telling me this is also related to the exponential random walk}
+
+\item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
+Small WIS values are better in terms of calibration \cite{bracher2020evaluating}.
+The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise, unlike a fully stochastic model that has inherent variability within the differential equation transitions.
+This suggest that even a deterministic model core can produce well calibrated prediction intervals by capturing the uncertainty in parameter estimation and observation noise.
+
 \end{itemize}
+
+\begin{figure}
+\includegraphics[scale=.2]{bias_by_timezero.png}
+
+\caption{Bias of MechBayes and COVID-Baseline as a function of time. }
+\label{fig:bias}
+\end{figure}
 
@@
 
+However, using relatively simple non-parametric methods we were able to beat the COVID-HUB baseline model in both point and probabilistic forecast evaluations.
+Our nested model comparisons show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.
+
 \newpage

### 2020-07-30 paper.tex
update

@@
 \begin{document}
+
+\maketitle
+
 \section*{Abstract}
@@
 %\end{itemize}
-
-\maketitle
 
@@
 used a fixed detection probability to model undiagnosed individuals in Spain and Italy \cite{lopez2020modified}.
-Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling}\cite{yang2020modified} \cite{bertozzi2020challenges}\cite{prem2020effect}.
+Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling}\cite{yang2020modified} \cite{bertozzi2020challenges}\cite{prem2020effect}\cite{flaxman2020estimating}.
+Perhaps the most similar model was put forth by Pei et al.
+\cite{pei2020differential}.
+This model was an extension to the SEIR model with time varying transmission and detection probability.
+However, their model fixed detection probability across time and was fit using Kalman Filter techniques instead of Bayesian HMC.
+They also used the model mostly for counterfactual scenario projections instead of forecasting.
+Another relatively similar model was put forth by Abbot et al.
+\cite{abbott2020estimating}.
+This model does use a time-varying transmission rate and detection rate, but does not use a differential equation model as the core component, but rather parameterizes new cases as a function of the time varying reproduction number.
 
@@
 \begin{figure}
-  \centering
 
@@
 
-\caption{Example fit and forecast for states with over 50 incident deaths.}
+\caption{Example fit and forecast for states with over 50 incident deaths.
+Grey bands represent 95\% prediction intervals.
+Blue line represents median forecast.
+MechBayes is able to produce well calibrated fits to the data, even under extreme reporting variation such as in North Carolina.
+It is also able to handle extreme outliers, such as in Illinois or New Jersey.
+\textbf{Is this totally true or should we mention manual adjustments.}}
 \label{fig:fit_and_forecast_results}
@@
   \centering
-     \begin{subfigure}{.5\textwidth}
+     \begin{subfigure}{1\textwidth}
   \centering
-    \includegraphics[scale=.1]{beta_t_plot.png}
+    \includegraphics[scale=.2]{beta_t_plot.png}
     \caption{Time varying-beta.}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
+\end{subfigure}
+
+\begin{subfigure}{1\textwidth}
   \centering
-    \includegraphics[scale=.1]{detection_plot.png}
+    \includegraphics[scale=.2]{detection_plot.png}
     \caption{Time-varying detection probability}
 \end{subfigure} 
- \caption{  A) Time varying beta parameter for four example states.
-We can clearly see that the random walk is able to non-parametrically account for the non-pharamceutical interventions.
+ \caption{  A) Time varying beta parameter for four all geographies with a day exceeding 50 incident deaths.
+We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions.
 B) Time varying detection probability for four example states.
@@
 
-We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes version 3 is able to accurately model the observed data in the three example states (NY,CA,FL) as well as the US.
+We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes version 3 is able to accurately model the observed data.
 The model is able to adapt to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves.
 The model is also able capture the uncertainty of the differential equation parameters well enough to produce well calibrated prediction intervals.
-Figure \ref{fig:fit_and_forecast_results}  shows intervals at the 50\%,80\% and 95\% levels.
+Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level.
 However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below.
@@
 \item \textbf{MechBayes is biased high}.
-We can see from Figure \ref{fig:bias} that MechBayes was biased high with respect to the median forecast.
+We can see from Figure \ref{fig:results_discussion} that MechBayes was biased high with respect to the median forecast.
 We suspect this is due to the inherent exponential growth of an SEIR model when the number of susceptibles is small.
@@
 
+\item \textbf{MechBayes Full outperforms the other versions on probabilistic scoring measures}.
+As can be seen by Figure \ref{fig:results_discussion}, MechBayes Full achieves the best log-score \textbf{sub for WIS } out the three nested models.
+This suggested that our extensions to the basic SEIR model not only improve point-forecasts, but also calibration.
+
 \end{itemize}
@@
 \begin{figure}
-\includegraphics[scale=.2]{bias_by_timezero.png}
-
-\caption{Bias of MechBayes and COVID-Baseline as a function of time. }
-\label{fig:bias}
+\begin{subfigure}{.5\textwidth}
+\includegraphics[scale=.15]{bias_by_timezero.png}
+\subcaption{Bias of MechBayes and COVID-Baseline as a function of time. }
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+\includegraphics[scale=.15]{ablation_calibration.png}
+\subcaption{Comparison of nested models using the probabilistic scoring mechanism}
+\end{subfigure}
+\caption{A) Bias of MechBayes over time. B) Probabilistic Calibration}
+\label{fig:results_discussion}
 \end{figure}

### 2020-07-30 ablation.R

@@ -1,6 +1,8 @@
 ablation_res <- read.csv("/Users/gcgibson/mech_bayes_paper/summary-states.csv")
 ablation_res$mae <- ablation_res$MAE
 library(ggplot2)
+
+### 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
 
 ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
@@ -34,4 +36,6 @@ ggsave("/Users/gcgibson/mech_bayes_paper/ablation_4.png",ablation_4,device="png"
 
 
 
-
+ablation_calibration <- ggplot(ablation_res %>% group_by(forecast_date,model) %>% 
+        summarize(ls=mean(LS)),aes(x=forecast_date,y=ls,col=model)) + geom_point(size=1.5) + theme_bw()
+ggsave("/Users/gcgibson/mech_bayes_paper/ablation_calibration.png",ablation_calibration,device="png",width=8,height=4)

### 2020-07-30 analysis.R

@@ -9,14 +9,14 @@ library(sjstats) #use for r2 functions
 
 np <- import("numpy")
 files = list.files("/Users/gcgibson/covid-intervention-analysis/samples/")
-files <- c("US.npz","FL.npz","NY.npz","CA.npz")
+files <- states_to_plot
 beta <- matrix(NA,ncol=length(files),nrow=137 )
 col_idx <- 1
 states = c()
 for (file in files){
   if (file !="DC.npz"){
     print (file)
-    npz1 <- np$load(paste0("/Users/gcgibson/covid-intervention-analysis/samples/",file),allow_pickle = T)
+    npz1 <- np$load(paste0("/Users/gcgibson/covid-intervention-analysis/samples/",file,".npz"),allow_pickle = T)
     beta[,col_idx] = colMeans(npz1$f$mcmc_samples[[1]]$beta)
     states <-c(states,substr(file,1,2))
     col_idx <- col_idx + 1
@@ -32,13 +32,13 @@ library(openintro)
 
 library(stringr)
 
-bet_t_plot <- ggplot(data_for_lmer, aes(x=date,y=R0,group=state,col=state)) + geom_line() + facet_grid(~state) + theme_bw() + ylab("Beta(t)") + xlab("Date")
+bet_t_plot <- ggplot(data_for_lmer, aes(x=date,y=R0,group=state,col=state)) + geom_line() + facet_wrap(~state) + theme_bw() + ylab("Beta(t)") + xlab("Date")
 ggsave("/Users/gcgibson/mech_bayes_paper/beta_t_plot.png",bet_t_plot,device="png",width=8,height=4)
 
-det_probs <- matrix(NA,nrow=137,ncol=4)
+det_probs <- matrix(NA,nrow=137,ncol=length(states_to_plot))
 counter <- 1
-for (file in c("US.txt","CA.txt","NY.txt","FL.txt")){
-  npz1 <- read.csv(paste0("/Users/gcgibson/covid-intervention-analysis/summary/",file))
+for (file in states_to_plot){
+  npz1 <- read.csv(paste0("/Users/gcgibson/covid-intervention-analysis/summary/",file,".txt"))
   
   rw  <-c()
   for (line in 10:146){
@@ -60,7 +60,7 @@ for (file in c("US.txt","CA.txt","NY.txt","FL.txt")){
 data_for_detection  = data.frame(R0 = as.numeric(c((det_probs))), t= rep(1:137,length(states)), state=rep(states,each=137))
[…]

### 2020-07-30 data.R

@@ -50,12 +50,12 @@ plot_df <- data.frame(time=rep(seq(as.Date("2020-03-05"),as.Date("2020-07-26"),b
                       median=c(t(fit_plot_df[,1,])),upper=c(t(fit_plot_df[,2,])),
                       lower=c(t(fit_plot_df[,3,])))
 plot_df$location <- rep(states,each=144)
-plot_df_subset <- plot_df[plot_df$location %in% state.abb[match(state_deaths_subset_idx,state.name)], ]
+plot_df_subset <- plot_df[plot_df$location %in% c("US",state.abb[match(state_deaths_subset_idx,state.name)]), ]
 state_deaths_subset_for_addition <- state_deaths_subset#[state_deaths_subset$location_name != "New Jersey",]
 state_deaths_subset_for_addition <- state_deaths_subset_for_addition[state_deaths_subset_for_addition$date %in% plot_df_subset$time,]
 state_deaths_subset_for_addition$location <- state.abb[match(state_deaths_subset_for_addition$location_name,state.name)]
-
+state_deaths_subset_for_addition[is.na(state_deaths_subset_for_addition$location),]$location <- "US"
 fit_and_forecast_results <- ggplot(plot_df_subset,aes(x=time,y=median)) + geom_line(col="cornflowerblue") +
-  geom_ribbon(aes(ymin=lower, ymax=upper),alpha=0.2,size=0,colour="cornflowerblue")+geom_point(data=state_deaths_subset_for_addition,aes(x=date,y=value),size=.5)  +facet_wrap(~location,scales='free') 
+  geom_ribbon(aes(ymin=lower, ymax=upper),alpha=0.2,size=0,colour="cornflowerblue")+geom_point(data=state_deaths_subset_for_addition,aes(x=date,y=value),size=.5)  +facet_wrap(~location,scales='free') + theme_bw() 
 ggsave("/Users/gcgibson/mech_bayes_paper/fit_and_forecast_results.png",fit_and_forecast_results,device="png",width=8,height=8)
 

### 2020-07-30 score.R

@@ -46,4 +46,5 @@ ggsave("/Users/gcgibson/mech_bayes_paper/wis_results_by_target.png",wis_results_
 
 error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(unit,model) %>% summarize(error=mean(error)),aes(x=unit,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
 
-error_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(error=mean(error)),aes(x=timezero,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+bias_by_timezero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(error=mean(error)),aes(x=timezero,y=error,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+ggsave("/Users/gcgibson/mech_bayes_paper/bias_by_timezero.png",bias_by_timezero,device="png",width=8,height=4)

### 2020-07-30 paper.tex
update

@@
 Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
-We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for the weeks of XXXX.
+We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for the weeks of 2020-05-05, 2020-05-10, 2020-05-17, 2020-05-24 2020-05-31, 2020-06-07, 2020-06-14, 2020-06-2, 2020-06-28, 2020-07-05, 2020-07-12, 2020-07-19, and 2020-07-26.
+Note that not all targets are observed at all weeks.
+This is due to 4 week ahead targets for weeks 2020-07-12 and beyond not being observable by 2020-08-01.
 % \begin{itemize}
@@
 
+\item \textbf{MechBayes is remarkably accurate}.
+As we can see from Figure \ref{fig:covidhub} the MAE values, when averaged over geographical unit, are as low as 20 deaths.
+On average the MAE values are near 50 deaths.
+Although there is an open question as to what is accurate enough to be actionable by public health officials, MechBayes seems to be able to provide an indicator within a reasonable range of error to make decisions about resource allocation.
+However, as expected, there is wide variance by geography due to differing population size.
+\textbf{would it make sense to normalize MAE by population size?}
+
 \end{itemize}
@@
 \begin{subfigure}{.5\textwidth}
-\includegraphics[scale=.15]{bias_by_timezero.png}
+\includegraphics[scale=.125]{bias_by_timezero.png}
 \subcaption{Bias of MechBayes and COVID-Baseline as a function of time. }
@@
 \begin{subfigure}{.5\textwidth}
-\includegraphics[scale=.15]{ablation_calibration.png}
+\includegraphics[scale=.125]{ablation_calibration.png}
 \subcaption{Comparison of nested models using the probabilistic scoring mechanism}

### 2020-07-30 paper.tex
update

@@
 \title{Mechanistic Bayesian Forecasts of COVID19}
-\author{Graham Gibson, Nicholas Reich, Laura Balzer, Dan Sheldon}
+\author{Graham Gibson, Nicholas Reich, Dan Sheldon}
 %\date{}                                           % Activate to display a given date or no date
@@
 In addition, there has also been severe temporal variation in the percent of symptomatic cases  \cite{russel2020using}.
+There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}).
+Most notably, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include "probable" deaths as well as confirmed deaths due to COVID-19.
+However, many other states have at least one outlying value, usually due to backlog reporting, where a large quantity of previous deaths is reported on a single day.
+We can also see from Figure \ref{fig:data} that some states do not report on particular days (usually weekends) leading to 0 incident deaths for the day.
 We can also see relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends.
-Example incident death data is show in \ref{fig:data}.
+Finally, we can see that while some states have reached their first peak and dropped off, others are starting to see a rapid increase in confirmed deaths, such as the Carolinas, Texas, and Florida.
 
-In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the COVID-HUB forecast repository and visualization tool \cite{COVID-HUB}.
+In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the COVID-HUB forecast repository and visualization tool \cite{covidhub}.
 The COVID-HUB team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths since the beginning of the April 2020.
 A COVID-HUB forecast is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
-
-\begin{equation}
-\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}
-\end{equation}
+The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$
 %\begin{itemize}
@@
 This allows forecasts to be evaluated both point-wise and probabilistically.
+We can think of WIS as an approximation the the log-score, a commonly used metric in probabilistic forecasting \cite{bracher2019multibin}\cite{reich2019reply}.
 
@@
      \centering
-     \includegraphics[scale=.175]{data_plot.png}
+     \includegraphics[scale=.14]{data_plot.png}
      \caption{Deaths by state for any state with over 50 incident deaths for a given day.
 Notice the large variability in incident reporting.
-There appears to be a weekly cycle where deaths are under-reported on the weekend.
+There are significant data-dumps, with New Jersey reporting over 1500 backlogged deaths when switching to include "probable" deaths.
+In some states, there appears to be a weekly cycle where deaths are under-reported on the weekend.
+This is especially pronounced for the U.S.
+as a whole.
 We can also see that there are some negative incident deaths, where data are revised to account for deaths that were incorrectly attributed to COVID.
@@
 
-This model assumes that forecasts are made on the current level of interventions.
+This model assumes that forecasts are made on the current level of interventions because $\mathbb{E}[log(\beta_{t+1})] = log(\beta_t)$.
+That is, the expected value of a random walk in the forecasting stage is simply $\beta_t$ at the last observed value of $t$.
+
+However, this non-parametric model is particularly susceptible to noise in reporting of cases, since $\beta(t)$ is the parameter that takes individuals from $S$ into $I$.
+To avoid instability issues, especially when forecasting, we set the random walk value for forecasting to the average of the last 10 days.
+This is a large enough window to smooth over most reporting issues (excepting data dumps).
 
@@
 This is due to 4 week ahead targets for weeks 2020-07-12 and beyond not being observable by 2020-08-01.
+
+ In the real-time evaluation we also made manual adjustments to account for data dumps through a quality-assurance process. This involved,
+
+ \begin{itemize}
+ \item Identifying outliers in recently reported incident cases.
+ \item Search for documented evidence of a data dump. These are usually recorded on state department of health websites and sometimes local news outlets.
+ \item Manually redistribute the incident deaths evenly over the time-frame mentioned by the department of health or news outlet for the backlog window.
+ \end{itemize}
+
+ This process ensured that the observed data does not contain any identifiable outliers (meaning documented by outside sources).
+In real-time this is necessary to avoid drastic over-predictions caused by data dumps.
+For example, New Jersey reported nearly 1600 daily deaths as it switched from reporting only confirmed deaths from COVID-19, to confirmed and probable.
+This would have caused a drastic increase in predictions if not properly identified as data dump.
 % \begin{itemize}
@@
 This allows to perform a nested model comparison using MAE as the scoring metric.
-\textbf{Include something with beta}
+
+ We also fix all non-model component variation.
+That is, we average over the last 10 days of $\beta(t)$ when forecasting, as well as manually redistributing data dumps.
+This ensures that the comparison is only on model components.
 
@@
 Blue line represents median forecast.
-MechBayes is able to produce well calibrated fits to the data, even under extreme reporting variation such as in North Carolina.
-It is also able to handle extreme outliers, such as in Illinois or New Jersey.
-\textbf{Is this totally true or should we mention manual adjustments.}}
+MechBayes is able to produce well calibrated fits to the data.}
 \label{fig:fit_and_forecast_results}
@@
     \includegraphics[scale=.2]{beta_t_plot.png}
-    \caption{Time varying-beta.}
+    \caption{Time-varying beta ($\beta(t)$).}
 \end{subfigure}
@@
     \includegraphics[scale=.2]{detection_plot.png}
-    \caption{Time-varying detection probability}
+    \caption{Time-varying detection probability ($p_{c,t}$)}
 \end{subfigure} 
- \caption{  A) Time varying beta parameter for four all geographies with a day exceeding 50 incident deaths.
+ \caption{  A) Time varying beta parameter for all geographies with a day exceeding 50 incident deaths.
 We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions.
-B) Time varying detection probability for four example states.
+While it may seem that many states follow the same trend, this is slightly misleading since the prior pulls $\beta(t)$ upward for $t$ close to 0 in states that did not experience community transmission in March.
+B) Time varying detection probability for example states.
 The time varying detection random walk is able account for the increase in testing.
-}
+There is remarkable similarity among the estimated detection probability trajectories between states.}
 \label{fig:model_details}
@@
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+Note that targets averaging over timezero include an average over all model versions.
 }
@@
 
-   \section{Ablation Results}
+   \section{Nested Model Comaprison}
 
@@
 
-\caption{Scores from COVID-HUB broken down by region, target and timezero.
-Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+\caption{Nested model comparison MAE broken down by timezero and target.
+We can see that MechBayes Full performs better than MechBayes No Detection Random Walk and MechBayes Only death in almost all breakdowns.
+We can also see that the improvement becomes more pronounced at larger horizons, suggesting that MechBayes Full is a better longer term forecasting model.
 }
@@
 in the submission.
-\textbf{is this true} \textbf{something is telling me this is also related to the exponential random walk}
-
+\textbf{is this true}
 \item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
@@
 However, as expected, there is wide variance by geography due to differing population size.
-\textbf{would it make sense to normalize MAE by population size?}
-
+If we normalize by population size, then the highest MAE by geographical unit is no more than 55 deaths per 100,000 people, with an average of 20 deaths per 1000,000 people.
 \end{itemize}
@@
 \begin{subfigure}{.5\textwidth}
-\includegraphics[scale=.125]{bias_by_timezero.png}
+\includegraphics[scale=.1]{bias_by_timezero.png}
 \subcaption{Bias of MechBayes and COVID-Baseline as a function of time. }
@@
 \begin{subfigure}{.5\textwidth}
-\includegraphics[scale=.125]{ablation_calibration.png}
+\includegraphics[scale=.1]{ablation_calibration.png}
 \subcaption{Comparison of nested models using the probabilistic scoring mechanism}

### 2020-07-30 ablation.R

@@ -3,35 +3,59 @@ ablation_res$mae <- ablation_res$MAE
 library(ggplot2)
 
 ### 
-ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
 
+ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
+ablation_res_subset$model <- as.factor(ablation_res_subset$model)
+levels(ablation_res_subset$model) <- c("Full", "No Det RW", "Only Death")
+ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
+  
+  
 ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-         summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()
+         summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + 
+  theme_bw() + ylab("MAE") + xlab("Date") 
 ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png",width=6,height=6)
 
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(13),]
+ablation_res_subset$model <- as.factor(ablation_res_subset$model)
+levels(ablation_res_subset$model) <- c("Full", "No Det RW", "Only Death")
+ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
+
+
+ablation_res_subset$model <- as.factor(ablation_res_subset$model)
+levels(ablation_res_subset$model) <- c("Full", "No Det RW", "Only Death")
+ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
+
 
 ablation_2 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()
+                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()+ ylab("MAE") + xlab("Date")
 ggsave("/Users/gcgibson/mech_bayes_paper/ablation_2.png",ablation_2,device="png",width=6,height=6)
 
 
 
 
[…]

### 2020-07-30 data.R

@@ -13,7 +13,7 @@ state_deaths_subset_idx  <- unique(state_deaths[state_deaths$value >= 50,]$locat
 state_deaths_subset <- state_deaths[state_deaths$location_name %in% state_deaths_subset_idx,]
 library(ggplot2)
 
-data_plot <- ggplot(state_deaths_subset,aes(x=date,y=value)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free") + theme_bw()
+data_plot <- ggplot(state_deaths_subset,aes(x=date,y=value)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free") + theme_bw() +ylab("Incident Deaths") + xlab("Date")
 ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=10,height=10)
 
 

### 2020-07-30 score.R

@@ -20,12 +20,24 @@ tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>
 tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-04-20",]
 tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
 library(ggplot2)
-mae_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+mae_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90)) + 
+  geom_vline(xintercept=as.numeric(as.Date("2020-05-10")),linetype=4) + geom_vline(xintercept=as.numeric(as.Date("2020-05-24")),linetype=4) +
 ggsave("/Users/gcgibson/mech_bayes_paper/mae_results_by_time_zero.png",mae_results_by_time_zero,device="png",width=8,height=4)
 
 
 mae_results_by_region <- ggplot(tmp_subset %>% group_by(unit,model) %>% summarize(mae=mean(mae)),aes(x=unit,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
 ggsave("/Users/gcgibson/mech_bayes_paper/mae_results_by_region.png",mae_results_by_region,device="png",width=8,height=4)
+
+
+#normalize by population
+pops <- read.csv("/Users/gcgibson/Downloads/nst-est2019-alldata.csv")
+pops$state <- pops$STATE
+library(dplyr)
+tmp_subset$state <- as.integer(tmp_subset$unit)
+tmp_subset_w_pop <- tmp_subset %>% left_join(pops,by = "state")
[…]

### 2020-07-31 paper.tex
first draft

@@
 \usepackage{amssymb}
+\usepackage{url}
 \usepackage{epstopdf}
@@
 
-As of August 1st 2020 the Covid-19 pandemic has caused over 4 million infections and over 150,000 deaths in the United States alone.
+As of August 1st 2020 the Covid-19 pandemic has caused over 4.5 million infections and over 150,000 deaths in the United States alone.
 Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials.
 However, data quality during a pandemic suffers from under-reporting, backlog reporting, and limited testing.
-Classical infectious disease differential equation models, such as the susceptible-exposed-infected-recovered model, do not take these data issues into account.
-We propose a novel Bayesian compartmental model that allows for non-parametric modeling of non-pharamacuetical interventions, time-varying testing, and joint observations on case counts and deaths.
-The model has been used to submit forecasts to the Covid-hub repository organized by the Reich Lab.
-We examine the performance relative to the baseline model in addition to performing a nested model comparison of our extensions to the classic models.
-We demonstrate a significant gain in both MAE and probabilistic scoring measures when taking into account data quality issues during a real-time pandemic.
+Classical infectious disease differential equation models, such as the susceptible-exposed-infected-recovered (SEIR) model, do not take these data issues into account.
+We propose a novel Bayesian compartmental model (MechBayes) that allows for non-parametric modeling of non-pharamacuetical interventions, time-varying testing, and joint observations on case counts and deaths.
+The model has been used to submit forecasts to the COVID19-ForecsatHub repository organized by the Reich Lab.
+We examine the performance relative to a baseline model in addition to performing a nested model comparison of our extensions to the classic SEIR models.
+We demonstrate a significant gain in both point and probabilistic forecast scoring measures when taking into account data quality issues during a real-time pandemic.
 %\begin{itemize}
@@
 The emergence of COVID-19 in early 2020 in the United States developed into the largest pandemic the country has seen in over a century.
-As of July 2020, there are over 4 million confirmed infections and over 150,000 deaths due to COVID-19 \cite{dong2020interactive}.
-Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic recession, and political stability.
+As of August 1st 2020, there are over 4.5 million confirmed infections and over 150,000 deaths due to COVID-19 \cite{dong2020interactive}.
+Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic impact, and clinical response.
 Forecasts of incident and cumulative deaths due to COVID may help in resource allocation, vaccine clinical trial planning, and re-opening strategies.
-Along with non-pharamacutical interventions, forecasts are one of the few public health tools available to help fight the pandemic.
-Infectious disease forecasting has been demonstrated to benefit public health decision makers during annual influenza outbreaks \cite{lutz2019applying}.
+Along with non-pharamacutical interventions, forecasts are one of the few public health tools available to help fight the pandemic in the absence of a vaccine.
+Infectious disease forecasting has been demonstrated to benefit public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.
 However, many forecasts of seasonal disease, such as influenza, often rely on ample historical data to look for patterns that can be projected forward into the future.
@@
 The COVID-19 pandemic has seen a resurgence in the use of differential equation models to explain the underlying transmission of a disease through a population.
-First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically the susceptible, exposed, infected, and recovered compartment \cite{kermack1927contribution}.
-The model is specified by setting the rates of flow between compartments.
+First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically either the susceptible, exposed, infected, or recovered compartment \cite{kermack1927contribution}.
+The model is specified by setting the rates of flow of individuals between compartments.
 While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time.
+
 Compartmental models have been used to effectively model and forecast disease in non-pandemic situations.
-These include complex compartmental models for Influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, including a retrospective model evaluation of the 1918 Influenza pandemic \cite{hall2007real}.
+These include complex compartmental models for Influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, even including a retrospective model evaluation of the 1918 Influenza pandemic \cite{hall2007real}.
 Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, Measles \cite{bokler1993chaos}, Dengue \cite{syafruddin2012seir} and a wide variety of other disease.
@@
 Compartmental models have also been adopted into a Bayesian framework before, including both stochastic disease dynamics and deterministic dynamics \cite{hotta2010bayesian}\cite{dukic2012tracking}.
-In fact, non-parametric transmissibility was included in a Bayesian SEIR model to study Ebola by Frasso and Lambert \cite{frasso2016bayesian}.
-Time-varying transmissibility has also been studied in the frequentist setting \cite{smirnova2019forecasting}.
-With the outbreak of COVID-19, accounting for testing has become a critical element in using an SEIR model.
+Non-parametric transmissibility was included in a Bayesian SEIR model to study Ebola by Frasso and Lambert \cite{frasso2016bayesian}.
+Time-varying transmissibility has also been studied in the frequentist setting using complex non-parametric functions \cite{smirnova2019forecasting}.
+Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling}\cite{yang2020modified} \cite{bertozzi2020challenges}\cite{prem2020effect}\cite{flaxman2020estimating}.
+
+With the outbreak of COVID-19, accounting for testing has become a critical element in effectively using an SEIR model.
 Lopez et al.
 used a fixed detection probability to model undiagnosed individuals in Spain and Italy \cite{lopez2020modified}.
-Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling}\cite{yang2020modified} \cite{bertozzi2020challenges}\cite{prem2020effect}\cite{flaxman2020estimating}.
-Perhaps the most similar model was put forth by Pei et al.
+Perhaps the most similar model to ours was put forth by Pei et al.
 \cite{pei2020differential}.
@@
 They also used the model mostly for counterfactual scenario projections instead of forecasting.
-Another relatively similar model was put forth by Abbot et al.
-\cite{abbott2020estimating}.
-This model does use a time-varying transmission rate and detection rate, but does not use a differential equation model as the core component, but rather parameterizes new cases as a function of the time varying reproduction number.
+Models that have accounted for time-varying testing have mostly used the renewal style equations \cite{abbott2020estimating}\cite{flaxman2020estimating}.
+These models do not use a differential equation model as the core component, but rather parameterizes new cases as a function of the time varying reproduction number and the serial interval.
 
-Emerging pandemics create a unique set of challenges for accurately predicting future deaths.
-These include, but are not limited to, severe under reporting of cases due to asymptomatic transmission, time-varying testing rates, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
+Emerging pandemics create a unique set of challenges for accurately predicting future deaths using differential equation models.
+These include, but are not limited to, severe under reporting of cases due to asymptomatic transmission, time-varying testing rates, backlog data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
 In this work, we introduce a set of extensions to the classic compartmental model that are able to account for the real-world and real-time complexities of infectious disease forecasting during a pandemic.
-We demonstrate the success of the model in both real-time forecast submissions  as well as an ablation test to demonstrate the additional forecast accuracy of our extensions.
+We demonstrate the success of the model in both real-time forecast submissions submitted to the Covid19-ForecastHub as well as a nested model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
 In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
@@
 %\item Cite Flu Forecasting
-%\item Describe COVID-HUB
+%\item Describe Covid19-ForecastHub
 %\begin{itemize}
@@
 %\begin{itemize}
-%\item COVID-HUB relative to baseline model 
+%\item Covid19-ForecastHub relative to baseline model 
 %\item Ablation test relative to basic SEIR model
@@
 	In this analysis we use confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}.
-This a time series dataset which we truncate to begin March 1st 2020 to April 27th 2020 and captures all 50 states, as well as Guam, Puerto Rico, and American Samoa.
+This a time series dataset which we truncate to begin March 1st 2020 to August 1st 2020 and captures all 50 states, as well as Guam, Puerto Rico, and American Samoa.
 As noted in \cite{krantz2020level}, COVID-19 cases are often dramatically under-reported, with reporting rates for the U.S.
@@
 We can also see from Figure \ref{fig:data} that some states do not report on particular days (usually weekends) leading to 0 incident deaths for the day.
-We can also see relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends.
+Some states exhibit relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends.
+This effect is most pronounced at the aggregate U.S.
+level, with a clear weekly cycle in reporting.
 Finally, we can see that while some states have reached their first peak and dropped off, others are starting to see a rapid increase in confirmed deaths, such as the Carolinas, Texas, and Florida.
 
-In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the COVID-HUB forecast repository and visualization tool \cite{covidhub}.
-The COVID-HUB team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths since the beginning of the April 2020.
-A COVID-HUB forecast is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
-The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$
+In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the Covid19-ForecastHub forecast repository and visualization tool \cite{covidhub}.
+The repository is meant to standardize probabilistic forecast formats, submission times, and targets across the COVID-19 modeling community, allowing for both ensemble modeling and easy comparison against baseline models.
+The Covid19-ForecastHub team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths since the beginning of the April 2020.
+A single submission file contains a forecast distribution for 1-4 week ahead incident and cumulative deaths for all geographies.
+An individual forecast distribution is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
+The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$, with the median ($.5$ quantile) representing the point-forecast.
 %\begin{itemize}
@@
 % \begin{figure}
-In a given time-step (e.g. one day), each member of the population of  interest belongs is assumed to be in one of the mutually-and-exhaustive compartments:
+In a given time-step (e.g. one day), each member of the population of interest belongs to one of the following mutually-and-exhaustive compartments:
  Susceptible $S$, Exposed but not yet infectious $E$, Infectious $I$, Recovered $R$, hospitalized before death $D_1$ , and finally deceased $D_2$.
-Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter.
+Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature.
+We omit an explicit hospitalization compartment since the available hospitalization data is highly variable by state and suffers even more reporting issues than case data.
 For simplicity, we assume a closed population of size $N$.
@@
 \item $\gamma$: rate of transition from  the infectious state $I$ to no longer being infectious; i.e., $1/\gamma$ is the expected duration of the infectious period
-\item $\rho$: fatality rate %(i.e., probability of transitioning from I to $D_1$ instead of I to R)
+\item $\rho$: fatality rate (i.e., probability of transitioning from I to $D_1$ instead of I to R)
 \item $\lambda$: rate of transition from $D_1$ to $D_2$ 
@@
 For a given time-step $t$, the following differential equations describe the changes in each compartment:
-$$\begin{aligned} 
+
+\begin{equation}
+\begin{aligned} 
 \frac{dS}{dt} &= - \beta(t) \frac{SI}{N} \\
@@
 \frac{dC}{d} &= \sigma E
-\end{aligned} $$
+\end{aligned}
+\end{equation}
+
+Here, we include the $C$ compartment to be able to observe the cumulative count of new infections. This captures only the flow into $I$.
 
 We can write this in a state space representation as follows: \[
-X(t) = (S(t), E(t), I(t), R(t), D_1(t), D_2(t))
+X(t) = (S(t), E(t), I(t), R(t), D_1(t), D_2(t), C(t))
 \]
 The update from time $t$ to time $t+1$ can be solved numerically as
-$$\bm{X}(t+1)= \text{RK4}\left(\bm{X}(t),\frac{dX}{dt},\beta(t)\right)$$,
-where RK4 is the Runge-Katta 4th order approximation (see numpyro ode docs) \cite{numpyro}.
+\begin{equation}
+\bm{X}(t+1)= \text{RK4}\left(\bm{X}(t),\frac{dX}{dt},\beta(t)\right)
+\end{equation},
+where RK4 is the Runge-Katta 4th order approximation (see numpyro ode docs) \cite{phan2019composable}.
 
@@
 To add to the complexity, these interventions have been implemented and repealed at different time points.
-They also face compliance issues.
+They also face compliance issues \cite{simonov2020persuasive}.
 In order to capture the aggregate effect of the interventions non-parametrically we choose a flexible model for the time-varying transmission parameter.
@@
 
-$$log(\beta(t)) \sim N(log(\beta(t-1), \sigma_{\beta}^2)$$
+\begin{equation}
+log(\beta(t)) \sim N(log(\beta(t-1), \sigma_{\beta}^2)
+\end{equation}
 
-This model assumes that forecasts are made on the current level of interventions because $\mathbb{E}[log(\beta_{t+1})] = log(\beta_t)$.
-That is, the expected value of a random walk in the forecasting stage is simply $\beta_t$ at the last observed value of $t$.
+This model assumes that forecasts are made on the current level of interventions because $\mathbb{E}[log(\beta(t+1))] = log(\beta(t))$.
+That is, the expected value of a random walk in the forecasting stage is simply $\beta(t)$ at the last observed value of $t$.
 
-However, this non-parametric model is particularly susceptible to noise in reporting of cases, since $\beta(t)$ is the parameter that takes individuals from $S$ into $I$.
+However, this non-parametric model is particularly susceptible to noise in reporting of cases, since $\beta(t)$ is the parameter that takes individuals from $S$ into $E$.
 To avoid instability issues, especially when forecasting, we set the random walk value for forecasting to the average of the last 10 days.
 This is a large enough window to smooth over most reporting issues (excepting data dumps).
+
+\begin{equation}
+\beta^{forecast}(t+k) = \frac{1}{10}\sum_{i=0}^{9}\beta(t-i)
+\end{equation}
 
@@
 
- The observed data used to fit the model is based on time-series data of confirmed cases $Cases_{t}$ and recorded deaths $Deaths_{t}$. 
-For a given state and day, the change in the  confirmed cases and reported deaths are subset of the  number of infections $I(t)$ and underlying number of deaths $D2(t)$, respectively.
+ The observed data used to fit the model is based on time-series data of incident confirmed cases $Cases_{t}$ and incident recorded deaths $Deaths_{t}$. 
+For a given state and day, the change in the  confirmed cases and reported deaths are subset of the cumulative number of new infections $C(t)$ and cumulative number of deaths $D2(t)$, respectively.
 Therefore, we introduce two additional parameters for the detection probability of cases $p_c$ and the detection probability of deaths $p_d$.
@@
 
-We also assume the probability that a COVID-19 death is reported $p_d$ has a prior distribution given by  $p_d \sim \Beta(90, 10)$. This prior satisfies $\E[p_d] = 0.9$ with concentration 100.
+We also assume the probability that a COVID-19 death is reported $p_d$ has a prior distribution given by  $p_d \sim \Beta(90, 10)$.
+This prior satisfies $\E[p_d] = 0.9$ with concentration 100.
+That is, we assume that deaths due to COVID-19 are most often correctly reported.
 
@@
 \begin{equation}
-\text{Cases}_{t} \sim NB(p_{c,t}*I_{t},\sigma_{c}^2)
+\text{Cases}_{t} \sim NB(p_{c,t}*[C_{t} - C_{t-1}],\sigma_{c}^2)
 \end{equation}
@@
 \begin{equation}
-\text{Deaths}_{t} \sim NB(p_d*D_{2_{t}}, \sigma_d^2)
+\text{Deaths}_{t} \sim NB(p_d*[D_{2_{t}} - D_{2_{t-1}}], \sigma_d^2)
 \end{equation}
 
+Where the difference in $C_{t}-C_{t-1}$ allows us to translate cumulative new cases to incident cases and similarly with deaths. 
 %  \begin{itemize}
@@
 We do this by assigning uniform probability to all initial states where the number of people in any given compartment at time zero does not exceed 2\% of the total population.
-This is a highly conservative estimate for the number of infected and exposed people at the start of the epidemic.
+This is a highly conservative estimate for the number of infected, exposed, dead and recovered people at the start of the epidemic.
 
@@
 This allows us to initialize the process model:
- $$ X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0) $$ 
+ \begin{equation} 
+ X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0)
+ \end{equation}
 %
@@
  Our prior on rate for leaving the exposed compartment $\sigma$ satisfies $\E[\sigma] = 1/\hat{d}_E$, where $\hat{d}_E$ is an initial guess of the duration of the latent period.
-Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) [cite].
+Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) \cite{midas}.
  %
  Our prior on the rate for leaving the infectious compartment $\gamma$ satisfies $\E[\gamma] = 1/\hat{d}_I$, where $\hat{d}_I$ is an initial guess for the duration of infectiousness.
-The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset (cite).
+The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset \cite{heffner2020emotional}.
 Our prior on the initial transmission rate is derived from the relationship between the basic reproductive number $R(0)$ and the length of the infectious period: $R(0) = \beta(0)/\gamma = \beta(0)\times \hat{d}_I$.
@@
 
-The identifiability of model parameters in compartmental models where the data consists of only a time series of incident cases and deaths presents a problem for uninformative priors.
+The identifiability of model parameters in compartmental models where the data consists only of a time series of incident cases and deaths presents a problem for uninformative priors.
 Using the renewal style equations, it can be shown that the number of newly infected at time $t$ is a function of the time-varying reproductive number, serial interval and previously reported new infections \cite{wallinga2007generation}.
@@
 Therefore, the time series of incident cases is not enough to uniquely identify $\gamma,\sigma,\beta(t)$.
-In order to make the model identifiable, we impose tight priors on the parameters $\sigma$ and $\gamma$ as estimated by the literature, in essence fixing the serial interval and we let $\beta(t)$ vary freely.
+In order to make the model identifiable, we impose tight priors on the parameters $\sigma$ and $\gamma$ as estimated by the literature (in essence fixing the serial interval), and we let $\beta(t)$ vary freely.
 This reflects the underlying biology of the system, since the reciprocal of the sum of $\sigma$ and $\gamma$ may be interpreted as the average time from when an individual becomes infected to when they infect someone else, given that they infect someone else.
@@
 This highlights a fundamental philosophical difference between using compartmental models for forecasting rather than interpreting parameters for epidemiological purposes.
-\textbf{I want to say a little more}
-
+However, putting relatively informative priors on $\sigma$ and $\gamma$, instead of fixing them, still allows for variation by state due to differing demographic characteristics such as age structure.
+Fitting the model in a Bayesian way allows for this unique trade off.
 % \begin{itemize}
@@
   \subsection{Fitting}
-We use the hamiltonian monte carlo algorithm implemented in numpyro (some citation here) to fit the model to data.
-That is, given a time series of confirmed cases ($C_{1:t}$) and confirmed deaths ($D_{1:t}$) we use Bayesian inference (via HMC) to obtain
+We use the hamiltonian monte carlo algorithm implemented in numpyro to fit the model to data \cite{numpyro}.
+That is, given a time series of confirmed cases ($\text{Cases}_{1:t}$) and confirmed deaths ($\text{Deaths}_{1:t}$) we use Bayesian inference (via HMC) to obtain
 
-$$f(\bm{\theta} | C_{1:t},D_{1:t}) \propto f(C_{1:t},D_{1:t} | \bm{\theta})f(\bm{\theta})$$
-
+\begin{equation}
+f(\bm{\theta} | C_{1:t},D_{1:t}) \propto f(C_{1:t},D_{1:t} | \bm{\theta})f(\bm{\theta})
+\end{equation}
 where $\bm{\theta}$ is a vector containing all model parameters.
 
-$$\bm{\theta} =[\beta_{t} ,
+\begin{equation}
+\bm{\theta} =[\beta_{t} ,
 \sigma ,
@@
 D_{2_0} ,
-R_0 ]$$
+R_0 ]
+\end{equation}
 
-We use the numpyro probabilistic framework to fit the model.
-
+We draw 1000 warm-up sample and then 1000 posterior samples.
+The advantage in numpyro is its speed and ease of use on a cluster.
+In practice, we launch each state simultaneously on a cluster leading a total run time of under 30 minutes.
+This allows for rapid model development and evaluation.
+Convergence is monitored through $n_eff$ and $\hat{R}$.
 %
@@
  To evaluate our model, we examine two different scenarios.
-First, we describe the submission process and infrastructure used for the real time evaluation as part of the COVID-HUB consortium.
+First, we describe the submission process and infrastructure used for the real time evaluation as part of the Covid19-ForecastHub consortium.
 Second, we describe the internal evaluation used to demonstrate our model enhancements improve accuracy over a naive compartmental model.
 
- \subsection{COVID-HUB}
- The COVID-HUB began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
+ \subsection{Covid19-ForecastHub}
+ The Covid19-ForecastHub began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
 We began submitting the first version on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
@@
 Model had normal observation noise (instead of negative binomial) and a non-time varying case detection probability.
-Model was fit to cumulative deaths.
+Model was fit to cumulative deaths and cases.
+ \begin{equation}
+\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
+\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+\end{equation}
+
   \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
 Model had normal observation noise (instead of negative binomial) and time varying case detection probability.
-Model was fit to cumulative deaths.
-    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}. Model had negative binomial observation noise  and time varying case detection probability. Model was fit to incident deaths. 
+Model was fit to cumulative deaths and cases.
+
+   \begin{equation}
+\text{Cases}_{t} \sim N(p_{t,c}*C_{t} ,\sigma_{c}^2) \ \ \  
+\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+\end{equation}
+
+    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}.
+Model had negative binomial observation noise  and time varying case detection probability.
+Model was fit to incident deaths and cases.
+     \begin{equation}
+\text{Cases}_{t} \sim NB(p_{c,t}*[C_{t}-C_{t-1}] ,\sigma_{c}^2) \ \ \  
+\text{Deaths}_{t} \sim NB(p_d*[D_{2_{t}} - D_{2_{t-1}}] , \sigma_d^2)
+\end{equation}
+
  \end{itemize}
@@
 
- \subsection{Ablation Test}
+ \subsection{Nested Model Comparison}
 
- While real-time model evaluation is valuable for understanding evolving model performance, we also perform a retrospective evaluation using three model variants.
+ While real-time model evaluation is valuable for understanding evolving model performance, we also perform a retrospective evaluation using three model variants to demonstrate the improvement in accuracy over a baseline SEIR model.
 We define the following variations on MechBayes,
@@
  Note that these are nested models, with MechBayes Death Only contained in MechBayes Fixed Case Detection Probability contained in MechBayes Full.
-This allows to perform a nested model comparison using MAE as the scoring metric.
 
@@
 That is, we average over the last 10 days of $\beta(t)$ when forecasting, as well as manually redistributing data dumps.
-This ensures that the comparison is only on model components.
+This ensures that the comparison is only on model components, and not on data discrepancies.
+Note that we do not include a model without a time varying transmissibility parameter.
+This is because such a model would assume no interventions were put in place, which clearly violates the data-generating process.
+We feel that previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential} \cite{abbott2020estimating}\cite{flaxman2020estimating} \cite{smirnova2019forecasting}.
 
-\section{COVID-Hub Results}
+\section{Covid19-ForecastHub Results}
 
@@
 
-\caption{Example fit and forecast for states with over 50 incident deaths.
+\caption{Example fit and forecast for states with over 50 incident deaths and the U.S..
 Grey bands represent 95\% prediction intervals.
 Blue line represents median forecast.
-MechBayes is able to produce well calibrated fits to the data.}
+MechBayes is able to produce well calibrated fits to the data as well as accurately tracking trends in incident deaths.}
 \label{fig:fit_and_forecast_results}
@@
 \end{subfigure} 
- \caption{  A) Time varying beta parameter for all geographies with a day exceeding 50 incident deaths.
-We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions.
+ \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S..
+We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions, with transmission dropping in states like New York quite rapidly.
 While it may seem that many states follow the same trend, this is slightly misleading since the prior pulls $\beta(t)$ upward for $t$ close to 0 in states that did not experience community transmission in March.
 B) Time varying detection probability for example states.
-The time varying detection random walk is able account for the increase in testing.
+The time varying detection random walk is able account for the increase in testing and other reporting anomolies.
 There is remarkable similarity among the estimated detection probability trajectories between states.}
@@
     \includegraphics[scale=.1]{mae_results_by_time_zero.png}
-    \caption{MAE by timezero}
+    \caption{MAE by forecast week}
 \end{subfigure}%
@@
     \includegraphics[scale=.1]{wis_results_by_time_zero.png}
-    \caption{WIS by timezero}
+    \caption{WIS by forecast week}
 \end{subfigure}
@@
     \includegraphics[scale=.1]{mae_results_by_target.png}
-    \caption{WIS by target}
+    \caption{MAE by target}
 \end{subfigure}%
@@
 
-\caption{Scores from COVID-HUB broken down by region, target and timezero.
+\caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
-Note that targets averaging over timezero include an average over all model versions.
+Note that targets averaging over forecast week include an average over all model versions.
 }
@@
 The model is also able capture the uncertainty of the differential equation parameters well enough to produce well calibrated prediction intervals.
-Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level.
+Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level, with roughly 95\% of observations falling within the bounds for each state.
 However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below.
@@
 
-We next turn to the comparison of MechBayes against the COVID-HUB baseline model.
+We next turn to the comparison of MechBayes against the Covid19-ForecastHub baseline model.
 This baseline model uses the previous daily incident as the mean forecast for the current daily incidence, along with  bootstrapped prediction intervals from historical changes in daily incidence.
-See COVID-HUB for more details.
-\cite{COVID-HUB}.
+See Covid19-ForecastHub for more details.
+\cite{covidhub}.
 
-We begin by breaking down the results by week the forecast was made (timezero) and averaging over both region and target.
-As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 did not outperform the COVID-HUB baseline on MAE or WIS when broken down by timezero.
-However, MechBayes V2 (the introduction of the time-varying detection probability), outperformed the COVID-HUB baseline model on both MAE and WIS when broken down by timezero.
-The same is true for MechBayes V3 with the exception of June 26th, where the COVID-HUB baseline model slightly outperformed MechBayes V3.
+We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target.
+As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 did not outperform the Covid19-ForecastHub baseline on MAE or WIS when broken down by forecast week.
+However, MechBayes V2 (the introduction of the time-varying detection probability), outperformed the Covid19-ForecastHub baseline model on both MAE and WIS when broken down by forecast week.
+The same is true for MechBayes V3 with the exception of June 26th, where the Covid19-ForecastHub baseline model slightly outperformed MechBayes V3.
 However, as seen in Figure \ref{fig:data}, overall incident deaths were at their lowest across the U.S.
@@
 We also break down the results by geographical region, as seen in Figure \ref{fig:covidhub} (C,D).
-Note that this break-down is averaged over target, but also timezero, meaning that we average over each of the model versions.
+Note that this break-down is averaged over target, but also forecast week, meaning that we average over each of the model versions.
 We feel this is important, as it reflects the real-time accuracy of our evolving model efforts, instead of \textbf{cherry-picking} the best model.
@@
 
-Finally, we break down the results by target by averaging over region and timezero  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS.
+Finally, we break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS.
 We can also see that the MAE increase as horizon increase, which is to be expected.
@@
 %\begin{itemize}
-%\item MechBayes is almost always better than baseline model when broken down by region, target, and timezero after June 1st 2020.
+%\item MechBayes is almost always better than baseline model when broken down by region, target, and forecast week after June 1st 2020.
 %\item MechBayes has improved over time.
@@
 
-   \section{Nested Model Comaprison}
+   \section{Nested Model Comparison Reults}
 
-We can see from \ref{fig:ablation} that MechBayes Full is consistently better than MechBayes Death Only or MechBayes Fixed Case Detection Probability.
+We can see from Figure \ref{fig:ablation} that MechBayes Full is consistently better than MechBayes Death Only or MechBayes Fixed Case Detection Probability.
 The only exception seems to be June 22nd 2020 for the two week ahead target.
@@
 MechBayes Full forecasted an exponential growth in cases, which was reflected in a large over prediction two weeks later on July 4th.
+Finally, we can also see that the difference in MAE between MechBayes Full and the competing models increases as forecast horizon increases.
+This suggest that MechBayes Full is a better long term forecasting model.
 
@@
 
-\caption{Nested model comparison MAE broken down by timezero and target.
+\caption{Nested model comparison MAE broken down by forecast week and target.
 We can see that MechBayes Full performs better than MechBayes No Detection Random Walk and MechBayes Only death in almost all breakdowns.
@@
 
-MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world modeling challenges during a pandemic. Our experiments led us to the following conclusions.
+MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world data challenges during a pandemic. Our experiments led us to the following conclusions.
 
@@
 There is remarkable similarity between the increase across geographical locations as well, suggesting (as testing data reflects) an overall increase in the number of detected cases across the U.S.
+Adjusting for case data quality also requires a QA process, where data dumps are redistributed across a specific time-window.
+This avoids the situation where MechBayes interprets data dumps as a drastic change in disease dynamics.
+This is visible in Figure \ref{fig:model_details}, where the time-varying detection probability in New Jersey follows the typical trend of other states, until the data dump, where the detection probability dips significantly down.
+
+\item \textbf{Models must evolve in order to be successful in real-time pandemic forecasting}.
+As Figure \ref{fig:covidhub} shows, MechBayes significantly improved over time relative to the baseline model.
+Forecasting during a real-time pandemic is subject to a wide variety of data reporting issues.
+Continuously responding to challenges is key to being able to forecast well.
+This extends to manual QA processes and extensive forecast validation outside of the model itself.
+
+\item \textbf{MechBayes is biased high}.
+We can see from Figure \ref{fig:results_discussion} that MechBayes was biased high with respect to the median forecast.
+We suspect this is due to the inherent exponential growth of an SEIR model when the number of susceptibles is small.
+Since overall prevalence in any state is well below the number needed to reach herd immunity, when the model sees an increase in cases it treats this as exponential growth, which translates into exponential growth in deaths.
+However, in the recent forecasts the bias is consistently lower than the Covid19-ForecastHub baseline model.
+Note the extreme low bias in the first submission is due to the omission of the U.S.
+in the submission.
+\textbf{is this true}
+
+\item \textbf{Most epidemiological parameters are unidentifiable from the data}.
+MechBayes requires relatively informative priors on $\sigma$, $\gamma$, $\rho$ and $\lambda$.
+These parameters reflect the biology of the disease, from latent incubation period to average time from symptom onset to death.
+Since the data used to fit MechBayes is only a time series of confirmed cases and deaths, these parameters are simply not identifiable from the data.
+However, using a Bayesian framework we can simultaneously set priors based on the literature and allow for small deviations from the prior due to variations across states.
+
+\item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
+Small WIS values are better in terms of calibration \cite{bracher2020evaluating}.
+The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise, unlike a fully stochastic model that has inherent variability within the differential equation transitions.
+This suggest that even a deterministic model core can produce well calibrated prediction intervals by capturing the uncertainty in parameter estimation and observation noise.
+
+\item \textbf{MechBayes is remarkably accurate}.
+As we can see from Figure \ref{fig:covidhub} the MAE values, when averaged over geographical unit, are as low as 20 deaths.
+On average the MAE values are near 50 deaths.
+Although there is an open question as to what is accurate enough to be actionable by public health officials, MechBayes seems to be able to provide an indicator within a reasonable range of error to make decisions about resource allocation.
+However, as expected, there is wide variance by geography due to differing population size.
+If we normalize by population size, then the highest MAE by geographical unit is no more than 55 deaths per 100,000 people, with an average of 20 deaths per 1000,000 people.
 
@@
 
-\item \textbf{Models must evolve in order to be successful in real-time pandemic forecasting}.
-As Figure \ref{fig:covidhub} shows, MechBayes significantly improved over time relative to the baseline model.
-Forecasting during a real-time pandemic is subject to a wide variety of data reporting issues.
-Continuously responding to challenges is key to being able to forecast well.
-
-\item \textbf{MechBayes is biased high}.
-We can see from Figure \ref{fig:results_discussion} that MechBayes was biased high with respect to the median forecast.
-We suspect this is due to the inherent exponential growth of an SEIR model when the number of susceptibles is small.
-Since overall prevalence in any state is well below the number needed to reach herd immunity, when the model sees an increase in cases it treats this as exponential growth, which translates into exponential growth in deaths.
-However, in the recent forecasts the bias is consistently lower than the COVID-HUB baseline model.
-Note the extreme low bias in the first submission is due to the omission of the U.S.
-in the submission.
-\textbf{is this true}
-\item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
-Small WIS values are better in terms of calibration \cite{bracher2020evaluating}.
-The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise, unlike a fully stochastic model that has inherent variability within the differential equation transitions.
-This suggest that even a deterministic model core can produce well calibrated prediction intervals by capturing the uncertainty in parameter estimation and observation noise.
-
 \item \textbf{MechBayes Full outperforms the other versions on probabilistic scoring measures}.
-As can be seen by Figure \ref{fig:results_discussion}, MechBayes Full achieves the best log-score \textbf{sub for WIS } out the three nested models.
-This suggested that our extensions to the basic SEIR model not only improve point-forecasts, but also calibration.
-
-\item \textbf{MechBayes is remarkably accurate}.
-As we can see from Figure \ref{fig:covidhub} the MAE values, when averaged over geographical unit, are as low as 20 deaths.
-On average the MAE values are near 50 deaths.
-Although there is an open question as to what is accurate enough to be actionable by public health officials, MechBayes seems to be able to provide an indicator within a reasonable range of error to make decisions about resource allocation.
-However, as expected, there is wide variance by geography due to differing population size.
-If we normalize by population size, then the highest MAE by geographical unit is no more than 55 deaths per 100,000 people, with an average of 20 deaths per 1000,000 people.
+For simplicity, we only included MAE results for the nested model comparison.
+However, as can be seen from Figure \ref{covidhub}, MAE and WIS are highly correlated.
+We therefore conclude that MechBayes Full is also better calibrated than the other two competing models.
 \end{itemize}
@@
 \begin{figure}
-\begin{subfigure}{.5\textwidth}
-\includegraphics[scale=.1]{bias_by_timezero.png}
-\subcaption{Bias of MechBayes and COVID-Baseline as a function of time. }
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-\includegraphics[scale=.1]{ablation_calibration.png}
-\subcaption{Comparison of nested models using the probabilistic scoring mechanism}
-\end{subfigure}
-\caption{A) Bias of MechBayes over time. B) Probabilistic Calibration}
+\centering
+\includegraphics[scale=.15]{bias_by_timezero.png}
+\caption{Bias of MechBayes and COVID-Baseline as a function of time.  }
 \label{fig:results_discussion}
@@
 We have seen that MechBayes is a powerful Bayesian compartmental model that can capture the real-world complexities of forecasting during a pandemic.
-Through internal and external evaluation, we have demonstrated the success of MechBayes in forecasting.
+Through real-time and retrospective evaluation, we have demonstrated the success of MechBayes in forecasting Covid19.
 The model is able to improve over a naive baseline model as well as a naive compartmental model.
@@
 
-However, using relatively simple non-parametric methods we were able to beat the COVID-HUB baseline model in both point and probabilistic forecast evaluations.
+However, using relatively simple non-parametric methods we were able to beat the Covid19-ForecastHub baseline model in both point and probabilistic forecast evaluations.
 Our nested model comparisons show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.

### 2020-07-31 ablation.R

@@ -1,7 +1,7 @@
 ablation_res <- read.csv("/Users/gcgibson/mech_bayes_paper/summary-states.csv")
 ablation_res$mae <- ablation_res$MAE
 library(ggplot2)
-
+library(ggthemes)
 ### 
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
@@ -12,7 +12,7 @@ ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
   
 ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
          summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + 
-  theme_bw() + ylab("MAE") + xlab("Date") 
+  theme_bw() + ylab("MAE") + xlab("Date")  + scale_color_brewer(palette="Set3")
 ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png",width=6,height=6)
 
 
@@ -28,7 +28,7 @@ ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
 ablation_2 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()+ ylab("MAE") + xlab("Date")
+                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()+ ylab("MAE") + xlab("Date") + scale_color_brewer(palette="Set3")
 ggsave("/Users/gcgibson/mech_bayes_paper/ablation_2.png",ablation_2,device="png",width=6,height=6)
 
 
@@ -41,7 +41,7 @@ ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
 ablation_3 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-                       summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + theme_bw()+ ylab("MAE") + xlab("Date")
[…]

### 2020-07-31 score.R

@@ -22,6 +22,9 @@ tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_
 library(ggplot2)
 mae_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90)) + 
   geom_vline(xintercept=as.numeric(as.Date("2020-05-10")),linetype=4) + geom_vline(xintercept=as.numeric(as.Date("2020-05-24")),linetype=4) +
+  annotate("text", x =as.Date("2020-05-08"), y = 400, label = "V1") +  annotate("text", x =as.Date("2020-05-22"), y = 400, label = "V2") +   annotate("text", x =as.Date("2020-07-22"), y = 400, label = "V3")
+
+
 ggsave("/Users/gcgibson/mech_bayes_paper/mae_results_by_time_zero.png",mae_results_by_time_zero,device="png",width=8,height=4)
 
 
@@ -43,7 +46,10 @@ mae_results_by_target <- ggplot(tmp_subset %>% group_by(target,model) %>% summar
 ggsave("/Users/gcgibson/mech_bayes_paper/mae_results_by_target.png",mae_results_by_target,device="png",width=8,height=4)
 
 
-wis_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(wis=mean(wis)),aes(x=timezero,y=wis,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90))
+wis_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(wis=mean(wis)),aes(x=timezero,y=wis,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90)) + geom_vline(xintercept=as.numeric(as.Date("2020-05-10")),linetype=4) + geom_vline(xintercept=as.numeric(as.Date("2020-05-24")),linetype=4) +
+  geom_vline(xintercept=as.numeric(as.Date("2020-05-10")),linetype=4) + geom_vline(xintercept=as.numeric(as.Date("2020-05-24")),linetype=4) +
[…]

### 2020-08-18 paper.tex
rework intro

@@
 
-As of August 1st 2020 the Covid-19 pandemic has caused over 4.5 million infections and over 150,000 deaths in the United States alone.
+The COVID-19 pandemic emerged in late December 2019.
+In the first six months of the global outbreak, the US reported more cases and deaths than any other country in the world..
 Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials.
-However, data quality during a pandemic suffers from under-reporting, backlog reporting, and limited testing.
-Classical infectious disease differential equation models, such as the susceptible-exposed-infected-recovered (SEIR) model, do not take these data issues into account.
-We propose a novel Bayesian compartmental model (MechBayes) that allows for non-parametric modeling of non-pharamacuetical interventions, time-varying testing, and joint observations on case counts and deaths.
-The model has been used to submit forecasts to the COVID19-ForecsatHub repository organized by the Reich Lab.
+However, data quality during a pandemic suffers from under-reporting, delayed reporting of cases and deaths due to reporting infrastructure issues, and limited testing.
+We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize Covid-19 in forecasting in real time.
+This includes non-parametric modeling of non-pharamacuetical interventions, time-varying rates of testing for the disease, and joint observations on new case counts and new deaths.
+The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab.
 We examine the performance relative to a baseline model in addition to performing a nested model comparison of our extensions to the classic SEIR models.
-We demonstrate a significant gain in both point and probabilistic forecast scoring measures when taking into account data quality issues during a real-time pandemic.
+We demonstrate a significant gain in both point and probabilistic forecast scoring measures using MechBayes.
 %\begin{itemize}
@@
 The emergence of COVID-19 in early 2020 in the United States developed into the largest pandemic the country has seen in over a century.
-As of August 1st 2020, there are over 4.5 million confirmed infections and over 150,000 deaths due to COVID-19 \cite{dong2020interactive}.
-Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic impact, and clinical response.
+Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic impact, treatments and public health response.
 Forecasts of incident and cumulative deaths due to COVID may help in resource allocation, vaccine clinical trial planning, and re-opening strategies.
-Along with non-pharamacutical interventions, forecasts are one of the few public health tools available to help fight the pandemic in the absence of a vaccine.
-Infectious disease forecasting has been demonstrated to benefit public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.
+Forecasts provide important data to decision-makers and the general public and can improve situational awareness of current trends and how they will continue in coming weeks.
+Infectious disease forecasting, at the time horizon of up to 4 weeks in the future, has benefited public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.
 However, many forecasts of seasonal disease, such as influenza, often rely on ample historical data to look for patterns that can be projected forward into the future.
 In an emerging pandemic situation, models must be able to fit to limited data.
-The COVID-19 pandemic has seen a resurgence in the use of differential equation models to explain the underlying transmission of a disease through a population.
+In modeling the COVID-19 pandemic, many research groups have turned to the use of differential equation models to explain the underlying transmission of a disease through a population.
 First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically either the susceptible, exposed, infected, or recovered compartment \cite{kermack1927contribution}.
 The model is specified by setting the rates of flow of individuals between compartments.
-While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time.
+While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time at both local and global scales .
 
-Compartmental models have been used to effectively model and forecast disease in non-pandemic situations.
-These include complex compartmental models for Influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, even including a retrospective model evaluation of the 1918 Influenza pandemic \cite{hall2007real}.
-Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, Measles \cite{bokler1993chaos}, Dengue \cite{syafruddin2012seir} and a wide variety of other disease.
+Compartmental models have been used to effectively model and forecast disease in non-pandemic situations both retrospectively and in real-time.
+These include complex compartmental models for real-time influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, even including a retrospective model evaluation of the 1918 influenza pandemic \cite{hall2007real}.
+Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, measles \cite{bokler1993chaos}, dengue \cite{syafruddin2012seir} and a wide variety of other diseases
+
+In this work, we introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19.
+These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
+The case and death data is reported as new cases, instead of prevalent, and new deaths, instead of cumulative, which we handle using an observation model on top of the underlying SEIR dynamics.
+Finally, we choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting.
+This also allows for implementation in a cutting edge probabilistic programming framework for speed and accuracy.
+
+.
+Our main goal in this work is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting.
+However, our m We are not interested in identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data.
+We are willing to accept biologically unrealistic parameter values or priors for latent variables; but not ones that lead to gross pathologies during inference and forecasting.
+
+ MechBayes is able to account for these operational concerns when forecasting in real-time.
+We demonstrate the success of the model in both forecast submissions submitted to the US Centers for Disease Control via the Covid19-ForecastHub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
+In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
+
+\section{Related Work}
 
@@
 These models do not use a differential equation model as the core component, but rather parameterizes new cases as a function of the time varying reproduction number and the serial interval.
-
-Emerging pandemics create a unique set of challenges for accurately predicting future deaths using differential equation models.
-These include, but are not limited to, severe under reporting of cases due to asymptomatic transmission, time-varying testing rates, backlog data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
-In this work, we introduce a set of extensions to the classic compartmental model that are able to account for the real-world and real-time complexities of infectious disease forecasting during a pandemic.
-We demonstrate the success of the model in both real-time forecast submissions submitted to the Covid19-ForecastHub as well as a nested model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
-In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
 

### 2020-08-19 paper_2020-08-19.tex
update

--- 2020-08-18 paper.tex
@@
 	In this analysis we use confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}.
-This a time series dataset which we truncate to begin March 1st 2020 to August 1st 2020 and captures all 50 states, as well as Guam, Puerto Rico, and American Samoa.
-As noted in \cite{krantz2020level}, COVID-19 cases are often dramatically under-reported, with reporting rates for the U.S.
+This a time series dataset which we truncate to begin March 1st 2020 to August 1st 2020 and captures all 50 states, as well as Guam, Puerto Rico, American Samoa, District of Columbia, Northern Mariana Islands and U.S.
+Virgin Islands.
+As noted in \cite{krantz2020level}, COVID-19 cases are under-reported, with the fraction of all infections reported as cases for the U.S.
 estimated at 20-30\% \cite{russel2020using}.
-In addition, there has also been severe temporal variation in the percent of symptomatic cases  \cite{russel2020using}.
 	There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}).
-Most notably, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include "probable" deaths as well as confirmed deaths due to COVID-19.
-However, many other states have at least one outlying value, usually due to backlog reporting, where a large quantity of previous deaths is reported on a single day.
+For example, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include \"probable" deaths as well as confirmed deaths due to COVID-19 on June 25th 2020.
+However, many other states have at least one outlying value, usually due to backlog reporting, where a large number of deaths occurring in previous weeks are reported on a single day.
 We can also see from Figure \ref{fig:data} that some states do not report on particular days (usually weekends) leading to 0 incident deaths for the day.
+Some states also revise the cause of death, allowing for negative incident deaths.
 Some states exhibit relatively regular weekly reporting cycles, with reporting dropping off significantly on the weekends.
 This effect is most pronounced at the aggregate U.S.
-level, with a clear weekly cycle in reporting.
-Finally, we can see that while some states have reached their first peak and dropped off, others are starting to see a rapid increase in confirmed deaths, such as the Carolinas, Texas, and Florida.
+level, which shows a clear weekly cycle in reporting.
 
-In order to organize collaborative efforts across the modeling community, the ReichLab of UMass Amherst has developed the Covid19-ForecastHub forecast repository and visualization tool \cite{covidhub}.
-The repository is meant to standardize probabilistic forecast formats, submission times, and targets across the COVID-19 modeling community, allowing for both ensemble modeling and easy comparison against baseline models.
-The Covid19-ForecastHub team have been soliciting forecasts for 1-4 week ahead incident and cumulative deaths since the beginning of the April 2020.
-A single submission file contains a forecast distribution for 1-4 week ahead incident and cumulative deaths for all geographies.
+We made probabilistic forecasts  for 1-4 week ahead incident and cumulative deaths for all geographies.
 An individual forecast distribution is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
 The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$, with the median ($.5$ quantile) representing the point-forecast.
-%\begin{itemize}
-%\item We use cases/deaths from JHU
-%\item Under-reporting issues
-%\item Batch reporting issues
-%\item Revision Issues
-%\item Weekly cycle issues
-%\item High variability in incident data
-%\item Ref Figure 1
-%\end{itemize}
-
-Forecasts are due on Monday evenings and therefore use the incident data up until the Sunday before.
+Forecasts were made on Monday evenings and therefore use the incident data up until the Sunday before.
 A one-week ahead target corresponds to the following Saturday.
 A two-week ahead corresponds two the next Saturday and so on.
-Forecasts are stored in the generic repository Zoltar, which allows teams to access both their model and a baseline model \cite{reich2020zoltar}.
-Zoltar also scores models according to mean-absolute-error (MAE) and the weighted-interval score (WIS) \cite{bracher2020evaluating}.
-This allows forecasts to be evaluated both point-wise and probabilistically.
-We can think of WIS as an approximation the the log-score, a commonly used metric in probabilistic forecasting \cite{bracher2019multibin}\cite{reich2019reply}.
+Forecasts were evaluated using mean-absolute-error (MAE) and the weighted-interval score (WIS) \cite{bracher2020evaluating}.
+The WIS is an approximation to the continuously ranked probability score (CRPS), a commonly used metric in probabilistic forecasting \cite{bracher2019multibin}\cite{reich2019reply}.
 
@@
  \section{Compartmental Model}
-% \begin{itemize}
-% \item Basic SEIR model description
-% \item Definition of compartmental model (description of diff eq as flow between compartments)
-% \item Ref Figure 2
-% \item Interpretation of parameters in basic SEIR model 
-% \begin{itemize}
-% \item Sigma
-% \item Gamma
-% \item beta
-% \end{itemize}
-% 
-% \end{itemize}
-% 
-% \begin{figure}
-In a given time-step (e.g. one day), each member of the population of interest belongs to one of the following mutually-and-exhaustive compartments:
- Susceptible $S$, Exposed but not yet infectious $E$, Infectious $I$, Recovered $R$, hospitalized before death $D_1$ , and finally deceased $D_2$.
-Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature.
+
+In a given time-step (e.g. one day), each member of the population of a single geography belongs to one of the following mutually exhaustive compartments:
+ Susceptible $S$, Exposed but not yet infectious $E$, Infectious $I$, Recovered $R$, hospitalized before death $D_1$ , and  deceased $D_2$ (Figure \ref{fig:seird}).
+Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature \cite{russell2020estimating}.
 We omit an explicit hospitalization compartment since the available hospitalization data is highly variable by state and suffers even more reporting issues than case data.
@@
 \item $\beta(t)$: transmission rate, which we allow to vary by time $t$
-\item $\sigma$: rate of transition from the exposed state $E$ to infectious state $I$; i.e., $1/\sigma$ is the expected duration of the latent period
-\item $\gamma$: rate of transition from  the infectious state $I$ to no longer being infectious; i.e., $1/\gamma$ is the expected duration of the infectious period
+\item $\sigma$: rate of transition from the exposed state $E$ to infectious state $I$; i.e., $1/\sigma$ is the expected duration of the time between exposure and symptom onset. 
+\item $\gamma$: rate of transition from  the infectious state $I$ to no longer being infectious (either to state $D_1$ or $R$); i.e., $1/\gamma$ is the expected duration of the infectious period
 \item $\rho$: fatality rate (i.e., probability of transitioning from I to $D_1$ instead of I to R)
@@
 \end{equation},
-where RK4 is the Runge-Katta 4th order approximation (see numpyro ode docs) \cite{phan2019composable}.
+where RK4 is the Runge-Katta 4th order approximation \cite{phan2019composable}.
 
@@
     \node [int, right=of E] (I) {$I$};
+        \node [int, above=of E] (C) {$C$};
+
     \node [int, right=of I] (D_1) {$D_1$};
+
     \node [int, right=of D_1] (D_2) {$D_2$};
@@
                           (E) edge node[yshift=12pt] {$\sigma$       \\[.2em] } (I) 
+                           (E) edge node[yshift=12pt] {$\sigma$       \\[.2em] } (C) 
                            (I) edge node[yshift=12pt] {$\rho\gamma$       \\[.2em] } (D_1)
-                           (I) edge[bend right=90] node[xshift=-20pt,yshift=-20pt] {$(1-\rho)\gamma$       \\[.4em] } (R)
+                           (I) edge node[xshift=30pt,yshift=-5pt] {$(1-\rho)\gamma$       \\[.4em] } (R)
                             (D_1) edge node[yshift=12pt] {$\lambda$       \\[.2em] } (D_2)  ;
@@
 However, this non-parametric model is particularly susceptible to noise in reporting of cases, since $\beta(t)$ is the parameter that takes individuals from $S$ into $E$.
-To avoid instability issues, especially when forecasting, we set the random walk value for forecasting to the average of the last 10 days.
-This is a large enough window to smooth over most reporting issues (excepting data dumps).
+To avoid instability issues, especially when forecasting, we smooth each posterior sample of $\beta$ to be the average over the last 10 days from the sample.
+This is a large enough window to smooth over most reporting issues (excepting delayed reporting).
 
 \begin{equation}
-\beta^{forecast}(t+k) = \frac{1}{10}\sum_{i=0}^{9}\beta(t-i)
+\beta^*_{forecast}(t+k) = \frac{1}{10}\sum_{i=0}^{9}\beta^*(t-i)
 \end{equation}
+
+where $\beta^*(t)$ denotes a sample from the posterior at time $t$.
 
@@
 
-In more detail, $p_c$ is the probability that an infectious person receives a positive test result and is confirmed as case.
-We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$ with concentration 50.
-This means that we expect 30\% of cases to be detected initially.
+In more detail, $p_c$ is the probability that an infected person receives a positive test result and is reported by a state or local health authority as a case.
+We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$  with 90\% probability between \[ .22,0.38\].
+This means that we expect 30\% of cases to be detected initially, as suggested by the literature \cite{midas}.
 However, we also allow this to vary by time.
@@
 This prior satisfies $\E[p_d] = 0.9$ with concentration 100.
-That is, we assume that deaths due to COVID-19 are most often correctly reported.
+That is, we assume that deaths due to COVID-19 are most often correctly reported \cite{weinberger2020estimation}.
 
-Using the above SEIR model and these detection probabilities, we can then express the observed numbers of confirmed cases and deaths as follows.
+Using the above SEIR model and these detection probabilities, we can then express the observed incident numbers of confirmed cases and deaths as follows.
 \begin{equation}
@@
 
-\subsection{Seeding Epidemic}
-Due to the under-reporting of cases, we cannot use the observed data to seed the epidemic.
-We instead allow the model to find the initial state values for all compartments except the number of susceptible people, which we take as the population size of the geographic region minus the sum of the initial values for the other compartments to enforce the constraint that the entire system size sums to the population size.
-We do this by assigning uniform probability to all initial states where the number of people in any given compartment at time zero does not exceed 2\% of the total population.
-This is a highly conservative estimate for the number of infected, exposed, dead and recovered people at the start of the epidemic.
+\subsection{Epidemiological Model Parameters}
 
-\begin{align*}
- E_0 &\sim \Unif(0, 0.02 N) \\
-I_0 &\sim \Unif(0, 0.02 N) \\
-  D_{1_0} &\sim \Unif(0, 0.02 N) \\
-   D_{2_0} &\sim \Unif(0, 0.02 N) \\   
-   R_{0} &\sim \Unif(0, 0.02 N) \\   
-\end{align*}
-
-This allows us to initialize the process model:
- \begin{equation} 
- X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0)
- \end{equation}
-%
-% \begin{itemize}
-% \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
-% \item Put priors on initial seed
-%  \end{itemize}
-
-  \subsection{Priors}
-We also place the following priors on the transition parameters: 
-\begin{align*}
-\sigma &\sim \Gamma(5, 5 \hat{d}_E)\\
-\gamma & \sim \Gamma(7, 7 \hat{d}_I) \\
-\beta(0) &\sim \Gamma(1, \hat{d}_I/\hat{R}) \\
-    \rho &\sim \Beta(10, 90)\\ 
-\lambda &\sim \Gamma(10, 100)
-\end{align*}
- Our prior on rate for leaving the exposed compartment $\sigma$ satisfies $\E[\sigma] = 1/\hat{d}_E$, where $\hat{d}_E$ is an initial guess of the duration of the latent period.
-Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) \cite{midas}.
- %
- Our prior on the rate for leaving the infectious compartment $\gamma$ satisfies $\E[\gamma] = 1/\hat{d}_I$, where $\hat{d}_I$ is an initial guess for the duration of infectiousness.
-The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset \cite{heffner2020emotional}.
-Our prior on the initial transmission rate is derived from the relationship between the basic reproductive number $R(0)$ and the length of the infectious period: $R(0) = \beta(0)/\gamma = \beta(0)\times \hat{d}_I$.
-Therefore, we set our prior on the initial transmission rate to satisfy $\E[\beta(0)] = \hat{R}/\hat{d}_I$ where $\hat{R} = 3.0$ is an initial guess for $R(0)$ and $\hat{d}_I = 2.0$, as described above.
-%%
-Our prior on the fatality rate $\rho$ satisfies $\E[\rho] = 0.1$ with concentration of $100$.
-Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
-
-The identifiability of model parameters in compartmental models where the data consists only of a time series of incident cases and deaths presents a problem for uninformative priors.
+We use relatively informative priors for epidemiological parameters, such as $\gamma$,$\sigma$, $\rho$,  $\lambda$, and initial compartment values.
+The details are described in the Appendix A1.
+However, the identifiability of model parameters in compartmental models where the data consists only of a time series of incident cases and deaths presents a problem for uninformative priors.
 Using the renewal style equations, it can be shown that the number of newly infected at time $t$ is a function of the time-varying reproductive number, serial interval and previously reported new infections \cite{wallinga2007generation}.
@@
 Fitting the model in a Bayesian way allows for this unique trade off.
-% \begin{itemize}
-% \item We choose tight priors based on literature
-%  \item Fundamental unidentifiability due to renewal equation expression where $I(t)$ is a convolution between $R_t$ and Seiral interval
-%  \item Cannot estimate both simultaneously 
-%  \item Most flexibility comes in with beta and detection random walk
-%  \item Ideal for forecasting, maybe not for inference
-%  \end{itemize}
-%  
+
   \subsection{Fitting}
-We use the hamiltonian monte carlo algorithm implemented in numpyro to fit the model to data \cite{numpyro}.
+We use the Hamiltonian Monte Carlo algorithm implemented in numpyro to fit the model to data \cite{numpyro}.
 That is, given a time series of confirmed cases ($\text{Cases}_{1:t}$) and confirmed deaths ($\text{Deaths}_{1:t}$) we use Bayesian inference (via HMC) to obtain
@@
 \begin{equation}
-f(\bm{\theta} | C_{1:t},D_{1:t}) \propto f(C_{1:t},D_{1:t} | \bm{\theta})f(\bm{\theta})
+f(\bm{\theta} | \text{Cases}_{1:t},D_{1:t}) \propto f(\text{Cases}_{1:t},\text{Deaths}_{1:t} | \bm{\theta})f(\bm{\theta})
 \end{equation}
@@
 
-We draw 1000 warm-up sample and then 1000 posterior samples.
-The advantage in numpyro is its speed and ease of use on a cluster.
-In practice, we launch each state simultaneously on a cluster leading a total run time of under 30 minutes.
-This allows for rapid model development and evaluation.
-Convergence is monitored through $n_eff$ and $\hat{R}$.
-%
-% \begin{itemize}
-% \item Fully Bayesian HMC on parameters
-% \item Why we choose deterministic compartmental model with only uncertainty on parameters and observations
-% \item Estimation in numpyro very fast
-%   \end{itemize}
-%   
-%   
-%   
+We draw 1000 warm-up sample and then 1000 posterior samples of model parameters. This allows us to forecast using the posterior predictive distribution.
+
+\begin{equation}
+f(C_{t:(t+h)}, D_{t:(t+h)}) =  \int_{\theta} f(C_{1:t},D_{1:t} | \bm{\theta})f(\bm{\theta})
+\end{equation}
+
+\textbf{For dan to fill in}
  \section{Experimental Setup}
@@
 
- \subsection{Covid19-ForecastHub}
- The Covid19-ForecastHub began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
-We began submitting the first version on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
+ \subsection{Real-Time Forecast Evaluation}
+We began submitting forecasts to the U.S.
+Centers for Disease Control on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
 The forecasts use daily data up to and including the Sunday before submission the next Monday.
 The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
-Our model went through three distinct iterations as we evaluated performance in real-time.
+Our model went through two distinct iterations as we evaluated performance in real-time.
 
@@
 
- \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
+% 
+% \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
 Model had normal observation noise (instead of negative binomial) and a non-time varying case detection probability.
 Model was fit to cumulative deaths and cases.
- \begin{equation}
-\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
-\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
-\end{equation}
+% \begin{equation}
+%\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
+%\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+%\end{equation}
 
-  \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
+  \item \textbf{Version 1: May 10th 2020 through May 24th 2020}.
 Model had normal observation noise (instead of negative binomial) and time varying case detection probability.
@@
 
-    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}.
+    \item \textbf{Version 2: May 24th 2020 through August 1st 2020}.
 Model had negative binomial observation noise  and time varying case detection probability.
@@
 
- These three versions highlight the complexities of forecasting during a real-time pandemic.
+ These two versions highlight the complexities of forecasting during a real-time pandemic.
 Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
-We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for the weeks of 2020-05-05, 2020-05-10, 2020-05-17, 2020-05-24 2020-05-31, 2020-06-07, 2020-06-14, 2020-06-2, 2020-06-28, 2020-07-05, 2020-07-12, 2020-07-19, and 2020-07-26.
+We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
 Note that not all targets are observed at all weeks.
-This is due to 4 week ahead targets for weeks 2020-07-12 and beyond not being observable by 2020-08-01.
+This is due to 4 week ahead targets for weeks 2020-07-13 and beyond not being observable by 2020-08-01.
 
- In the real-time evaluation we also made manual adjustments to account for data dumps through a quality-assurance process. This involved,
+ In the real-time evaluation we also made manual adjustments to account for delayed reporting through a quality-assurance process. This involved,
 
@@
  \item Identifying outliers in recently reported incident cases.
- \item Search for documented evidence of a data dump. These are usually recorded on state department of health websites and sometimes local news outlets.
- \item Manually redistribute the incident deaths evenly over the time-frame mentioned by the department of health or news outlet for the backlog window.
+ \item Searching for documented evidence of a data dump. These are usually recorded on state department of health websites and sometimes local news outlets.
+ \item Manually redistributing the incident deaths evenly over the time-frame mentioned by the department of health or news outlet for the backlog window.
  \end{itemize}
@@
 In real-time this is necessary to avoid drastic over-predictions caused by data dumps.
-For example, New Jersey reported nearly 1600 daily deaths as it switched from reporting only confirmed deaths from COVID-19, to confirmed and probable.
+For example, New Jersey reported nearly 1,600 daily deaths as it switched from reporting only confirmed deaths from COVID-19, to confirmed and probable on 2020-06-25.
 This would have caused a drastic increase in predictions if not properly identified as data dump.
+On 2020-07-07 Texas removed 3,000 confirmed cases when they discovered the reported cases were a result of anitgen testing, which were not considered reportable.
 % \begin{itemize}
@@
 
- \subsection{Nested Model Comparison}
+ \subsection{Ablation Test}
 
@@
 This is because such a model would assume no interventions were put in place, which clearly violates the data-generating process.
-We feel that previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential} \cite{abbott2020estimating}\cite{flaxman2020estimating} \cite{smirnova2019forecasting}.
-
-\section{Covid19-ForecastHub Results}
+Previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential} \cite{abbott2020estimating}\cite{flaxman2020estimating} \cite{smirnova2019forecasting}.
 
@@
 
-\caption{Example fit and forecast for states with over 50 incident deaths and the U.S..
+\caption{Example fit and forecast for 2020-07-26 for states with over 50 incident deaths and the U.S..
 Grey bands represent 95\% prediction intervals.
@@
 \end{subfigure} 
- \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S..
-We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions, with transmission dropping in states like New York quite rapidly.
+ \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S.
+by 2020-07-26.
+We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions retrospectively, with transmission dropping in states like New York quite rapidly.
 While it may seem that many states follow the same trend, this is slightly misleading since the prior pulls $\beta(t)$ upward for $t$ close to 0 in states that did not experience community transmission in March.
@@
 \end{subfigure}%
+\section{Real-Time Model Results}
 
@@
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+We can also see that MAE varies heavily by region, which is an artifact of both population size and number of covid deaths.
+However, in regions with large MAE (left side) we see a significant improvement over the baseline model in terms of both MAE and WIS.
+Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both MAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
 Note that targets averaging over forecast week include an average over all model versions.
@@
 The model is also able capture the uncertainty of the differential equation parameters well enough to produce well calibrated prediction intervals.
-Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level, with roughly 95\% of observations falling within the bounds for each state.
+Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level, with 92.3\% of observations falling within the bounds for each state.
 However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below.
@@
 This results in the model underestimating transmissibility to reduce the flow through the compartments.
-\textbf{Need to think on this a bit}
 
@@
 We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target.
-As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 did not outperform the Covid19-ForecastHub baseline on MAE or WIS when broken down by forecast week.
+As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 outperform the Covid19-ForecastHub baseline on MAE or WIS when broken down by forecast week.
 However, MechBayes V2 (the introduction of the time-varying detection probability), outperformed the Covid19-ForecastHub baseline model on both MAE and WIS when broken down by forecast week.
@@
 
+\section{Appendix}
+\subsection{A1}
+\subsection{Seeding Epidemic}
+Due to the under-reporting of cases, we cannot use the observed data to seed the epidemic.
+We instead allow the model to find the initial state values for all compartments except the number of susceptible people, which we take as the population size of the geographic region minus the sum of the initial values for the other compartments to enforce the constraint that the entire system size sums to the population size.
+We do this by assigning uniform probability to all initial states where the number of people in any given compartment at time zero does not exceed 2\% of the total population.
+This is a highly conservative estimate for the number of infected, exposed, dead and recovered people at the start of the epidemic which is most likely much lower than 2\% of the population.
+
+\begin{align*}
+ E_0 &\sim \Unif(0, 0.02 N) \\
+I_0 &\sim \Unif(0, 0.02 N) \\
+  D_{1_0} &\sim \Unif(0, 0.02 N) \\
+   D_{2_0} &\sim \Unif(0, 0.02 N) \\   
+   R_{0} &\sim \Unif(0, 0.02 N) \\   
+\end{align*}
+
+This allows us to initialize the process model:
+ \begin{equation} 
+ X(0) = (S(0), E(0), I(0), R(0), D_1(0), D_2(0), C(0)) = (N-E_0-I_0-D_{1_0}-D_{2_0},E_0,I_0, R_0, D_{1_0}, D_{2_0}, I_0)
+ \end{equation}
+%
+% \begin{itemize}
+% \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
+% \item Put priors on initial seed
+%  \end{itemize}
+
+  \subsection{Priors}
+We also place the following priors on the transition parameters: 
+\begin{align*}
+\sigma &\sim \Gamma(5, 5 \hat{d}_E)\\
+\gamma & \sim \Gamma(7, 7 \hat{d}_I) \\
+\beta(0) &\sim \Gamma(1, \hat{d}_I/\hat{R}) \\
+    \rho &\sim \Beta(10, 90)\\ 
+\lambda &\sim \Gamma(10, 100)
+\end{align*}
+ Our prior on rate for leaving the exposed compartment $\sigma$ satisfies $\E[\sigma] = 1/\hat{d}_E$, where $\hat{d}_E$ is an initial guess of the duration of the latent period.
+Currently, we assume $\hat{d}_E = 4.0$ based on published estimates (shortened slightly to account for possible infectiousness prior to developing symptoms) \cite{midas}.
+ %
+ Our prior on the rate for leaving the infectious compartment $\gamma$ satisfies $\E[\gamma] = 1/\hat{d}_I$, where $\hat{d}_I$ is an initial guess for the duration of infectiousness.
+The current setting is $\hat{d}_I = 2.0$ to model the likely isolation of individuals after symptom onset \cite{heffner2020emotional}.
+Our prior on the initial transmission rate is derived from the relationship between the basic reproductive number $R(0)$ and the length of the infectious period: $R(0) = \beta(0)/\gamma = \beta(0)\times \hat{d}_I$.
+Therefore, we set our prior on the initial transmission rate to satisfy $\E[\beta(0)] = \hat{R}/\hat{d}_I$ where $\hat{R} = 3.0$ is an initial guess for $R(0)$ and $\hat{d}_I = 2.0$, as described above.
+%%
+Our prior on the fatality rate $\rho$ satisfies $\E[\rho] = 0.1$ with 90\% probability of being between \[0.06,.14\].
+Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
+
+\newpage 
 \bibliographystyle{unsrt}

### 2020-08-19 data.R

@@ -1,19 +1,32 @@
 library(stringr)
 library(dplyr)
-deaths <- read.csv("/Users/gcgibson/Desktop/truth-Incident Deaths.txt")  # Number of students in bed
+library(RCurl)
+#deaths <- read.csv("/Users/gcgibson/Desktop/truth-Incident Deaths.txt")  # Number of students in bed
+x <- getURL("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-truth/truth-Cumulative%20Deaths.csv")
+y <- read.csv(text = x)
+deaths <- y[!grepl("County",y$location_name) , ]
 
 
 
 state_deaths <- deaths %>%
   filter(!str_detect(location_name, "County"))
-state_deaths$date <- as.Date(state_deaths$adate)
+state_deaths$date <- as.Date(state_deaths$date)
+state_deaths <- state_deaths %>% group_by(location_name) %>%mutate(incident=c(value[1],diff(value))) %>% ungroup()
 
-state_deaths_subset_idx  <- unique(state_deaths[state_deaths$value >= 50,]$location_name)
+state_deaths_subset_idx  <- unique(state_deaths[state_deaths$incident >= 50,]$location_name)
 
 state_deaths_subset <- state_deaths[state_deaths$location_name %in% state_deaths_subset_idx,]
 library(ggplot2)
 
-data_plot <- ggplot(state_deaths_subset,aes(x=date,y=value)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free") + theme_bw() +ylab("Incident Deaths") + xlab("Date")
+
+state_deaths_subset_total_deaths <- state_deaths_subset %>% group_by(location) %>% mutate(total_deaths = sum(value)) %>% ungroup()
+state_deaths_subset <- left_join(state_deaths_subset,state_deaths_subset_total_deaths)
+state_deaths_subset$total_deaths <- as.factor(state_deaths_subset$total_deaths)
+state_deaths_subset$total_deaths <- factor(state_deaths_subset$total_deaths,levels = rev(levels(state_deaths_subset$total_deaths)))
+state_deaths_subset$location_name <- as.factor(state_deaths_subset$location_name)
+state_deaths_subset <- state_deaths_subset[order(state_deaths_subset$total_deaths),]
[…]

### 2020-08-19 score.R

@@ -4,12 +4,15 @@ library(lubridate)
 zoltar_connection <- new_connection()
 zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
 ## read in fips
+fips_csv <- read.csv(url("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-locations/locations.csv"))
 
 ## access scores for the COVID-19 Forecasts project
 tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
 
   mutate(wis = (abs_error+.01*interval_2+.025*interval_5+.05*interval_10+.1*interval_20+.15*interval_30+.2*interval_40+.25*interval_50+.3*interval_60+.35*interval_70+.40*interval_80+.45*interval_90)/12,
          mae=abs_error,error=error)
+
+tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
 tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
 tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
 #tmp_mech_bayes <- tmp_mech_baye
@@ -20,15 +23,24 @@ tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>
 tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-04-20",]
 tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
 library(ggplot2)
-mae_results_by_time_zero <- ggplot(tmp_subset %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90)) + 
-  geom_vline(xintercept=as.numeric(as.Date("2020-05-10")),linetype=4) + geom_vline(xintercept=as.numeric(as.Date("2020-05-24")),linetype=4) +
-  annotate("text", x =as.Date("2020-05-08"), y = 400, label = "V1") +  annotate("text", x =as.Date("2020-05-22"), y = 400, label = "V2") +   annotate("text", x =as.Date("2020-07-22"), y = 400, label = "V3")
[…]

### 2020-08-19 paper_2020-08-19.tex
update

@@
 We began submitting forecasts to the U.S.
-Centers for Disease Control on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
+Centers for Disease Control for incident deaths on May 10th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
 The forecasts use daily data up to and including the Sunday before submission the next Monday.
 The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
-Our model went through two distinct iterations as we evaluated performance in real-time.
-
- \begin{itemize}
-
-% 
-% \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
-Model had normal observation noise (instead of negative binomial) and a non-time varying case detection probability.
-Model was fit to cumulative deaths and cases.
-% \begin{equation}
-%\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
-%\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
-%\end{equation}
-
-  \item \textbf{Version 1: May 10th 2020 through May 24th 2020}.
-Model had normal observation noise (instead of negative binomial) and time varying case detection probability.
-Model was fit to cumulative deaths and cases.
-
-   \begin{equation}
-\text{Cases}_{t} \sim N(p_{t,c}*C_{t} ,\sigma_{c}^2) \ \ \  
-\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
-\end{equation}
-
-    \item \textbf{Version 2: May 24th 2020 through August 1st 2020}.
-Model had negative binomial observation noise  and time varying case detection probability.
-Model was fit to incident deaths and cases.
-     \begin{equation}
-\text{Cases}_{t} \sim NB(p_{c,t}*[C_{t}-C_{t-1}] ,\sigma_{c}^2) \ \ \  
-\text{Deaths}_{t} \sim NB(p_d*[D_{2_{t}} - D_{2_{t-1}}] , \sigma_d^2)
-\end{equation}
-
- \end{itemize}
-
- These two versions highlight the complexities of forecasting during a real-time pandemic.
-Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
 We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
@@
 
-\begin{figure}
-  \centering
-     \begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{mae_results_by_time_zero.png}
-    \caption{MAE by forecast week}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
-    \caption{WIS by forecast week}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{mae_results_by_region.png}
-    \caption{MAE by region}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{wis_results_by_region.png}
-    \caption{WIS by region}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{mae_results_by_target.png}
-    \caption{MAE by target}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{wis_results_by_target.png}
-    \caption{WIS by target}
-\end{subfigure}%
-\section{Real-Time Model Results}
-
-\caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
-Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
-We can also see that MAE varies heavily by region, which is an artifact of both population size and number of covid deaths.
-However, in regions with large MAE (left side) we see a significant improvement over the baseline model in terms of both MAE and WIS.
-Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both MAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
-Note that targets averaging over forecast week include an average over all model versions.
-}
-\label{fig:covidhub}
-\end{figure}
-
 We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes version 3 is able to accurately model the observed data.
@@
 
+\begin{figure}
+  \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_time_zero_inc.png}
+    \caption{MAE by forecast week}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_time_zero_inc.png}
+    \caption{WIS by forecast week}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_region_inc.png}
+    \caption{MAE by region}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_region_inc.png}
+    \caption{WIS by region}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_target_inc.png}
+    \caption{MAE by target}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_target_inc.png}
+    \caption{WIS by target}
+\end{subfigure}%
+\section{Real-Time Model Results}
+
+\caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
+Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+We can also see that MAE varies heavily by region, which is an artifact of both population size and number of covid deaths.
+However, in regions with large MAE (left side) we see a significant improvement over the baseline model in terms of both MAE and WIS.
+Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both MAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
+Note that targets averaging over forecast week include an average over all model versions.
+}
+\label{fig:covidhub_appendix}
+\end{figure}
+
 We next turn to the comparison of MechBayes against the Covid19-ForecastHub baseline model.
@@
 We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target.
-As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 outperform the Covid19-ForecastHub baseline on MAE or WIS when broken down by forecast week.
-However, MechBayes V2 (the introduction of the time-varying detection probability), outperformed the Covid19-ForecastHub baseline model on both MAE and WIS when broken down by forecast week.
-The same is true for MechBayes V3 with the exception of June 26th, where the Covid19-ForecastHub baseline model slightly outperformed MechBayes V3.
-However, as seen in Figure \ref{fig:data}, overall incident deaths were at their lowest across the U.S.
-in late June and remarkably stable, meaning this week was the hardest of the weeks to beat the baseline model.
+As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 outperformed the Covid19-ForecastHub baseline on MAE or WIS when broken down by forecast week.
+The same is true for MechBayes V2 with the exception of June 26th, where the Covid19-ForecastHub baseline model slightly outperformed MechBayes V2.
+As we can see from \ref{fig:data}, in weeks with the largest increase in deaths (mostly during the month of May) MechBayes significantly outperformed the baseline model.
+However, in weeks with a small increase or a decrease in deaths, the scores were much closer.
+This suggests that MechBayes performs well where it counts, when the epidemic is taking off nearly exponentially.
 
 We also break down the results by geographical region, as seen in Figure \ref{fig:covidhub} (C,D).
-Note that this break-down is averaged over target, but also forecast week, meaning that we average over each of the model versions.
-We feel this is important, as it reflects the real-time accuracy of our evolving model efforts, instead of \textbf{cherry-picking} the best model.
-However, we still consistent improvements in MAE under MechBayes when broken down by region, with the exception of New York.
-This is mainly due to MechBayes V1 large MAE for New York in March, where most of the cases (and therefore contribution to MAE) occurred.
-The WIS results are similar.
+Note that this break-down is averaged over target (1-4 week ahead), but also forecast week, meaning that we average over each of the model versions.
+We feel this is important, as it reflects the real-time accuracy of our evolving model efforts, instead of cherry-picking the best model.
+However, we still see consistent improvements in MAE under MechBayes when broken down by region.
+We see the largest improvements in the regions where the baseline model had the highest MAE, meaning that MechBayes improves forecasts in regions with more significant viral activity.
 
@@
 
-   \section{Nested Model Comparison Reults}
+   \section{Ablation Test Results}
 
@@
     \includegraphics[scale=.15]{ablation_1.png}
-    \caption{Ablation results for 1 week ahead.}
+    \caption{1 week ahead.}
 \end{subfigure}%
@@
     \includegraphics[scale=.15]{ablation_2.png}
-    \caption{Ablation results for 2 week ahead.}
+    \caption{2 week ahead.}
 \end{subfigure}
@@
     \includegraphics[scale=.15]{ablation_3.png}
-    \caption{Ablation results 3 week ahead.}
+    \caption{3 week ahead.}
 \end{subfigure}%
@@
     \includegraphics[scale=.15]{ablation_4.png}
-    \caption{Ablation results 4 week ahead. }
+    \caption{4 week ahead. }
 \end{subfigure}
@@
 
-MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world data challenges during a pandemic. Our experiments led us to the following conclusions.
+MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world data challenges during a pandemic.
+This model produced consistently accurate real-time forecasts over the course of 3 months, and was ranked as one of the top 3 of 11 models on an independent scoreboard of COVID-19 forecast models \cite{yyg}.
+Our experiments led us to the following conclusions about the performance of this model and the underlying methodology.
 
@@
 \item \textbf{Adding case data when predicting deaths is helpful but only when accounting for data quality issues}.
-Our ablation test (Figure \ref{fig:ablation}) clearly shows that time-varying detection probability is a key feature in the model for reducing MAE of forecasts.
-As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the detection probability at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
+Our ablation tests (Figure \ref{fig:ablation}) clearly shows that time-varying detection probability is a key feature in the model for reducing MAE of forecasts.
+The "Full" model that both incorporates incident cases into the model likelihood and allows for a flexible time-varying case detection rate is consistently more accurate than a model that does not account for cases at all ("Death Only" model) and a model that does account for cases but does not account for a time-varying case detection probability.
+
+%As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the detection probability at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
 There is remarkable similarity between the increase across geographical locations as well, suggesting (as testing data reflects) an overall increase in the number of detected cases across the U.S.
@@
 Continuously responding to challenges is key to being able to forecast well.
-This extends to manual QA processes and extensive forecast validation outside of the model itself.
+This extends to manual QA processes and extensive forecast validation outside of the model itself as described in section 5.1.
 
@@
 
-\item \textbf{MechBayes is remarkably accurate}.
-As we can see from Figure \ref{fig:covidhub} the MAE values, when averaged over geographical unit, are as low as 20 deaths.
-On average the MAE values are near 50 deaths.
+\item \textbf{MechBayes is accurate, especially at short horizons}.
+As we can see from Figure \ref{fig:covidhub} the MAE values, when averaged over geographical unit, are 31 for 1 week ahead, 44 for 2 week ahead, 51 for 3 week ahead and 63 for 4 week ahead.
 Although there is an open question as to what is accurate enough to be actionable by public health officials, MechBayes seems to be able to provide an indicator within a reasonable range of error to make decisions about resource allocation.
 However, as expected, there is wide variance by geography due to differing population size.
-If we normalize by population size, then the highest MAE by geographical unit is no more than 55 deaths per 100,000 people, with an average of 20 deaths per 1000,000 people.
+If we normalize by population size, then the highest MAE by geographical unit is no more than 60 deaths per 100,000 people, with an average of 1 deaths per 1000,000 people.
 
@@
 Our ablation test explicitly did not include a model that fixed $\beta$ across time.
-This is because the model would not converge without the flexibility to capture interventions.
+This is because the model would not converge without the flexibility to capture changes in transmission.
 While non-parametrically modeling interventions is appealing from a forecasting perspective, it does modify the philosophy behind compartmental modeling.
@@
 For simplicity, we only included MAE results for the nested model comparison.
-However, as can be seen from Figure \ref{covidhub}, MAE and WIS are highly correlated.
+However, as can be seen from Figure \ref{fig:covidhub}, MAE and WIS are highly correlated.
 We therefore conclude that MechBayes Full is also better calibrated than the other two competing models.
@@
 We have seen that MechBayes is a powerful Bayesian compartmental model that can capture the real-world complexities of forecasting during a pandemic.
-Through real-time and retrospective evaluation, we have demonstrated the success of MechBayes in forecasting Covid19.
+Through real-time and retrospective evaluation, we have demonstrated the success of MechBayes in forecasting COVID-19.
 The model is able to improve over a naive baseline model as well as a naive compartmental model.
@@
 
+\subsection{A2}
+
+The Covid19-ForecastHub began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
+We began submitting the first version on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
+The forecasts use daily data up to and including the Sunday before submission the next Monday.
+The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
+Our model went through three distinct iterations as we evaluated performance in real-time.
+
+ \begin{itemize}
+
+ \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
+Model had normal observation noise (instead of negative binomial) and a non-time varying case detection probability.
+Model was fit to cumulative deaths and cases.
+ \begin{equation}
+\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
+\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+\end{equation}
+
+  \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
+Model had normal observation noise (instead of negative binomial) and time varying case detection probability.
+Model was fit to cumulative deaths and cases.
+
+   \begin{equation}
+\text{Cases}_{t} \sim N(p_{t,c}*C_{t} ,\sigma_{c}^2) \ \ \  
+\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+\end{equation}
+
+    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}.
+Model had negative binomial observation noise  and time varying case detection probability.
+Model was fit to incident deaths and cases.
+     \begin{equation}
+\text{Cases}_{t} \sim NB(p_{c,t}*[C_{t}-C_{t-1}] ,\sigma_{c}^2) \ \ \  
+\text{Deaths}_{t} \sim NB(p_d*[D_{2_{t}} - D_{2_{t-1}}] , \sigma_d^2)
+\end{equation}
+
+ \end{itemize}
+
+ These three versions highlight the complexities of forecasting during a real-time pandemic.
+Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
+We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for the weeks of 2020-05-05, 2020-05-10, 2020-05-17, 2020-05-24 2020-05-31, 2020-06-07, 2020-06-14, 2020-06-2, 2020-06-28, 2020-07-05, 2020-07-12, 2020-07-19, and 2020-07-26.
+Note that not all targets are observed at all weeks.
+This is due to 4 week ahead targets for weeks 2020-07-12 and beyond not being observable by 2020-08-01.
+
+ These two versions highlight the complexities of forecasting during a real-time pandemic.
+Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
+
+\begin{figure}
+  \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_time_zero.png}
+    \caption{MAE by forecast week}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
+    \caption{WIS by forecast week}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_region.png}
+    \caption{MAE by region}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_region.png}
+    \caption{WIS by region}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{mae_results_by_target.png}
+    \caption{MAE by target}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.1]{wis_results_by_target.png}
+    \caption{WIS by target}
+\end{subfigure}%
+\section{Real-Time Model Results}
+
+\caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
+Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
+We can also see that MAE varies heavily by region, which is an artifact of both population size and number of covid deaths.
+However, in regions with large MAE (left side) we see a significant improvement over the baseline model in terms of both MAE and WIS.
+Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both MAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
+Note that targets averaging over forecast week include an average over all model versions.
+}
+\label{fig:covidhub_appendix}
+\end{figure}
+
 \newpage 

### 2020-08-19 ablation.R

@@ -6,29 +6,28 @@ library(ggthemes)
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full", "No Det RW", "Only Death")
+levels(ablation_res_subset$model) <- c("Full","Fixed Case Detection Probability","Death Only")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
   
-  
 ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-         summarize(mae=mean(mae)),aes(x=forecast_date,y=mae,col=model)) + geom_point(size=1.5) + 
-  theme_bw() + ylab("MAE") + xlab("Date")  + scale_color_brewer(palette="Set3")
-ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png",width=6,height=6)
+         summarize(se = sqrt(var(mae)),mae_t=mean(mae)),aes(x=as.factor(forecast_date),y=mae_t,col=model)) + geom_point(size=1.5) +theme_bw() + ylab("MAE") + xlab("Date")  + scale_color_brewer(palette="Dark2") + ylim(0,800) + theme(legend.position = "none")  
+ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png",width=4,height=6)
 
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(13),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full", "No Det RW", "Only Death")
+levels(ablation_res_subset$model) <- c("Full","Fixed Case Detection Probability","Death Only")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full", "No Det RW", "Only Death")
+levels(ablation_res_subset$model) <- c("Full","Fixed Case Detection Probability","Death Only")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
[…]

### 2020-08-19 score.R

@@ -12,7 +12,7 @@ tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %
   mutate(wis = (abs_error+.01*interval_2+.025*interval_5+.05*interval_10+.1*interval_20+.15*interval_30+.2*interval_40+.25*interval_50+.3*interval_60+.35*interval_70+.40*interval_80+.45*interval_90)/12,
          mae=abs_error,error=error)
 
-tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
+tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead cum death"),]
 tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
 tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
 #tmp_mech_bayes <- tmp_mech_baye
@@ -23,11 +23,11 @@ tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>
 tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-04-20",]
 tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
 library(ggplot2)
-mae_results_by_time_zero <- ggplot(tmp_subset[tmp_subset$timezero >= "2020-05-10",] %>% group_by(timezero,model) %>% summarize(mae=mean(mae)),aes(x=timezero,y=mae,col=model)) + geom_point() + theme_bw() +  theme(axis.text.x = element_text(angle = 90)) + 
-  #annotate("text", x =as.Date("2020-04-15"), y = 150, label = "V1") + 
-  annotate("text", x =as.Date("2020-05-12"), y = 150, label = "V1") +   
-  annotate("text", x =as.Date("2020-07-01"), y = 150, label = "V2") + 
-  #geom_ribbon(aes(xmin =as.Date("2020-04-05"),xmax=as.Date("2020-05-05") ),alpha=.1,fill='thistle1',colour='thistle1') +
[…]

### 2020-08-20 paper_2020-08-19.tex
update

@@
 Lopez et al.
-used a fixed detection probability to model undiagnosed individuals in Spain and Italy \cite{lopez2020modified}.
+used a fixed case and death deviation model to model undiagnosed individuals in Spain and Italy \cite{lopez2020modified}.
 Perhaps the most similar model to ours was put forth by Pei et al.
 \cite{pei2020differential}.
-This model was an extension to the SEIR model with time varying transmission and detection probability.
-However, their model fixed detection probability across time and was fit using Kalman Filter techniques instead of Bayesian HMC.
+This model was an extension to the SEIR model with time varying transmission and case and death deviation model.
+However, their model fixed case and death deviation model across time and was fit using Kalman Filter techniques instead of Bayesian HMC.
 They also used the model mostly for counterfactual scenario projections instead of forecasting.
@@
 %\item Observations on cases and deaths
-%\item Time-varying detection probability for varied testing
+%\item Time-varying case and death deviation model for varied testing
 %\item Non-parametric model of interventions
@@
 A two-week ahead corresponds two the next Saturday and so on.
-Forecasts were evaluated using mean-absolute-error (MAE) and the weighted-interval score (WIS) \cite{bracher2020evaluating}.
+Forecasts were evaluated using relative mean-absolute-error (relMAE) and the weighted-interval score (WIS) \cite{bracher2020evaluating}.
+Relative mean absolute error is the ratio of the mean absolute error of the model of interest divided by the baseline model \cite{reich2016case}.
 The WIS is an approximation to the continuously ranked probability score (CRPS), a commonly used metric in probabilistic forecasting \cite{bracher2019multibin}\cite{reich2019reply}.
@@
 For a given state and day, the change in the  confirmed cases and reported deaths are subset of the cumulative number of new infections $C(t)$ and cumulative number of deaths $D2(t)$, respectively.
-Therefore, we introduce two additional parameters for the detection probability of cases $p_c$ and the detection probability of deaths $p_d$.
+Therefore, we introduce two additional parameters for the case and death deviation model of cases $p_c$ and the case and death deviation model of deaths $p_d$.
 For both, we set fairly flat priors to reflect these parameters are poorly determined from observed data.
@@
 % \item Logistic random walk 
-% \item Fixed detection probability on deaths
+% \item Fixed case and death deviation model on deaths
 %
@@
 The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
-We use the model submissions made in real-time, under the corresponding version of the model as dated above, evaluated on both MAE and WIS for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
+We use the model submissions made in real-time evaluated on both relMAE and WIS for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
 Note that not all targets are observed at all weeks.
@@
 
- \item \textbf{MechBayes Fixed Case Detection Probability} MechBayes Full with $p_{c,t}$ fixed to $p_c$, that is, removing the time-varying detection probability.
+ \item \textbf{MechBayes Fixed case and death deviation model} MechBayes Full with $p_{c,t}$ fixed to $p_c$, that is, removing the time-varying case and death deviation model.
 
@@
 
- Note that these are nested models, with MechBayes Death Only contained in MechBayes Fixed Case Detection Probability contained in MechBayes Full.
+ Note that these are nested models, with MechBayes Death Only contained in MechBayes Fixed case and death deviation model contained in MechBayes Full.
 
@@
     \includegraphics[scale=.2]{detection_plot.png}
-    \caption{Time-varying detection probability ($p_{c,t}$)}
+    \caption{Time-varying case and death deviation model ($p_{c,t}$)}
 \end{subfigure} 
@@
 While it may seem that many states follow the same trend, this is slightly misleading since the prior pulls $\beta(t)$ upward for $t$ close to 0 in states that did not experience community transmission in March.
-B) Time varying detection probability for example states.
+B) Time varying case and death deviation model for example states.
 The time varying detection random walk is able account for the increase in testing and other reporting anomolies.
-There is remarkable similarity among the estimated detection probability trajectories between states.}
+There is remarkable similarity among the estimated case and death deviation model trajectories between states.}
 \label{fig:model_details}
@@
 since March 2020.
-Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in detection probability before climbing again.
+Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in case and death deviation model before climbing again.
 Note that interpreting this strictly as time-varying detection is obscuring the fact that this parameter $p_{t,d}$ can soak-up any excess variation beyond the ability of cases and the case-fatility ratio to explain the number of deaths.
-That is, the time-varying detection probability is able to "de-couple" cases and deaths beyond the case-fataility ratio regardless of the underlying reason (whether that be an increase in testing or shifting age distribution of cases).
+That is, the time-varying case and death deviation model is able to "de-couple" cases and deaths beyond the case-fataility ratio regardless of the underlying reason (whether that be an increase in testing or shifting age distribution of cases).
 Thus, interpretation of this parameter as a strict mapping to testing is incorrect.
@@
   \centering
-    \includegraphics[scale=.1]{mae_results_by_time_zero_inc.png}
-    \caption{MAE by forecast week}
+    \includegraphics[scale=.1005]{mae_results_by_time_zero_inc.png}
+    \caption{relMAE by forecast week}
 \end{subfigure}%
@@
   \centering
-    \includegraphics[scale=.1]{wis_results_by_time_zero_inc.png}
+    \includegraphics[scale=.1005]{wis_results_by_time_zero_inc.png}
     \caption{WIS by forecast week}
@@
   \centering
-    \includegraphics[scale=.1]{mae_results_by_region_inc.png}
-    \caption{MAE by region}
+    \includegraphics[scale=.1005]{mae_results_by_region_inc.png}
+    \caption{relMAE by region}
 \end{subfigure}%
@@
   \centering
-    \includegraphics[scale=.1]{wis_results_by_region_inc.png}
+    \includegraphics[scale=.1005]{wis_results_by_region_inc.png}
     \caption{WIS by region}
@@
   \centering
-    \includegraphics[scale=.1]{mae_results_by_target_inc.png}
-    \caption{MAE by target}
+    \includegraphics[scale=.1005]{mae_results_by_target_inc.png}
+    \caption{relMAE by target}
 \end{subfigure}%
@@
   \centering
-    \includegraphics[scale=.1]{wis_results_by_target_inc.png}
+    \includegraphics[scale=.1005]{wis_results_by_target_inc.png}
     \caption{WIS by target}
@@
 \caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
+We use the relative MAE metric which takes the MAE of MechBayes model and divides by the MAE of the baseline model.
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
@@
 However, in regions with large MAE (left side) we see a significant improvement over the baseline model in terms of both MAE and WIS.
-Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both MAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
+Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both relMAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
 Note that targets averaging over forecast week include an average over all model versions.
@@
 %\begin{itemize}
-%\item First two submissions did not have detection probability random walk
+%\item First two submissions did not have case and death deviation model random walk
 %\item Observations were Normally distributed on cumulative deaths
@@
 
-We can see from Figure \ref{fig:ablation} that MechBayes Full is consistently better than MechBayes Death Only or MechBayes Fixed Case Detection Probability.
+We can see from Figure \ref{fig:ablation} that MechBayes Full is consistently better than MechBayes Death Only or MechBayes Fixed case and death deviation model.
 The only exception seems to be June 22nd 2020 for the two week ahead target.
@@
 
-We can also see that the MechBayes Death Only is consistently better than MechBayes Fixed Case Detection Probability.
+We can also see that the MechBayes Death Only is consistently better than MechBayes Fixed case and death deviation model.
 This may be evidence that naively including case data, without adjusting for time-varying testing rates, may be worse than not including it at all.
@@
 \item \textbf{Adding case data when predicting deaths is helpful but only when accounting for data quality issues}.
-Our ablation tests (Figure \ref{fig:ablation}) clearly shows that time-varying detection probability is a key feature in the model for reducing MAE of forecasts.
-The "Full" model that both incorporates incident cases into the model likelihood and allows for a flexible time-varying case detection rate is consistently more accurate than a model that does not account for cases at all ("Death Only" model) and a model that does account for cases but does not account for a time-varying case detection probability.
+Our ablation tests (Figure \ref{fig:ablation}) clearly shows that time-varying case and death deviation model is a key feature in the model for reducing MAE of forecasts.
+The "Full" model that both incorporates incident cases into the model likelihood and allows for a flexible time-varying case detection rate is consistently more accurate than a model that does not account for cases at all ("Death Only" model) and a model that does account for cases but does not account for a time-varying case and death deviation model.
 
-%As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the detection probability at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
+%As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the case and death deviation model at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
 There is remarkable similarity between the increase across geographical locations as well, suggesting (as testing data reflects) an overall increase in the number of detected cases across the U.S.
@@
 This avoids the situation where MechBayes interprets data dumps as a drastic change in disease dynamics.
-This is visible in Figure \ref{fig:model_details}, where the time-varying detection probability in New Jersey follows the typical trend of other states, until the data dump, where the detection probability dips significantly down.
+This is visible in Figure \ref{fig:model_details}, where the time-varying case and death deviation model in New Jersey follows the typical trend of other states, until the data dump, where the case and death deviation model dips significantly down.
 
@@
 The model is able to improve over a naive baseline model as well as a naive compartmental model.
-Allowing for time-varying interventions and detection probability is a necessary model component during a real-time pandemic forecasting effort.
+Allowing for time-varying interventions and case and death deviation model is a necessary model component during a real-time pandemic forecasting effort.
 
@@
 % \begin{itemize}
-% \item Because of detection probability we cannot seed using initial reported data, since this is an underestimate
+% \item Because of case and death deviation model we cannot seed using initial reported data, since this is an underestimate
 % \item Put priors on initial seed
@@
  \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
-Model had normal observation noise (instead of negative binomial) and a non-time varying case detection probability.
+Model had normal observation noise (instead of negative binomial) and a non-time varying case and death deviation model.
 Model was fit to cumulative deaths and cases.
@@
   \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
-Model had normal observation noise (instead of negative binomial) and time varying case detection probability.
+Model had normal observation noise (instead of negative binomial) and time varying case and death deviation model.
 Model was fit to cumulative deaths and cases.
@@
     \item \textbf{Version 3: May 24th 2020 through August 1st 2020}.
-Model had negative binomial observation noise  and time varying case detection probability.
+Model had negative binomial observation noise  and time varying case and death deviation model.
 Model was fit to incident deaths and cases.

### 2020-08-26 model_eval.Rmd
big update

---
title: "Untitled"
author: "Graham Casey Gibson"
date: "8/20/2020"
output:
  html_document:
    df_print: paged
---

```{r test, warning=FALSE,echo=FALSE, message=FALSE}

## read packages
library(lme4)
library(zoltr) ## devtools::install_github("reichlab/zoltr")
library(tidyverse)
library(lubridate)
## connect to zoltar
zoltar_connection <- new_connection()
zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
## read in fips
fips_csv <- read.csv(url("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-locations/locations.csv"))

## access scores for the COVID-19 Forecasts project
tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
  mutate(wis = (abs_error+.01*interval_2+.025*interval_5+.05*interval_10+.1*interval_20+.15*interval_30+.2*interval_40+.25*interval_50+.3*interval_60+.35*interval_70+.40*interval_80+.45*interval_90)/12,
         mae=abs_error,error=error)

## subset to inc
tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
## subset to models of interest
tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
## convert time to date
tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)

## subset to exactly what we want
tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis),error=mean(error))

## get geographies MB used
states_ <- unique(tmp_subset[tmp_subset$model == "UMass-MechBayes",]$unit) 
## remove goes we added in July 
states_ <- states_[states_!="US" & states_ != "60" & states_ != "66"&states_ != "69" & states_ != "78" & states_ != "72"]
## subset to common geos
tmp_subset <- tmp_subset[tmp_subset$unit %in% states_,]
## move timezero to V3
tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-12",]
## adjust timezero from monday to sunday to be consistent with MB
tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
print (head(tmp_subset))
print (nrow(tmp_subset))
```

```{r test2, fig.align='center',warning=FALSE,echo=FALSE, message=FALSE}
mae_results_by_time_zero_df <- tmp_subset[tmp_subset$timezero >= "2020-05-12" & tmp_subset$timezero <= "2020-08-02",] %>% group_by(timezero,model,unit,target) %>% summarize(mae=mean(mae))

mae_results_by_time_zero_df$timezero <- as.factor(mae_results_by_time_zero_df$timezero)
mae_results_by_time_zero <- ggplot(mae_results_by_time_zero_df,aes(x=timezero,y=log(mae),col=model)) + geom_boxplot() + theme_bw() +  theme(axis.text.x = element_text(angle = 90)) 
mae_results_by_time_zero <- mae_results_by_time_zero + geom_hline(yintercept = 0)
print (mae_results_by_time_zero)
```

# Model 1
In order to answer the question of whether there is a stisticaly significant difference between models in terms of MAE, we model log(MAE) as follows

\begin{eqnarray}
log(MAE)_{m,r,t,h}, &=& \beta_0 + \beta_1*\text{model} + \beta_2*\text{2 week ahead} + \beta_3*\text{3 week ahead} + \beta_4*\text{4 week ahead}  \\
&+&  \beta_5*\text{2 week ahead}*\text{model} + \beta_6*\text{3 week ahead}*\text{model} \\
&+& \beta_7*\text{4 week ahead}*\text{model} + \Sigma_{m,r,t,h}
\end{eqnarray}

where $m$ is a model index, $r$ is a geographic index, $t$ is a time index and $h$ is a horizon index. Here we let $\Sigma$ be an auto-regressive covariance matrix of the form,

\begin{equation}
Cov(log(MAE_{m,r,t_i,h}),log(MAE_{m,r,t_j,h})) = \sigma^2 * \rho^{|t_i-t_j|}
\end{equation}

```{r model 1, message=FALSE}

library(lme4)
library(nlme)
library(pander)
diagnostic_fit <- lme(log(mae + .00001)~ model*target ,random  = ~ 1  + timezero | unit,data=mae_results_by_time_zero_df)
diagnostic_fit <- update(diagnostic_fit, correlation = corAR1())
pander(summary(diagnostic_fit))

```

# Model 2

We can also fit a pure random effects model, with a general covariance structure.

\begin{eqnarray}
log(MAE)_{m,r,t,h}, &=& \beta_0 + \beta_1*\text{model} + \beta_2*\text{2 week ahead} + \beta_3*\text{3 week ahead} + \beta_4*\text{4 week ahead}  \\
&+&  \beta_5*\text{2 week ahead}*\text{model} + \beta_6*\text{3 week ahead}*\text{model}  \\
&+& \beta_7*\text{4 week ahead}*\text{model} +  \beta_8*\text{time} + b_0_r + b_1_r*\text{time} \\
b_0_r &~& N(0,\Sigma_{b_0}^2)\\
b_1_r &~& N(0,\Sigma_{b_1}^2)\\
\end{eqnarray}

```{r model 2, message=FALSE}
diagnostic_fit <- lmer(log(mae+.00001)~ model*target +timezero+  ( 1  + timezero | unit),data=mae_results_by_time_zero_df)

summary(diagnostic_fit)
```

### 2020-08-26 paper_2020-08-19.tex
big update

@@
 \title{Mechanistic Bayesian Forecasts of COVID19}
-\author{Graham Gibson, Nicholas Reich, Dan Sheldon}
+\author{Graham C. Gibson, Nicholas Reich, Dan Sheldon}
 %\date{}                                           % Activate to display a given date or no date
@@
 The COVID-19 pandemic emerged in late December 2019.
-In the first six months of the global outbreak, the US reported more cases and deaths than any other country in the world..
+In the first six months of the global outbreak, the US reported more cases and deaths than any other country in the world.
 Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials.
-However, data quality during a pandemic suffers from under-reporting, delayed reporting of cases and deaths due to reporting infrastructure issues, and limited testing.
+However, data quality during a pandemic suffers from under-reporting, delayed reporting due to reporting infrastructure issues, and limited testing.
+Traditionally, compartmental models have been used to model emerging pandemics, however, the basic compartmental framework is difficult to operationalize in a complex data environment.
 We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize Covid-19 in forecasting in real time.
-This includes non-parametric modeling of non-pharamacuetical interventions, time-varying rates of testing for the disease, and joint observations on new case counts and new deaths.
+This includes non-parametric modeling of varying transmission rates, non-parametric modelling of case detection versus death detection, and finally joint observation likelihood on new case counts and new deaths.
 The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab.
-We examine the performance relative to a baseline model in addition to performing a nested model comparison of our extensions to the classic SEIR models.
+We examine the performance relative to a baseline model in addition to performing an ablation test of our extensions to the classic SEIR models.
 We demonstrate a significant gain in both point and probabilistic forecast scoring measures using MechBayes.
@@
 The model is specified by setting the rates of flow of individuals between compartments.
-While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time at both local and global scales .
+While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time at both local and global scales.
 
@@
 Finally, we choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting.
-This also allows for implementation in a cutting edge probabilistic programming framework for speed and accuracy.
+We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.
+Implementation in a Bayesian framework also allows for implementation in a cutting edge probabilistic programming framework for speed and accuracy.
 
-.
 Our main goal in this work is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting.
@@
     \path[->, auto=false] (S) edge node[yshift=12pt] {$\beta(t) I$ \\[.2em]} (E)
-                          (E) edge node[yshift=12pt] {$\sigma$       \\[.2em] } (I) 
-                           (E) edge node[yshift=12pt] {$\sigma$       \\[.2em] } (C) 
-                           (I) edge node[yshift=12pt] {$\rho\gamma$       \\[.2em] } (D_1)
+                          (E) edge node[yshift=5pt] {$\sigma$       \\[.2em] } (I) 
+                           (E) edge node[yshift=-12pt,xshift=5pt] {$\sigma$       \\[.2em] } (C) 
+                           (I) edge node[yshift=5pt] {$\rho\gamma$       \\[.2em] } (D_1)
                            (I) edge node[xshift=30pt,yshift=-5pt] {$(1-\rho)\gamma$       \\[.4em] } (R)
-                            (D_1) edge node[yshift=12pt] {$\lambda$       \\[.2em] } (D_2)  ;
+                            (D_1) edge node[yshift=5pt] {$\lambda$       \\[.2em] } (D_2)  ;
                           \end{tikzpicture}
@@
   \centering
-    \includegraphics[scale=.1005]{mae_results_by_time_zero_inc.png}
-    \caption{relMAE by forecast week}
+    \includegraphics[scale=.135]{mae_results_by_time_zero_inc.png}
+    \caption{MAE by forecast week}
 \end{subfigure}%
@@
   \centering
-    \includegraphics[scale=.1005]{wis_results_by_time_zero_inc.png}
+    \includegraphics[scale=.135]{wis_results_by_time_zero_inc.png}
     \caption{WIS by forecast week}
@@
   \centering
-    \includegraphics[scale=.1005]{mae_results_by_region_inc.png}
-    \caption{relMAE by region}
+    \includegraphics[scale=.135]{mae_results_by_target_inc.png}
+    \caption{MAE by target}
 \end{subfigure}%
@@
   \centering
-    \includegraphics[scale=.1005]{wis_results_by_region_inc.png}
-    \caption{WIS by region}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1005]{mae_results_by_target_inc.png}
-    \caption{relMAE by target}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1005]{wis_results_by_target_inc.png}
+    \includegraphics[scale=.135]{wis_results_by_target_inc.png}
     \caption{WIS by target}
 \end{subfigure}%
-\section{Real-Time Model Results}
 
-\caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
+\caption{Scores from Covid19-ForecastHub broken down by target and forecast week.
 We use the relative MAE metric which takes the MAE of MechBayes model and divides by the MAE of the baseline model.
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
-We can also see that MAE varies heavily by region, which is an artifact of both population size and number of covid deaths.
-However, in regions with large MAE (left side) we see a significant improvement over the baseline model in terms of both MAE and WIS.
 Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both relMAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
@@
 \end{figure}
+
+\begin{figure}
+  \centering
+
+\begin{subfigure}{\textwidth}
+  \centering
+    \includegraphics[scale=.2]{mae_results_by_region_inc.png}
+    \caption{MAE by region}
+\end{subfigure}
+\begin{subfigure}{\textwidth}
+  \centering
+    \includegraphics[scale=.20]{wis_results_by_region_inc.png}
+    \caption{WIS by region}
+\end{subfigure}
+\caption{Scores from Covid19-ForecastHub broken down by region for the 20 regions with the highest incident deaths.
+Regions are sorted by total deaths left to right.
+Here we can see that MechBayes has the greatest increase in states with large total death counts.
+States with medium or small death counts have more mixed results.
+The trends are similar for WIS, with an even more pronounced improvement for states with large death counts.
+}
+\label{fig:covidhub_region_results}
+\end{figure}
+
+\section{Real-Time Model Results}
 
@@
 
-Finally, we break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS.
+We break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS.
 We can also see that the MAE increase as horizon increase, which is to be expected.
 We can also see that incident MAE is lower than cumulative MAE, which is again to be expected due to the lower absolute numbers of incident deaths.
-Finally, we see the model is slightly better calibrated on incident deaths than cumulative.
+
+Finally, we include a formal test of the difference in MAE and WIS to demonstrate statistically significant advantages in using MechBayes over the baseline model.
+In order to do this, we just a random effects regression model of the form,
+
+\begin{eqnarray}
+log(MAE_{m,t,r,h} +\delta) &=& \beta_0 +  \beta1*h_1 + ....+ \beta_3*h_3 \\
+&+& \beta_4 *h_1*mb + .... \beta_8*h_4*mb\\
+ &+& b_{r} + \epsilon \\
+b_{r} &\sim &N(0,\Sigma_b^2)\\
+\epsilon &\sim& N(0,\sigma^2)
+\end{eqnarray}
+
+where $mb$ is an indicator for the MechBayes model, $t$ is timezero, $r$ is region, and $h$ is target horizon (1-4 week ahead).
+We chose this model because it explains the variation in MAE by model and horizon while allowing varying baseline MAE values by region.
+Here, variation over time in MAE within a specific region is explained by differences in model performance This leads to the following coefficient estimates for the fixed effects,
+
+\begin{table}[ht]
+\centering
+\begin{tabular}{rrrrrr}
+  \hline
+ & Estimate & Std. Error & df & t value & Pr($>$$|$t$|$) \\ 
+  \hline
+$\beta_0$  & 1.96 & 0.24 & 56.74 & 8.24 & $0.00^*$ \\ 
+ $\beta_{h_1}$  & 0.45 & 0.10 & 4643.00 & 4.66 & $0.00^*$ \\ 
+   $\beta_{h_2} $ & 0.61 & 0.10 & 4643.00 & 6.33 & $0.00^*$ \\ 
+  $\beta_{h_3}$ & 0.75 & 0.10 & 4643.00 & 7.63 & $0.00^*$ \\ 
+$\beta_{h_1 mb}$ & -0.18 & 0.10 & 4643.00 & -1.88 & 0.06 \\ 
+  $\beta_{h_2 mb}$& -0.37 & 0.10 & 4643.00 & -3.85 & $0.00^*$ \\ 
+  $\beta_{h_3mb}$& -0.27 & 0.10 & 4643.00 & -2.78 & $0.01^*$ \\ 
+ $\beta_{h_4mb}$& -0.37 & 0.10 & 4643.00 & -3.68 & $0.00^*$ \\ 
+   \hline
+\end{tabular}
+\caption{Coefficient estimates and t-values for MAE evaluation model.
+We can see that MechBayes performs statistically significantly better than the baseline model for 2-4 weeks ahead.
+The performance increase does not seem to follow a particular pattern.}
+\end{table}
 %
@@
 
-\caption{Nested model comparison MAE broken down by forecast week and target.
+\caption{Ablation test MAE broken down by forecast week and target.
 We can see that MechBayes Full performs better than MechBayes No Detection Random Walk and MechBayes Only death in almost all breakdowns.
@@
 
-\item \textbf{Models must evolve in order to be successful in real-time pandemic forecasting}.
-As Figure \ref{fig:covidhub} shows, MechBayes significantly improved over time relative to the baseline model.
-Forecasting during a real-time pandemic is subject to a wide variety of data reporting issues.
-Continuously responding to challenges is key to being able to forecast well.
-This extends to manual QA processes and extensive forecast validation outside of the model itself as described in section 5.1.
+\item \textbf{MechBayes is accurate when compared to a baseline model}.
+As Figure \ref{fig:covidhub} shows, MechBayes improved relative to the baseline model when broken down by timezero and target.
+The results are more pronounced when breaking down by target, where the boxplots (95\%) are almost completely separated.
+These results are statistically significant for 2-4 weeks ahead, suggesting that the biggest improvements from MechByes over the baseline model comes in longer term forecasting.
+Additionally, as seen in Figure \ref{fig:covidhub_region_results} , the biggest gains in performance, in both MAE and WIS, occur in regions with the largest total deaths counts.
+This is a desirable feature of a pandemic forecasting model.
 
@@
 
-\item \textbf{MechBayes is accurate, especially at short horizons}.
-As we can see from Figure \ref{fig:covidhub} the MAE values, when averaged over geographical unit, are 31 for 1 week ahead, 44 for 2 week ahead, 51 for 3 week ahead and 63 for 4 week ahead.
-Although there is an open question as to what is accurate enough to be actionable by public health officials, MechBayes seems to be able to provide an indicator within a reasonable range of error to make decisions about resource allocation.
-However, as expected, there is wide variance by geography due to differing population size.
-If we normalize by population size, then the highest MAE by geographical unit is no more than 60 deaths per 100,000 people, with an average of 1 deaths per 1000,000 people.
-
 \item \textbf{Allowing for time-varying transmissibility is necessary to non-parametrically capture the effect of non-pharamceutical interventions}.
@@
 \item \textbf{MechBayes Full outperforms the other versions on probabilistic scoring measures}.
-For simplicity, we only included MAE results for the nested model comparison.
+For simplicity, we only included MAE results for the ablation test.
 However, as can be seen from Figure \ref{fig:covidhub}, MAE and WIS are highly correlated.
@@
 However, using relatively simple non-parametric methods we were able to beat the Covid19-ForecastHub baseline model in both point and probabilistic forecast evaluations.
-Our nested model comparisons show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.
+Our ablation tests show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.
 

### 2020-08-26 model_rank_test.R

@@ -1,0 +1,79 @@
+library(lme4)
+library(zoltr) ## devtools::install_github("reichlab/zoltr")
+library(tidyverse)
+library(lubridate)
+zoltar_connection <- new_connection()
+zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
+## read in fips
+fips_csv <- read.csv(url("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-locations/locations.csv"))
+
+## access scores for the COVID-19 Forecasts project
+tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
+  
+  mutate(wis = (abs_error+.01*interval_2+.025*interval_5+.05*interval_10+.1*interval_20+.15*interval_30+.2*interval_40+.25*interval_50+.3*interval_60+.35*interval_70+.40*interval_80+.45*interval_90)/12,
+         mae=abs_error,error=error,truth=truth)
+
+tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
+tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
+tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
+#tmp_mech_bayes <- tmp_mech_baye
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target,truth) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis),error=mean(error))
+states_ <- unique(tmp_subset[tmp_subset$model == "UMass-MechBayes",]$unit) 
+states_ <- states_[states_!="US" & states_ != "60" & states_ != "66"&states_ != "69" & states_ != "78" & states_ != "72"]
+tmp_subset <- tmp_subset[tmp_subset$unit %in% states_,]
+
+#tmp_subset$timezero <- as.factor(tmp_subset$timezero)
+tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-12",]
+tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
+
+
[…]

### 2020-08-26 score_calibration.R

@@ -1,0 +1,102 @@
+library(zoltr) ## devtools::install_github("reichlab/zoltr")
+library(tidyverse)
+library(lubridate)
+zoltar_connection <- new_connection()
+zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
+## read in fips
+fips_csv <- read.csv(url("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-locations/locations.csv"))
+
+## access scores for the COVID-19 Forecasts project
+#tmp <-  download_forecast(zoltar_connection, "https://www.zoltardata.com/api/model//")   
+tmp <- do_zoltar_query(zoltar_connection, "https://www.zoltardata.com/api/project/44/", c("MechBayes"))
+
+tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
+tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
+tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
+#tmp_mech_bayes <- tmp_mech_baye
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target,truth) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis),error=mean(error))
+states_ <- unique(tmp_subset[tmp_subset$model == "UMass-MechBayes",]$unit) 
+states_ <- states_[states_!="US" & states_ != "60" & states_ != "66"&states_ != "69" & states_ != "78" & states_ != "72"]
+tmp_subset <- tmp_subset[tmp_subset$unit %in% states_,]
+
+#tmp_subset$timezero <- as.factor(tmp_subset$timezero)
+tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-12",]
+tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
+library(ggplot2)
+library(tidyr)
+
+mae_results_by_time_zero_df <- tmp_subset[tmp_subset$timezero >= "2020-05-12" & tmp_subset$timezero <= "2020-08-02",] %>% group_by(timezero,model,unit,target) %>% summarize(mae=mean(mae))
+
+
[…]

### 2020-08-26 score_inc.R

@@ -1,0 +1,104 @@
+library(zoltr) ## devtools::install_github("reichlab/zoltr")
+library(tidyverse)
+library(lubridate)
+zoltar_connection <- new_connection()
+zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
+## read in fips
+fips_csv <- read.csv(url("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-locations/locations.csv"))
+
+## access scores for the COVID-19 Forecasts project
+tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
+  mutate(interval_2 = interval_2,
+         mae=abs_error,error=error,ls=log_single_bin,truth=truth)
+
+
+tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
+tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
+tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
+#tmp_mech_bayes <- tmp_mech_baye
+tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target,truth) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis),error=mean(error))
+states_ <- unique(tmp_subset[tmp_subset$model == "UMass-MechBayes",]$unit) 
+states_ <- states_[states_!="US" & states_ != "60" & states_ != "66"&states_ != "69" & states_ != "78" & states_ != "72"]
+tmp_subset <- tmp_subset[tmp_subset$unit %in% states_,]
+
+#tmp_subset$timezero <- as.factor(tmp_subset$timezero)
+tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-12",]
+tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
+library(ggplot2)
+library(tidyr)
+
+mae_results_by_time_zero_df <- tmp_subset[tmp_subset$timezero >= "2020-05-12" & tmp_subset$timezero <= "2020-08-02",] %>% group_by(timezero,model,unit,target) %>% summarize(mae=mean(mae))
+
+
[…]

### 2020-08-27 paper_2020-08-19.tex
big update

@@
 Forecasts provide important data to decision-makers and the general public and can improve situational awareness of current trends and how they will continue in coming weeks.
+
 Infectious disease forecasting, at the time horizon of up to 4 weeks in the future, has benefited public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.
@@
 We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.
-Implementation in a Bayesian framework also allows for implementation in a cutting edge probabilistic programming framework for speed and accuracy.
+Implementation in a Bayesian framework also allows for use of a cutting edge probabilistic programming framework for speed and rapid model development.
 
-Our main goal in this work is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting.
-However, our m We are not interested in identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data.
+Our main goal in developing MechBayes is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting.
+Our focus is not on identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data.
 We are willing to accept biologically unrealistic parameter values or priors for latent variables; but not ones that lead to gross pathologies during inference and forecasting.
+The non-parametric extensions to the underlying SEIR model also mean that any inference on epidemiological parameters should be carried out very carefully.
 
- MechBayes is able to account for these operational concerns when forecasting in real-time.
-We demonstrate the success of the model in both forecast submissions submitted to the US Centers for Disease Control via the Covid19-ForecastHub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
-In what follows we first describe the available data and forecast submission infrastructure, outline the basic susceptible-exposed-infected-recovered (SEIR) compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
+ We demonstrate the success of the model in both forecast submissions submitted to the US Centers for Disease Control via the COVID-19 Forecast Hub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
+In what follows we first describe the available data and forecast submission infrastructure, outline the basic SEIR compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
 
@@
 
-%\begin{itemize}
-%
-%\item Resource allocation
-%\item Intervention Planning
-%\item Disease burden
-%\end{itemize}
-%\item Cite Flu Forecasting
-%\item Describe Covid19-ForecastHub
-%\begin{itemize}
-%\item Number of participating teams as evidence of importance
-%\item Soliciting forecasts for incident and cumulative deaths
-%\end{itemize}
-%\item Mechanistic models have been around since K\&K
-%\item Demonstrated success in modeling infectious disease 
-%\item Since they were developed around the last pandemic their usefulness as applied forecasting models has gone untested
-%\item Extensions to basic model needed 
-%\begin{itemize}
-%\item Observations on cases and deaths
-%\item Time-varying case and death deviation model for varied testing
-%\item Non-parametric model of interventions
-%\item Full Bayesian fitting due to unidentifiability of parameters given only a time series of cases and deaths.
-%\end{itemize}
-%\item MechBayes outperforms in two settings
-%\begin{itemize}
-%\item Covid19-ForecastHub relative to baseline model 
-%\item Ablation test relative to basic SEIR model
-%\end{itemize}
-%\item We show improvements in MAE and WIS relative to baseline model in both experimental setups
-%
-%\end{itemize}
-
 \section{Data}
-	In this analysis we use confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}.
+	In this analysis we used confirmed case counts and deaths as reported by the Johns Hopkins University Center for Systems Science and Engineering \cite{dong2020interactive}.
 This a time series dataset which we truncate to begin March 1st 2020 to August 1st 2020 and captures all 50 states, as well as Guam, Puerto Rico, American Samoa, District of Columbia, Northern Mariana Islands and U.S.
@@
 A two-week ahead corresponds two the next Saturday and so on.
-Forecasts were evaluated using relative mean-absolute-error (relMAE) and the weighted-interval score (WIS) \cite{bracher2020evaluating}.
-Relative mean absolute error is the ratio of the mean absolute error of the model of interest divided by the baseline model \cite{reich2016case}.
-The WIS is an approximation to the continuously ranked probability score (CRPS), a commonly used metric in probabilistic forecasting \cite{bracher2019multibin}\cite{reich2019reply}.
+Forecasts were evaluated using mean-absolute-error (MAE) and empirical coverage probability at the 95\% level.
+This allows us to assses both point prediction and calibration of forecasts.
 
@@
  To evaluate our model, we examine two different scenarios.
-First, we describe the submission process and infrastructure used for the real time evaluation as part of the Covid19-ForecastHub consortium.
+First, we describe the submission process and infrastructure used for the real time evaluation as part of the COVID-19 Forecast Hub consortium.
 Second, we describe the internal evaluation used to demonstrate our model enhancements improve accuracy over a naive compartmental model.
@@
 The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
-We use the model submissions made in real-time evaluated on both relMAE and WIS for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
+We use the model submissions made in real-time evaluated on both MAE and coverage probability for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
 Note that not all targets are observed at all weeks.
 This is due to 4 week ahead targets for weeks 2020-07-13 and beyond not being observable by 2020-08-01.
+We subset to the 50 states and Washington D.C., which is the largest set of locations where forecasts were made for each date.
 
@@
 % \end{itemize}
+
+ We compare the results against the COVID-19 Forecast Hub baseline model. This is a purely statistical model that uses historical COVID-19 deaths only.
+
+\begin{eqnarray}
+\text{Deaths}_t &\sim& N(\text{Deaths}_{t-1},\hat{\sigma}^2)\\
+\hat{\sigma} &\sim& not sure
+\end{eqnarray}
+
+This model is fit state by state. This is similar to an auto-regressive order 1 model, a common time series baseline model. However, here the variance is determined from historical changes in deaths.
 
@@
   \centering
-    \includegraphics[scale=.2]{beta_t_plot.png}
+    \includegraphics[scale=.15]{beta_t_plot.png}
     \caption{Time-varying beta ($\beta(t)$).}
@@
   \centering
-    \includegraphics[scale=.2]{detection_plot.png}
+    \includegraphics[scale=.15]{detection_plot.png}
     \caption{Time-varying case and death deviation model ($p_{c,t}$)}
@@
   \centering
-    \includegraphics[scale=.135]{wis_results_by_time_zero_inc.png}
-    \caption{WIS by forecast week}
+    \includegraphics[scale=.115]{cp_results_by_time_zero.png}
+    \caption{Coverage probability by forecast week}
 \end{subfigure}
@@
   \centering
-    \includegraphics[scale=.135]{wis_results_by_target_inc.png}
-    \caption{WIS by target}
+    \includegraphics[scale=.115]{cp_results_by_target.png}
+    \caption{Coverage probability by target}
 \end{subfigure}%
 
-\caption{Scores from Covid19-ForecastHub broken down by target and forecast week.
+\caption{Scores from COVID-19 Forecast Hub broken down by target and forecast week.
 We use the relative MAE metric which takes the MAE of MechBayes model and divides by the MAE of the baseline model.
-Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
-Finally, we can see that as the horizon increases from 1 to 4 weeks ahead, both relMAE and WIS increase, reflecting an increase in difficulty of forecasting further ahead in time.
+Here we can see that the MechBayes model improves in both MAE and coverage probability over time, consistently beating the baseline model in the month of July 2020.
+We can see that as the horizon increases from 1 to 4 weeks ahead, MAE, reflecting an increase in difficulty of forecasting further ahead in time.
 Note that targets averaging over forecast week include an average over all model versions.
+We also see a uniform improvement in coverage probability when broken down by timezero and target.
+Here the horizontal line represents 95\% coverage interval.
 }
@@
   \centering
-    \includegraphics[scale=.20]{wis_results_by_region_inc.png}
-    \caption{WIS by region}
+    \includegraphics[scale=.20]{cp_results_by_location.png}
+    \caption{Coverage probability by region}
 \end{subfigure}
-\caption{Scores from Covid19-ForecastHub broken down by region for the 20 regions with the highest incident deaths.
+\caption{Scores from COVID-19 Forecast Hub broken down by region for the 20 regions with the highest incident deaths.
 Regions are sorted by total deaths left to right.
@@
 States with medium or small death counts have more mixed results.
-The trends are similar for WIS, with an even more pronounced improvement for states with large death counts.
+The improvements in coverage probability are more mixed, with MechBayes improving coverage probability in 12 of 20 regions.
 }
@@
 
-We next turn to the comparison of MechBayes against the Covid19-ForecastHub baseline model.
+We next turn to the comparison of MechBayes against the COVID-19 Forecast Hub baseline model.
 This baseline model uses the previous daily incident as the mean forecast for the current daily incidence, along with  bootstrapped prediction intervals from historical changes in daily incidence.
-See Covid19-ForecastHub for more details.
+See COVID-19 Forecast Hub for more details.
 \cite{covidhub}.
@@
 We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target.
-As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes V1 outperformed the Covid19-ForecastHub baseline on MAE or WIS when broken down by forecast week.
-The same is true for MechBayes V2 with the exception of June 26th, where the Covid19-ForecastHub baseline model slightly outperformed MechBayes V2.
+As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes outperformed the COVID-19 Forecast Hub baseline on MAE and coverage probability when broken down by forecast week.
 As we can see from \ref{fig:data}, in weeks with the largest increase in deaths (mostly during the month of May) MechBayes significantly outperformed the baseline model.
@@
 
-\begin{eqnarray}
-log(MAE_{m,t,r,h} +\delta) &=& \beta_0 +  \beta1*h_1 + ....+ \beta_3*h_3 \\
+\begin{eqnarray*}
+log(MAE_{m,t,r,h} +1) &=& \beta_0 +  \beta1*h_1 + ....+ \beta_3*h_3 \\
 &+& \beta_4 *h_1*mb + .... \beta_8*h_4*mb\\
@@
 \epsilon &\sim& N(0,\sigma^2)
-\end{eqnarray}
+\end{eqnarray*}
 
@@
   \hline
-$\beta_0$  & 1.96 & 0.24 & 56.74 & 8.24 & $0.00^*$ \\ 
- $\beta_{h_1}$  & 0.45 & 0.10 & 4643.00 & 4.66 & $0.00^*$ \\ 
-   $\beta_{h_2} $ & 0.61 & 0.10 & 4643.00 & 6.33 & $0.00^*$ \\ 
-  $\beta_{h_3}$ & 0.75 & 0.10 & 4643.00 & 7.63 & $0.00^*$ \\ 
-$\beta_{h_1 mb}$ & -0.18 & 0.10 & 4643.00 & -1.88 & 0.06 \\ 
-  $\beta_{h_2 mb}$& -0.37 & 0.10 & 4643.00 & -3.85 & $0.00^*$ \\ 
-  $\beta_{h_3mb}$& -0.27 & 0.10 & 4643.00 & -2.78 & $0.01^*$ \\ 
- $\beta_{h_4mb}$& -0.37 & 0.10 & 4643.00 & -3.68 & $0.00^*$ \\ 
+$\beta_0$  & 2.47 & 0.17 & 54.87 & 14.97 & 0.00 \\ 
+   $\beta_{h_1}$ & 0.30 & 0.06 & 4643.00 & 5.18 & 0.00 \\ 
+ $\beta_{h_2} $ & 0.44 & 0.06 & 4643.00 & 7.56 & 0.00 \\ 
+  $\beta_{h_3}$ & 0.59 & 0.06 & 4643.00 & 9.79 & 0.00 \\ 
+  $\beta_{h_1 mb}$  & -0.19 & 0.06 & 4643.00 & -3.19 & 0.00 \\ 
+  $\beta_{h_2 mb}$ & -0.29 & 0.06 & 4643.00 & -4.87 & 0.00 \\ 
+   $\beta_{h_3mb}$ & -0.23 & 0.06 & 4643.00 & -3.94 & 0.00 \\ 
+ $\beta_{h_4mb}$ & -0.29 & 0.06 & 4643.00 & -4.72 & 0.00 \\ 
    \hline
@@
 \end{table}
-%
-%\begin{itemize}
-%\item MechBayes is almost always better than baseline model when broken down by region, target, and forecast week after June 1st 2020.
-%\item MechBayes has improved over time.
-%\begin{itemize}
-%\item First two submissions did not have case and death deviation model random walk
-%\item Observations were Normally distributed on cumulative deaths
-%\item Switched to current model
-%\end{itemize}
-%\item Naive baseline is hard to beat
-%\item MechBayes is biased high. This comes from uncertainty in the random walk leading to potentially huge growth rates. 
-%\end{itemize}   
-%
 
@@
 Since overall prevalence in any state is well below the number needed to reach herd immunity, when the model sees an increase in cases it treats this as exponential growth, which translates into exponential growth in deaths.
-However, in the recent forecasts the bias is consistently lower than the Covid19-ForecastHub baseline model.
+However, in the recent forecasts the bias is consistently lower than the COVID-19 Forecast Hub baseline model.
 Note the extreme low bias in the first submission is due to the omission of the U.S.
@@
 
-However, using relatively simple non-parametric methods we were able to beat the Covid19-ForecastHub baseline model in both point and probabilistic forecast evaluations.
+However, using relatively simple non-parametric methods we were able to beat the COVID-19 Forecast Hub baseline model in both point and probabilistic forecast evaluations.
 Our ablation tests show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.
@@
 
-\subsection{A2}
-
-The Covid19-ForecastHub began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
+%\subsection{A2}
+%
+% 
+%The COVID-19 Forecast Hub began soliciting forecasts in the beginning of April 2020 for 1-4 week ahead cumulative deaths.
 We began submitting the first version on April 20th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
@@
 Our model went through three distinct iterations as we evaluated performance in real-time.
-
- \begin{itemize}
-
- \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
+% 
+% \begin{itemize}
+% 
+% \item \textbf{Version 1: April 20th 2020 through May 10th 2020}.
 Model had normal observation noise (instead of negative binomial) and a non-time varying case and death deviation model.
 Model was fit to cumulative deaths and cases.
- \begin{equation}
-\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
-\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
-\end{equation}
-
-  \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
+% \begin{equation}
+%\text{Cases}_{t} \sim N(p_{c}*C_{t} ,\sigma_{c}^2) \ \ \  
+%\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+%\end{equation}
+%
+%  \item \textbf{Version 2: May 10th 2020 through May 24th 2020}.
 Model had normal observation noise (instead of negative binomial) and time varying case and death deviation model.
 Model was fit to cumulative deaths and cases.
-
-   \begin{equation}
-\text{Cases}_{t} \sim N(p_{t,c}*C_{t} ,\sigma_{c}^2) \ \ \  
-\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
-\end{equation}
-
-    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}.
+%  
+%   \begin{equation}
+%\text{Cases}_{t} \sim N(p_{t,c}*C_{t} ,\sigma_{c}^2) \ \ \  
+%\text{Deaths}_{t} \sim N(p_d*D_{2_{t}} , \sigma_d^2)
+%\end{equation}
+%
+%
+%    \item \textbf{Version 3: May 24th 2020 through August 1st 2020}.
 Model had negative binomial observation noise  and time varying case and death deviation model.
 Model was fit to incident deaths and cases.
-     \begin{equation}
-\text{Cases}_{t} \sim NB(p_{c,t}*[C_{t}-C_{t-1}] ,\sigma_{c}^2) \ \ \  
-\text{Deaths}_{t} \sim NB(p_d*[D_{2_{t}} - D_{2_{t-1}}] , \sigma_d^2)
-\end{equation}
-
- \end{itemize}
-
- These three versions highlight the complexities of forecasting during a real-time pandemic.
+%     \begin{equation}
+%\text{Cases}_{t} \sim NB(p_{c,t}*[C_{t}-C_{t-1}] ,\sigma_{c}^2) \ \ \  
+%\text{Deaths}_{t} \sim NB(p_d*[D_{2_{t}} - D_{2_{t-1}}] , \sigma_d^2)
+%\end{equation}
+%    
+% \end{itemize} 
+% 
+% 
+% These three versions highlight the complexities of forecasting during a real-time pandemic.
 Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
@@
 This is due to 4 week ahead targets for weeks 2020-07-12 and beyond not being observable by 2020-08-01.
-
- These two versions highlight the complexities of forecasting during a real-time pandemic.
+%  
+% 
+% These two versions highlight the complexities of forecasting during a real-time pandemic.
 Models evolve due to real-time evaluations by responding to the unique data collection environments of each pandemic.
-
-\begin{figure}
-  \centering
-     \begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{mae_results_by_time_zero.png}
-    \caption{MAE by forecast week}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
-    \caption{WIS by forecast week}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{mae_results_by_region.png}
-    \caption{MAE by region}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{wis_results_by_region.png}
-    \caption{WIS by region}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{mae_results_by_target.png}
-    \caption{MAE by target}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.1]{wis_results_by_target.png}
-    \caption{WIS by target}
-\end{subfigure}%
-\section{Real-Time Model Results}
-
-\caption{Scores from Covid19-ForecastHub broken down by region, target and forecast week.
+%
+%
+%\begin{figure}
+%  \centering
+%     \begin{subfigure}{.5\textwidth}
+%  \centering
+%    \includegraphics[scale=.1]{mae_results_by_time_zero.png}
+%    \caption{MAE by forecast week}
+%\end{subfigure}%
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%    \includegraphics[scale=.1]{wis_results_by_time_zero.png}
+%    \caption{WIS by forecast week}
+%\end{subfigure}
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%    \includegraphics[scale=.1]{mae_results_by_region.png}
+%    \caption{MAE by region}
+%\end{subfigure}%
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%    \includegraphics[scale=.1]{wis_results_by_region.png}
+%    \caption{WIS by region}
+%\end{subfigure}
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%    \includegraphics[scale=.1]{mae_results_by_target.png}
+%    \caption{MAE by target}
+%\end{subfigure}%
+%\begin{subfigure}{.5\textwidth}
+%  \centering
+%    \includegraphics[scale=.1]{wis_results_by_target.png}
+%    \caption{WIS by target}
+%\end{subfigure}%
+%\section{Real-Time Model Results}
+%
+%\caption{Scores from COVID-19 Forecast Hub broken down by region, target and forecast week.
 Here we can see that the MechBayes model improves in both MAE and WIS over time, consistently beating the baseline model in the month of July 2020.
@@
 }
-\label{fig:covidhub_appendix}
-\end{figure}
-
+%\label{fig:covidhub_appendix}
+%\end{figure}
+%
+%
 \newpage 

### 2020-08-27 analysis.R

@@ -9,8 +9,15 @@ library(sjstats) #use for r2 functions
 
 np <- import("numpy")
 files = list.files("/Users/gcgibson/covid-intervention-analysis/samples/")
-files <- states_to_plot
+files <- state_deaths_subset_idx
+files <- files[files !="US"]
+files <- state.abb[match(files,state.name)]
+
+
 beta <- matrix(NA,ncol=length(files),nrow=137 )
+beta_upper <- matrix(NA,ncol=length(files),nrow=137 )
+beta_lower <- matrix(NA,ncol=length(files),nrow=137 )
+library(matrixStats)
 col_idx <- 1
 states = c()
 for (file in files){
@@ -18,29 +25,50 @@ for (file in files){
     print (file)
     npz1 <- np$load(paste0("/Users/gcgibson/covid-intervention-analysis/samples/",file,".npz"),allow_pickle = T)
     beta[,col_idx] = colMeans(npz1$f$mcmc_samples[[1]]$beta)
+    beta_upper[,col_idx] = t(rowQuantiles(t(npz1$f$mcmc_samples[[1]]$beta),probs = .975))
+    beta_lower[,col_idx] = t(rowQuantiles(t(npz1$f$mcmc_samples[[1]]$beta),probs = .025))
+    
     states <-c(states,substr(file,1,2))
     col_idx <- col_idx + 1
   }
 }
 
-X <- read.csv(url("https://raw.githubusercontent.com/CEIDatUGA/COVID-19-DATA/master/US/us-state-intervention-data/US_state_intervention_time_series.csv"))
 
-data_for_lmer  = data.frame(R0 = c((beta)), t= rep(1:137,length(states)), state=rep(states,each=137))
-data_for_lmer$state <- as.factor(data_for_lmer$state)
+data_for_lmer  = data.frame(R0 = c((beta)),
+                            R0_upper = c((beta_upper)),
+                            R0_lower = c((beta_lower)),
+                            t= rep(1:137,length(states)), state=rep(states,each=137))
+#data_for_lmer$state <- as.factor(data_for_lmer$state)
 data_for_lmer$date <- rep(seq(as.Date("2020-03-05"),as.Date("2020-07-19"),"days"),length(states))
 library(openintro)
 
 library(stringr)
+regions_to_show <- c("TX", "FL", "CA", "NY", "NJ", "IL", "PA", "MA", "AZ", "GA", 
[…]

### 2020-08-27 data.R

@@ -27,6 +27,7 @@ state_deaths_subset$location_name <- as.factor(state_deaths_subset$location_name
 state_deaths_subset <- state_deaths_subset[order(state_deaths_subset$total_deaths),]
 state_deaths_subset$location_name <- factor(state_deaths_subset$location_name,levels = unique(state_deaths_subset$location_name))
 data_plot <- ggplot(state_deaths_subset,aes(x=date,y=incident)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free") + theme_bw() +ylab("Incident Reported COVID-19 Deaths (per day)") + xlab("")
+#data_plot <- data_plot + ylab("Incident Deaths") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=10,height=10)
 
 
@@ -79,6 +80,7 @@ plot_df_subset[is.na(plot_df_subset$location_name), ]$location_name <- "US"
 fit_and_forecast_results <- ggplot(plot_df_subset,aes(x=time,y=median)) + geom_line(col="cornflowerblue") +
   geom_ribbon(aes(ymin=lower, ymax=upper),alpha=0.2,size=0,colour="cornflowerblue")+geom_point(data=state_deaths_subset_for_addition,aes(x=date,y=incident),size=.5)  +facet_wrap(~location_name,scales='free') +
   theme_bw() + geom_vline(xintercept = as.Date("2020-07-26"),alpha=.75)
+fit_and_forecast_results <- fit_and_forecast_results + ylab("Incident Deaths") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ggsave("/Users/gcgibson/mech_bayes_paper/fit_and_forecast_results.png",fit_and_forecast_results,device="png",width=8,height=8)
 
 

### 2020-08-27 model_rank_test.R

@@ -37,7 +37,7 @@ library(nlme)
 
 
 #diagnostic_fit <- update(diagnostic_fit, correlation = corAR1())
-diagnostic_fit <- lmer(log(mae+.01)~ factor(target) + factor(model):factor(target) +  ( 1   | unit),data=mae_results_by_time_zero_df)
+diagnostic_fit <- lmer(log(mae+1)~ factor(target) + factor(model):factor(target) +  ( 1   | unit),data=mae_results_by_time_zero_df)
 
 summary(diagnostic_fit)
 library(sjPlot)

### 2020-08-27 score_calibration.R

@@ -3,100 +3,94 @@ library(tidyverse)
 library(lubridate)
 zoltar_connection <- new_connection()
 zoltar_authenticate(zoltar_connection, "gcgibson", "casey!ili")
-## read in fips
-fips_csv <- read.csv(url("https://raw.githubusercontent.com/reichlab/covid19-forecast-hub/master/data-locations/locations.csv"))
 
-## access scores for the COVID-19 Forecasts project
-#tmp <-  download_forecast(zoltar_connection, "https://www.zoltardata.com/api/model//")   
-tmp <- do_zoltar_query(zoltar_connection, "https://www.zoltardata.com/api/project/44/", c("MechBayes"))
 
-tmp <- tmp[tmp$target %in% paste0(1:4, " wk ahead inc death"),]
-tmp_mech_bayes <- tmp[tmp$model %in%  c("UMass-MechBayes","COVIDhub-baseline"),]
-tmp_mech_bayes$timezero <- as.Date(tmp_mech_bayes$timezero)
-#tmp_mech_bayes <- tmp_mech_baye
-tmp_subset <- tmp_mech_bayes %>%  dplyr::group_by(timezero,unit,model,target,truth) %>% dplyr::summarize(mae=mean(mae),wis=mean(wis),error=mean(error))
-states_ <- unique(tmp_subset[tmp_subset$model == "UMass-MechBayes",]$unit) 
-states_ <- states_[states_!="US" & states_ != "60" & states_ != "66"&states_ != "69" & states_ != "78" & states_ != "72"]
-tmp_subset <- tmp_subset[tmp_subset$unit %in% states_,]
+tmp <- scores(zoltar_connection, "https://www.zoltardata.com/api/project/44/") %>%
+  mutate(truth=truth)
 
-#tmp_subset$timezero <- as.factor(tmp_subset$timezero)
-tmp_subset <- tmp_subset[tmp_subset$timezero > "2020-05-12",]
-tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero <- tmp_subset[tmp_subset$model == "COVIDhub-baseline",]$timezero -1
+
+mb_prefix <- "/Users/gcgibson/mech_bayes_paper/UMass-MechBayes/"
+baseline_prefix <- "/Users/gcgibson/mech_bayes_paper/COVIDhub-baseline/"
+
+baseline_files <- list.files(baseline_prefix)
+mb_files <- list.files(mb_prefix)
[…]

### 2020-08-27 paper_2020-08-26.tex
big update

--- 2020-08-27 paper_2020-08-19.tex
@@
 However, data quality during a pandemic suffers from under-reporting, delayed reporting due to reporting infrastructure issues, and limited testing.
-Traditionally, compartmental models have been used to model emerging pandemics, however, the basic compartmental framework is difficult to operationalize in a complex data environment.
-We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize Covid-19 in forecasting in real time.
-This includes non-parametric modeling of varying transmission rates, non-parametric modelling of case detection versus death detection, and finally joint observation likelihood on new case counts and new deaths.
+Traditionally, compartmental models have been used to model emerging pandemics.
+However, the basic compartmental framework is difficult to operationalize in a complex data environment.
+We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize COVID-19 forecasting in real time.
+This includes non-parametric modeling of varying transmission rates, non-parametric modeling of case and death discrepancies due to testing and reporting issues, and finally joint observation likelihood on new case counts and new deaths.
 The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab.
@@
 In this work, we introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19.
-These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting data dumps, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
-The case and death data is reported as new cases, instead of prevalent, and new deaths, instead of cumulative, which we handle using an observation model on top of the underlying SEIR dynamics.
-Finally, we choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting.
-We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.
+These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
+We accomplish this by modeling transmission and deviations from cases and deaths (beyond the case fatality ratio) non-parametrically.
+Our observation model is able to handle case data which is reported as new cases, instead of prevalent, and new deaths, instead of cumulative.
+We choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting.
 Implementation in a Bayesian framework also allows for use of a cutting edge probabilistic programming framework for speed and rapid model development.
@@
 The non-parametric extensions to the underlying SEIR model also mean that any inference on epidemiological parameters should be carried out very carefully.
+We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.
 
- We demonstrate the success of the model in both forecast submissions submitted to the US Centers for Disease Control via the COVID-19 Forecast Hub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
+ We demonstrate the success of the model in both forecast submissions to the US Centers for Disease Control via the COVID-19 Forecast Hub as well as an ablation model comparison to demonstrate the additional forecast accuracy of our extensions beyond the basic SEIR model.
 In what follows we first describe the available data and forecast submission infrastructure, outline the basic SEIR compartment model, describe our extensions for real-world pandemic forecasting, and finally evaluate the model using both real-time evaluation from submissions and a retrospective model component analysis.
@@
 		There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}).
-For example, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include \"probable" deaths as well as confirmed deaths due to COVID-19 on June 25th 2020.
-However, many other states have at least one outlying value, usually due to backlog reporting, where a large number of deaths occurring in previous weeks are reported on a single day.
+For example, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include ``probable" deaths as well as confirmed deaths due to COVID-19 on June 25th 2020.
+However, many other states have at least one outlying value, usually due to delayed reporting, where a large number of deaths occurring in previous weeks are reported on a single day.
 We can also see from Figure \ref{fig:data} that some states do not report on particular days (usually weekends) leading to 0 incident deaths for the day.
@@
 An individual forecast distribution is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
+Since our model produced cumulative forecasts by aggregating incident, we choose to only evaluate incident.
 The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$, with the median ($.5$ quantile) representing the point-forecast.
@@
 Notice the large variability in incident reporting.
-There are significant data-dumps, with New Jersey reporting over 1500 backlogged deaths when switching to include "probable" deaths.
+There are significant delayed reporting, with New Jersey reporting over 1500 backlogged deaths when switching to include ``probable" deaths.
 In some states, there appears to be a weekly cycle where deaths are under-reported on the weekend.
@@
 as a whole.
-We can also see that there are some negative incident deaths, where data are revised to account for deaths that were incorrectly attributed to COVID.
+We can also see that there are some negative incident deaths, where data are revised to account for deaths that were incorrectly attributed to COVID-19.
 }
@@
 In a given time-step (e.g. one day), each member of the population of a single geography belongs to one of the following mutually exhaustive compartments:
- Susceptible $S$, Exposed but not yet infectious $E$, Infectious $I$, Recovered $R$, hospitalized before death $D_1$ , and  deceased $D_2$ (Figure \ref{fig:seird}).
-Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature \cite{russell2020estimating}.
+ susceptible $S$, exposed but not yet infectious $E$, infectious $I$, recovered $R$, hospitalized before death $D_1$, and deceased $D_2$ (Figure \ref{fig:seird}).
+Here we assume everyone who is hospitalized will eventually become deceased in order to separate the rate of flow into both a case fatality ratio (CFR) parameter as well as a time from symptoms to death parameter, which both have prior estimates from the literature \cite{russell2020estimating}.
 We omit an explicit hospitalization compartment since the available hospitalization data is highly variable by state and suffers even more reporting issues than case data.
@@
 This model assumes that forecasts are made on the current level of interventions because $\mathbb{E}[log(\beta(t+1))] = log(\beta(t))$.
-That is, the expected value of a random walk in the forecasting stage is simply $\beta(t)$ at the last observed value of $t$.
+That is, the expected value of a random walk in the forecasting stage is the estimated value at the final time point of available data.
 
@@
 For a given state and day, the change in the  confirmed cases and reported deaths are subset of the cumulative number of new infections $C(t)$ and cumulative number of deaths $D2(t)$, respectively.
-Therefore, we introduce two additional parameters for the case and death deviation model of cases $p_c$ and the case and death deviation model of deaths $p_d$.
+To handle this, we introduce two additional parameters.
+First, we introduce a case and death deviation, beyond the case fatality ration, of cases $p_c$ and the case and death deviation model of deaths $p_d$.
 For both, we set fairly flat priors to reflect these parameters are poorly determined from observed data.
 
-In more detail, $p_c$ is the probability that an infected person receives a positive test result and is reported by a state or local health authority as a case.
-We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$  with 90\% probability between \[ .22,0.38\].
+In more detail, $p_c$ can be thought of as an aggregate probability of a case being detected and then flowing through the compartments.
+We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$  with 90\% probability between $0.22,0.38$.
 This means that we expect 30\% of cases to be detected initially, as suggested by the literature \cite{midas}.
 However, we also allow this to vary by time.
+We do not intend for this to be interpretable as purely reflecting testing, but rather an aggregate measure of testing, reporting issues, and general departure from our prior estimate of the case fatality ratio.
 
@@
 We also assume the probability that a COVID-19 death is reported $p_d$ has a prior distribution given by  $p_d \sim \Beta(90, 10)$.
-This prior satisfies $\E[p_d] = 0.9$ with concentration 100.
+This prior satisfies $\E[p_d] = 0.9$ with 90\% probability between 0.89,0.92.
 That is, we assume that deaths due to COVID-19 are most often correctly reported \cite{weinberger2020estimation}.
@@
  This process ensured that the observed data does not contain any identifiable outliers (meaning documented by outside sources).
-In real-time this is necessary to avoid drastic over-predictions caused by data dumps.
+In real-time this is necessary to avoid drastic over-predictions caused by delayed reporting.
 For example, New Jersey reported nearly 1,600 daily deaths as it switched from reporting only confirmed deaths from COVID-19, to confirmed and probable on 2020-06-25.
@@
 
- We compare the results against the COVID-19 Forecast Hub baseline model. This is a purely statistical model that uses historical COVID-19 deaths only.
-
-\begin{eqnarray}
-\text{Deaths}_t &\sim& N(\text{Deaths}_{t-1},\hat{\sigma}^2)\\
-\hat{\sigma} &\sim& not sure
-\end{eqnarray}
-
-This model is fit state by state. This is similar to an auto-regressive order 1 model, a common time series baseline model. However, here the variance is determined from historical changes in deaths.
+ We compare the results against the COVID-19 Forecast Hub baseline model.
+This is a purely statistical model that uses historical COVID-19 deaths only.
+At all forecast horizons, the median of the forecast distribution from the baseline model is equal to the most recent reported incidence.
+The model obtains a non-parametric distribution around this median by adding and subtracting past observed differences in incidence from one week to the next for a specific location.
+This model is fit state by state.
 
@@
  \begin{itemize}
- \item \textbf{MechBayes Full} Mech Bayes as of model version 3.
+ \item \textbf{MechBayes Full} Mech Bayes as submitted to the CDC.
 That is, a model using negative binomial observation noise as well as a time-varying random walk, using a joint likelihood over cases and deaths.
@@
  We also fix all non-model component variation.
-That is, we average over the last 10 days of $\beta(t)$ when forecasting, as well as manually redistributing data dumps.
+That is, we average over the last 10 days of $\beta(t)$ when forecasting, as well as manually redistributing delayed reporting.
 This ensures that the comparison is only on model components, and not on data discrepancies.
@@
 by 2020-07-26.
-We can clearly see that the random walk is able to non-parametrically account for the non-pharmaceutical interventions retrospectively, with transmission dropping in states like New York quite rapidly.
-While it may seem that many states follow the same trend, this is slightly misleading since the prior pulls $\beta(t)$ upward for $t$ close to 0 in states that did not experience community transmission in March.
+We can clearly see that the random walk is able to account for a composite of intervention efforts.
 B) Time varying case and death deviation model for example states.
-The time varying detection random walk is able account for the increase in testing and other reporting anomolies.
-There is remarkable similarity among the estimated case and death deviation model trajectories between states.}
+The time varying case and death deviation model is able account for the increase in testing and other reporting anomolies.}
 \label{fig:model_details}
@@
 
-We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes version 3 is able to accurately model the observed data.
+\section{Real-time Forecasting Results}
+
+We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes is able to accurately model the observed data.
 The model is able to adapt to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves.
@@
 We can see from Figure \ref{fig:model_details} that MechBayes is able to learn to adapt to the evolving pandemic situation.
-Panel A shows the time-varying transmission $\beta(t)$ for three example states, CA, FL, and NY as well as the U.S.
+Panel A shows the time-varying transmission $\beta(t)$ for 20 states with the highest total deaths.
 We can see that $\beta(0)$ is centered on our prior but as data comes in, the estimate increases.
 This is especially true in NY, where the epidemic took off quickly in March.
-However, the model is then able to adjust to the varying levels of non-pharamceutical interventions present in each of the three states as well as across the U.S.
-This radically reduces the transmissibility parameter by the first week of April 2020.
+However, the model is then able to adjust to the varying levels of interventions present in each state.
+This radically reduces the transmissibility parameter by the first week of April 2020 for almost all states.
 This is consistent with a peak in overall deaths two weeks later in mid-April 2020.
@@
 Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in case and death deviation model before climbing again.
-Note that interpreting this strictly as time-varying detection is obscuring the fact that this parameter $p_{t,d}$ can soak-up any excess variation beyond the ability of cases and the case-fatility ratio to explain the number of deaths.
+
+%Note that interpreting this strictly as time-varying detection is obscuring the fact that this parameter $p_{t,d}$ can soak-up any excess variation beyond the ability of cases and the case-fatility ratio to explain the number of deaths.
 That is, the time-varying case and death deviation model is able to "de-couple" cases and deaths beyond the case-fataility ratio regardless of the underlying reason (whether that be an increase in testing or shifting age distribution of cases).
@@
 \caption{Scores from COVID-19 Forecast Hub broken down by target and forecast week.
-We use the relative MAE metric which takes the MAE of MechBayes model and divides by the MAE of the baseline model.
-Here we can see that the MechBayes model improves in both MAE and coverage probability over time, consistently beating the baseline model in the month of July 2020.
+MechBayes improves in average MAE in 9 out of 12 timepoints.
+MechBayes improves in MAE across all targets.
 We can see that as the horizon increases from 1 to 4 weeks ahead, MAE, reflecting an increase in difficulty of forecasting further ahead in time.
-Note that targets averaging over forecast week include an average over all model versions.
 We also see a uniform improvement in coverage probability when broken down by timezero and target.
@@
 }
-\label{fig:covidhub_appendix}
+\label{fig:covidhub}
 \end{figure}
@@
 As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes outperformed the COVID-19 Forecast Hub baseline on MAE and coverage probability when broken down by forecast week.
+However, the improvements broken down by time are small, and improves in 9 out of 12 evaluation weeks.
 As we can see from \ref{fig:data}, in weeks with the largest increase in deaths (mostly during the month of May) MechBayes significantly outperformed the baseline model.
@@
 We also break down the results by geographical region, as seen in Figure \ref{fig:covidhub} (C,D).
-Note that this break-down is averaged over target (1-4 week ahead), but also forecast week, meaning that we average over each of the model versions.
-We feel this is important, as it reflects the real-time accuracy of our evolving model efforts, instead of cherry-picking the best model.
-However, we still see consistent improvements in MAE under MechBayes when broken down by region.
-We see the largest improvements in the regions where the baseline model had the highest MAE, meaning that MechBayes improves forecasts in regions with more significant viral activity.
+Here we see consistent improvements in MAE under MechBayes for regions with high total death counts with an improvement in the average MAE in 16 out of the 20 states with the highest total death count.
+Out of the states with the 10 highest death counts, only in California did the baseline model outperform MechBayes on the average MAE.
+However, the results are more mixed for states with lower death counts.
 
-We break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and WIS.
+We break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and coverage probability.
 We can also see that the MAE increase as horizon increase, which is to be expected.
@@
 
-Finally, we include a formal test of the difference in MAE and WIS to demonstrate statistically significant advantages in using MechBayes over the baseline model.
+Finally, we include a formal test of the difference in MAE to demonstrate statistically significant advantages in using MechBayes over the baseline model.
 In order to do this, we just a random effects regression model of the form,
@@
 We chose this model because it explains the variation in MAE by model and horizon while allowing varying baseline MAE values by region.
-Here, variation over time in MAE within a specific region is explained by differences in model performance This leads to the following coefficient estimates for the fixed effects,
+Here, variation over time in MAE within a specific region is explained by differences in model performance This leads to the following coefficient estimates for the fixed effects.
 
@@
 \caption{Coefficient estimates and t-values for MAE evaluation model.
-We can see that MechBayes performs statistically significantly better than the baseline model for 2-4 weeks ahead.
-The performance increase does not seem to follow a particular pattern.}
+We can see that MechBayes performs statistically significantly better than the baseline model for 1-4 weeks ahead.
+The performance increase seems to grow as horizon increases.}
 \end{table}
+
+We can see that MechBayes is significantly better for 1-4 weeks ahead.
+The difference in skill seems to improve as the forecast horizon increases.
+The model estimates $\beta_{h_4mb}$ = -0.29, which translates to a multiplicative difference of $e^{-.29}=.75$ on the MAE scale.
+For example, if the basline model MAE for a particular 4 step ahead forecast is 30, one would expect the MechBayes forecast to have an MAE of 22.5.
+We demonstrate that the mixed model proposed above fits the MAE results reasonably in Appendix 2.
 
@@
 We can also see that the MechBayes Death Only is consistently better than MechBayes Fixed case and death deviation model.
-This may be evidence that naively including case data, without adjusting for time-varying testing rates, may be worse than not including it at all.
+This may be evidence that naively including case data, without adjusting for discrepancies between cases and deaths due to testing etc., may be worse than not including it at all.
 
@@
 \caption{Ablation test MAE broken down by forecast week and target.
-We can see that MechBayes Full performs better than MechBayes No Detection Random Walk and MechBayes Only death in almost all breakdowns.
+We can see that MechBayes Full performs better than MechBayes Fixed Case/Death Deviation and MechBayes Only death in almost all breakdowns.
 We can also see that the improvement becomes more pronounced at larger horizons, suggesting that MechBayes Full is a better longer term forecasting model.
@@
 \begin{itemize}
+
+\item \textbf{MechBayes is accurate when compared to a baseline model}.
+As Figure \ref{fig:covidhub} shows, MechBayes improved relative to the baseline model when broken down by timezero and target.
+The results are more pronounced when breaking down by target, where the boxplots (95\%) are almost completely separated.
+These results are statistically significant for 1-4 weeks ahead, with larger improvements at later horizons, suggesting that the biggest improvements from MechByes over the baseline model comes in longer term forecasting.
+Additionally, as seen in Figure \ref{fig:covidhub_region_results} , the biggest gains in performance in MAE occur in regions with the largest total deaths counts.
+This is a desirable feature of a pandemic forecasting model.
+
+\item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
+As we can see from Figure \ref{fig:covidhub} (B,D), the coverage probability of MechBayes is much closer to 95\%.
+When averaging over target, region, and timezero the coverage probability is 94.6\%.
+This suggests that MechBayes is very well calibrated at the 95\% level.
+The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise.
+The model is able to learn reasonable estimates of these variance parameters to produce well calibrated results.
+
 \item \textbf{Adding case data when predicting deaths is helpful but only when accounting for data quality issues}.
 Our ablation tests (Figure \ref{fig:ablation}) clearly shows that time-varying case and death deviation model is a key feature in the model for reducing MAE of forecasts.
-The "Full" model that both incorporates incident cases into the model likelihood and allows for a flexible time-varying case detection rate is consistently more accurate than a model that does not account for cases at all ("Death Only" model) and a model that does account for cases but does not account for a time-varying case and death deviation model.
+The "Full" model that both incorporates incident cases into the model likelihood and allows for a flexible deviation between cases and deaths is consistently more accurate than a model that does not account for cases at all ("Death Only" model) and a model that does account for cases but does not account for a time-varying deviation between cases and deaths.
 
@@
 This is visible in Figure \ref{fig:model_details}, where the time-varying case and death deviation model in New Jersey follows the typical trend of other states, until the data dump, where the case and death deviation model dips significantly down.
-
-\item \textbf{MechBayes is accurate when compared to a baseline model}.
-As Figure \ref{fig:covidhub} shows, MechBayes improved relative to the baseline model when broken down by timezero and target.
-The results are more pronounced when breaking down by target, where the boxplots (95\%) are almost completely separated.
-These results are statistically significant for 2-4 weeks ahead, suggesting that the biggest improvements from MechByes over the baseline model comes in longer term forecasting.
-Additionally, as seen in Figure \ref{fig:covidhub_region_results} , the biggest gains in performance, in both MAE and WIS, occur in regions with the largest total deaths counts.
-This is a desirable feature of a pandemic forecasting model.
 
@@
 
-\item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
-Small WIS values are better in terms of calibration \cite{bracher2020evaluating}.
-The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise, unlike a fully stochastic model that has inherent variability within the differential equation transitions.
-This suggest that even a deterministic model core can produce well calibrated prediction intervals by capturing the uncertainty in parameter estimation and observation noise.
-
-\item \textbf{Allowing for time-varying transmissibility is necessary to non-parametrically capture the effect of non-pharamceutical interventions}.
+\item \textbf{Allowing for time-varying transmissibility is necessary to non-parametrically capture the effect of  interventions}.
 Our ablation test explicitly did not include a model that fixed $\beta$ across time.
@@
 
-\item \textbf{MechBayes Full outperforms the other versions on probabilistic scoring measures}.
-For simplicity, we only included MAE results for the ablation test.
-However, as can be seen from Figure \ref{fig:covidhub}, MAE and WIS are highly correlated.
-We therefore conclude that MechBayes Full is also better calibrated than the other two competing models.
 \end{itemize}
@@
 Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
+\subsection{A2}
+\textbf{is this needed}
 

### 2020-08-27 ablation.R

@@ -6,22 +6,22 @@ library(ggthemes)
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case Detection Probability","Death Only")
+levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
   
 ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
-         summarize(se = sqrt(var(mae)),mae_t=mean(mae)),aes(x=as.factor(forecast_date),y=mae_t,col=model)) + geom_point(size=1.5) +theme_bw() + ylab("MAE") + xlab("Date")  + scale_color_brewer(palette="Dark2") + ylim(0,800) + theme(legend.position = "none")  
+         summarize(se = sqrt(var(mae)),mae_t=mean(mae)),aes(x=as.Date(forecast_date),y=mae_t,col=model)) + geom_point(size=1.5) +theme_bw() + ylab("MAE") + xlab("Date")  + scale_color_brewer(palette="Dark2") + ylim(0,800) + theme(legend.position = "none")  
 ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png",width=4,height=6)
 
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(13),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case Detection Probability","Death Only")
+levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case Detection Probability","Death Only")
+levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
[…]

### 2020-08-27 analysis.R

@@ -94,7 +94,9 @@ data_for_detection  = data.frame(R0 = as.numeric(c((det_probs))),
                                 R0_lower = as.numeric(c((det_probs_lower))),                     
                                  t= rep(1:137,length(states)), state=rep(states,each=137))
 
-data_for_detection$state <- as.factor(data_for_detection$state)
+#data_for_detection$state <- as.factor(data_for_detection$state)
+data_for_detection$state <- factor(data_for_detection$state,levels=regions_to_show)
+
 data_for_detection$date <- rep(seq(as.Date("2020-03-05"),as.Date("2020-07-19"),"days"),length(states))
 detection_plot <- ggplot(data_for_detection, aes(x=date,y=R0,group=state)) + geom_line() + facet_wrap(~state) + 
   theme_bw() + ylab("Detection Probability") + xlab("Date") #+ scale_y_discrete(breaks=seq(0,1,by=.1))

### 2020-10-20 paper_2020-10-20.tex
major update

--- 2020-08-27 paper_2020-08-26.tex
@@
 \title{Mechanistic Bayesian Forecasts of COVID19}
-\author{Graham C. Gibson, Nicholas Reich, Dan Sheldon}
+\author{Graham C. Gibson, Nicholas G. Reich, Daniel Sheldon}
 %\date{}                                           % Activate to display a given date or no date
@@
 \begin{document}
+\tableofcontents
 
@@
 Effective modeling of the course of the pandemic can help assist with public health resource planning, intervention efforts, and vaccine clinical trials.
-However, data quality during a pandemic suffers from under-reporting, delayed reporting due to reporting infrastructure issues, and limited testing.
-Traditionally, compartmental models have been used to model emerging pandemics.
-However, the basic compartmental framework is difficult to operationalize in a complex data environment.
+However, building applied forecasting models presents unique challenges during a pandemic.
+First, data quality suffers from under-reporting, delayed reporting due to reporting infrastructure issues, and limited testing.
+Second, interventions are time-varying across different geographies leading to large changes in transmissibility over the course of the pandemic.
+Finally, the unidentifiability of the epidemiological parameters leads to difficulty in model fitting.
 We propose a novel Bayesian compartmental model (MechBayes) that builds upon the classic compartmental framework of susceptible-exposed-infected-recovered (SEIR) model to operationalize COVID-19 forecasting in real time.
 This includes non-parametric modeling of varying transmission rates, non-parametric modeling of case and death discrepancies due to testing and reporting issues, and finally joint observation likelihood on new case counts and new deaths.
-The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID19-ForecsatHub repository organized by the Reich Lab.
+The model has been used to submit forecasts to the US Centers for Disease Control, through the COVID-19 ForecastHub \cite{covidhub}.
 We examine the performance relative to a baseline model in addition to performing an ablation test of our extensions to the classic SEIR models.
@@
 
-The emergence of COVID-19 in early 2020 in the United States developed into the largest pandemic the country has seen in over a century.
-Understanding the future trajectory of the pandemic is crucial for minimizing the impact across the nation in terms of healthcare burden, economic impact, treatments and public health response.
-Forecasts of incident and cumulative deaths due to COVID may help in resource allocation, vaccine clinical trial planning, and re-opening strategies.
+The emergence of COVID-19 in early 2020 in the United States led to the largest pandemic in over a century.
+Understanding the future trajectory of the impact in terms of healthcare burden, economic, treatments and public health response.
+Forecasts of incident and cumulative deaths due to COVID-19 help in resource allocation, vaccine clinical trial planning, and re-opening strategies \cite{ray2020ensemble}.
 Forecasts provide important data to decision-makers and the general public and can improve situational awareness of current trends and how they will continue in coming weeks.
 
-Infectious disease forecasting, at the time horizon of up to 4 weeks in the future, has benefited public health decision makers during annual influenza outbreaks \cite{lutz2019applying}\cite{myers2000forecasting}.
+Infectious disease forecasting, at the time horizon of up to 4 weeks in the future, has benefited public health decision makers during annual influenza outbreaks \cite{lutz2019applying,myers2000forecasting}.
 However, many forecasts of seasonal disease, such as influenza, often rely on ample historical data to look for patterns that can be projected forward into the future.
 In an emerging pandemic situation, models must be able to fit to limited data.
-In modeling the COVID-19 pandemic, many research groups have turned to the use of differential equation models to explain the underlying transmission of a disease through a population.
-First introduced by  Kermack and McKendrick, the model assumes the each individual is in one of a mutually exclusive set of compartments, typically either the susceptible, exposed, infected, or recovered compartment \cite{kermack1927contribution}.
-The model is specified by setting the rates of flow of individuals between compartments.
-While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their properties in real-time at both local and global scales.
 
-Compartmental models have been used to effectively model and forecast disease in non-pandemic situations both retrospectively and in real-time.
-These include complex compartmental models for real-time influenza forecasting \cite{shaman2012forecasting}\cite{osthus2017forecasting}\cite{ong2010real}, even including a retrospective model evaluation of the 1918 influenza pandemic \cite{hall2007real}.
-Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, measles \cite{bokler1993chaos}, dengue \cite{syafruddin2012seir} and a wide variety of other diseases
+In modeling the COVID-19 pandemic, many research groups have turned to the use of compartmental models to explain the underlying transmission of a disease through a population.
+First introduced by  Kermack and McKendrick, such models assume the each individual is in one of a mutually exclusive set of compartments, typically either the susceptible, exposed, infected, or recovered compartment \cite{kermack1927contribution}.
+A model is specified by setting the rates of flow of individuals between compartments.
+While these models have been used since their inception in the early 20th century, the COVID pandemic represents a unique opportunity to explore their forecasting properties in real-time at both local and global scales.
 
-In this work, we introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19.
+ Our main goal in developing MechBayes is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting.
+Our focus is not on inference but forecasting.
+Therefore, identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data.
+We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.
+These models, however, are often not flexible enough for real-time forecasting.
+
+We introduce a novel operational forecast model based on mechanistic foundations and tailored to the particular needs and data availability of COVID-19.
 These include, but are not limited to, severe under reporting of cases due to low testing rates especially in mild or asymptomatic cases, time-varying testing rates, delayed reporting, and both the addition and removal of control measures such as social distancing, lockdown, and mask use.
 We accomplish this by modeling transmission and deviations from cases and deaths (beyond the case fatality ratio) non-parametrically.
-Our observation model is able to handle case data which is reported as new cases, instead of prevalent, and new deaths, instead of cumulative.
 We choose a Bayesian framework that allows for uncertainty in the epidemiological parameters that are unidentifiable from the data, introducing flexibility suited to forecasting.
-Implementation in a Bayesian framework also allows for use of a cutting edge probabilistic programming framework for speed and rapid model development.
-
-Our main goal in developing MechBayes is to forecast observed incident deaths, and because we have limited historical data on COVID-19, we think compartmental models are among the most parsimonious models available for forecasting.
-Our focus is not on identifying internal parameters of the model, many of which are poorly determined or not identifiable from the available data.
-We are willing to accept biologically unrealistic parameter values or priors for latent variables; but not ones that lead to gross pathologies during inference and forecasting.
-The non-parametric extensions to the underlying SEIR model also mean that any inference on epidemiological parameters should be carried out very carefully.
-We distinguish this from scenario projection models, which require well identified epidemiological parameters that can be set to counterfactual values under different scenarios, such as an increase or decrease in intervention levels.
+Implementation in a Bayesian framework also allows for use of a cutting edge probabilistic programming framework for computational speed.
 
@@
 
-Compartmental models have also been adopted into a Bayesian framework before, including both stochastic disease dynamics and deterministic dynamics \cite{hotta2010bayesian}\cite{dukic2012tracking}.
+Compartmental models have been used to effectively model and forecast disease in non-pandemic situations both retrospectively and in real-time.
+These include complex compartmental models for real-time influenza forecasting \cite{shaman2012forecasting, osthus2017forecasting,ong2010real}, and a retrospective model evaluation of the 1918 influenza pandemic \cite{hall2007real}.
+Compartmental models have been used not just in respiratory disease but in Ebola \cite{lekone2006statistical}, measles \cite{bokler1993chaos}, dengue \cite{syafruddin2012seir} and a wide variety of other diseases
+
+Compartmental models have also been adopted into a Bayesian framework before, including both stochastic disease dynamics and deterministic dynamics \cite{hotta2010bayesian,dukic2012tracking}.
 Non-parametric transmissibility was included in a Bayesian SEIR model to study Ebola by Frasso and Lambert \cite{frasso2016bayesian}.
 Time-varying transmissibility has also been studied in the frequentist setting using complex non-parametric functions \cite{smirnova2019forecasting}.
-Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling}\cite{yang2020modified} \cite{bertozzi2020challenges}\cite{prem2020effect}\cite{flaxman2020estimating}.
+Many efforts have been made to use SEIR models in forecasting COVID-19 \cite{giordano2020modelling,yang2020modified,bertozzi2020challenges}\cite{prem2020effect,flaxman2020estimating}.
+With the outbreak of COVID-19, accounting for testing has become a critical element in effectively using an SEIR model  \cite{lopez2020modified,pei2020differential}.
 
-With the outbreak of COVID-19, accounting for testing has become a critical element in effectively using an SEIR model.
-Lopez et al.
-used a fixed case and death deviation model to model undiagnosed individuals in Spain and Italy \cite{lopez2020modified}.
-Perhaps the most similar model to ours was put forth by Pei et al.
-\cite{pei2020differential}.
-This model was an extension to the SEIR model with time varying transmission and case and death deviation model.
-However, their model fixed case and death deviation model across time and was fit using Kalman Filter techniques instead of Bayesian HMC.
-They also used the model mostly for counterfactual scenario projections instead of forecasting.
-Models that have accounted for time-varying testing have mostly used the renewal style equations \cite{abbott2020estimating}\cite{flaxman2020estimating}.
-These models do not use a differential equation model as the core component, but rather parameterizes new cases as a function of the time varying reproduction number and the serial interval.
+%*Models that have accounted for time-varying testing have mostly used the renewal style equations \cite{abbott2020estimating}\cite{flaxman2020estimating}. 
+%These models do not use a differential equation model as the core component, but rather parameterizes new cases as a function of the time varying reproduction number and the serial interval.
+
+\section{Model Overview}
+
+MechBayes extends the SEIR model framework to account for a joint observation model on incident cases and deaths, time-varying transmission, and time-varying discrepancies in the case fatality ratio due to testing, asymptomatic cases etc.
+We do this by choosing flexible non-parametric models that capture trends in disease parameters without relying on external data.
 
@@
 estimated at 20-30\% \cite{russel2020using}.
-		There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}).
+
+		%There are large discrepancies in reporting practices across the states (Figure \ref{fig:data}).
 For example, New Jersey reported an additional 1600 incident deaths when changing reporting practices to include ``probable" deaths as well as confirmed deaths due to COVID-19 on June 25th 2020.
@@
 
-We made probabilistic forecasts  for 1-4 week ahead incident and cumulative deaths for all geographies.
-An individual forecast distribution is represented by a set of quantiles for incident and cumulative deaths from 1-4 weeks ahead.
+\section{Targets}	
+We made probabilistic forecasts  for 1--4 week ahead incident and cumulative deaths for all geographies.
+An individual forecast distribution is represented by a set of quantiles, $\mathbb{Q} = {.01,.05,.10,....,.90,.95,.99}$, with the median ($.5$ quantile) representing the point-forecast.
 Since our model produced cumulative forecasts by aggregating incident, we choose to only evaluate incident.
-The quantiles used are $\mathbb{Q} = {.05,.10,....,.90,.95} \cup {.01,.99}$, with the median ($.5$ quantile) representing the point-forecast.
-Forecasts were made on Monday evenings and therefore use the incident data up until the Sunday before.
+
+%Forecasts were made on Monday evenings and therefore use the incident data up until the Sunday before.
 A one-week ahead target corresponds to the following Saturday.
@@
      \centering
-     \includegraphics[scale=.14]{data_plot.png}
+     \includegraphics[scale=.17]{data_plot.png}
      \caption{Deaths by state for any state with over 50 incident deaths for a given day.
@@
 
- \section{Compartmental Model}
+ \section{MechBayes}
 
@@
 
- \subsection{Time-varying transmission parameter}
+ \subsection{Time-varying transmission}
 
@@
 \begin{equation}
-log(\beta(t)) \sim N(log(\beta(t-1), \sigma_{\beta}^2)
+log(\beta(t)) \sim N(log(\beta(t-1)), \sigma_{\beta}^2)
 \end{equation}
@@
 To handle this, we introduce two additional parameters.
-First, we introduce a case and death deviation, beyond the case fatality ration, of cases $p_c$ and the case and death deviation model of deaths $p_d$.
+First, we introduce a case and death deviation, beyond the case fatality ration, of cases $p_{c,t}$ and the case and death deviation model of deaths $p_d$.
 For both, we set fairly flat priors to reflect these parameters are poorly determined from observed data.
 
-In more detail, $p_c$ can be thought of as an aggregate probability of a case being detected and then flowing through the compartments.
+In more detail, $p_{c,t}$ can be thought of as an aggregate probability of a case being detected and then flowing through the compartments.
 We assume its prior distribution is given by  $p \sim \Beta(15, 35)$, such that  $\E[p_c] = 0.3$  with 90\% probability between $0.22,0.38$.
@@
 
-We use relatively informative priors for epidemiological parameters, such as $\gamma$,$\sigma$, $\rho$,  $\lambda$, and initial compartment values.
+We use relatively informative priors for epidemiological parameters, such as $\gamma$, $\sigma$, $\rho$,  $\lambda$, and initial compartment values.
 The details are described in the Appendix A1.
@@
  To evaluate our model, we examine two different scenarios.
-First, we describe the submission process and infrastructure used for the real time evaluation as part of the COVID-19 Forecast Hub consortium.
+First, we describe the submission process and infrastructure used for the real time evaluation as part of the U.S.
+Centers for Disease Control (CDC) COVID-19 Forecast Hub consortium.
 Second, we describe the internal evaluation used to demonstrate our model enhancements improve accuracy over a naive compartmental model.
 
- \subsection{Real-Time Forecast Evaluation}
-We began submitting forecasts to the U.S.
-Centers for Disease Control for incident deaths on May 10th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
+ \subsection{CDC Forecast Evaluation}
+We began submitting forecasts to the CDC for incident deaths on May 10th 2020 and have since submitted forecasts every Monday from then until August 1st 2020.
 The forecasts use daily data up to and including the Sunday before submission the next Monday.
 The one week ahead forecast corresponds to the following Saturday, the two week ahead to the second following Saturday and so on.
-We use the model submissions made in real-time evaluated on both MAE and coverage probability for submissions made on 2020-05-04, 2020-05-11, 2020-05-18, 2020-05-25, 2020-06-01, 2020-06-08, 2020-06-15,2020-06-22, 2020-06-29, 2020-07-06, 2020-07-13, 2020-07-20, and 2020-07-27.
-Note that not all targets are observed at all weeks.
-This is due to 4 week ahead targets for weeks 2020-07-13 and beyond not being observable by 2020-08-01.
+We use the model submissions made in real-time evaluated on both MAE and coverage probability for 13 submissions.
 We subset to the 50 states and Washington D.C., which is the largest set of locations where forecasts were made for each date.
@@
 
- We compare the results against the COVID-19 Forecast Hub baseline model.
+ We compare the results against the CDC baseline model.
 This is a purely statistical model that uses historical COVID-19 deaths only.
@@
  \begin{itemize}
- \item \textbf{MechBayes Full} Mech Bayes as submitted to the CDC.
+ \item \textbf{MechBayes Case/Death Time-Varying} Mech Bayes as submitted to the CDC.
 That is, a model using negative binomial observation noise as well as a time-varying random walk, using a joint likelihood over cases and deaths.
 
- \item \textbf{MechBayes Fixed case and death deviation model} MechBayes Full with $p_{c,t}$ fixed to $p_c$, that is, removing the time-varying case and death deviation model.
+ \item \textbf{MechBayes Case/Death Fixed} MechBayes Case/Death Time-Varying with $p_{c,t}$ fixed to $p_c$, that is, removing the time-varying case and death deviation model.
 
- \item \textbf{MechBayes Death Only} MechBayes Full with observations on cases removed.
+ \item \textbf{MechBayes Death Fixed} MechBayes Case/Death Time-Varying with observations on cases removed.
  \end{itemize}
 
- Note that these are nested models, with MechBayes Death Only contained in MechBayes Fixed case and death deviation model contained in MechBayes Full.
+ Note that these are nested models, with MechBayes Death Fixed contained in MechBayes Case/Death Fixed contained in MechBayes Case/Death Time-Varying.
 
@@
 This is because such a model would assume no interventions were put in place, which clearly violates the data-generating process.
-Previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential} \cite{abbott2020estimating}\cite{flaxman2020estimating} \cite{smirnova2019forecasting}.
+Previous Covid-19 modeling attempts have established that time-varying transmissibility is essential \cite{pei2020differential,abbott2020estimating,flaxman2020estimating,smirnova2019forecasting}.
 
@@
 
-    \includegraphics[scale=.2]{fit_and_forecast_results.png}
+    \includegraphics[scale=.15]{fig3.png}
 
-\caption{Example fit and forecast for 2020-07-26 for states with over 50 incident deaths and the U.S..
+\caption{\textbf{A,B} Example fit and forecast for 2020-07-19 for four selected states.
 Grey bands represent 95\% prediction intervals.
 Blue line represents median forecast.
-MechBayes is able to produce well calibrated fits to the data as well as accurately tracking trends in incident deaths.}
+MechBayes is able to produce well calibrated fits to the data as well as accurately tracking trends in incident deaths.
+\textbf{B} Posterior of $\beta_t$ for each of the four states, with 95\% credible intervals in blue.
+Note that this parameter does not reflect interventions specifically, but rather any change in transmissibility.
+\textbf{C} Posterior of the case-death discrepancy for the four example states.
+Here we can see that over time, the number of observed cases that resulted in deaths increased.
+This does not necessarily reflect testing, but an overall increase in the ability to observe cases as compared to the predicted number of cases one would obtain through the case-fatality ratio.
+}
 \label{fig:fit_and_forecast_results}
@@
 
-\begin{figure}
-  \centering
-     \begin{subfigure}{1\textwidth}
-  \centering
-    \includegraphics[scale=.15]{beta_t_plot.png}
-    \caption{Time-varying beta ($\beta(t)$).}
-\end{subfigure}
-
-\begin{subfigure}{1\textwidth}
-  \centering
-    \includegraphics[scale=.15]{detection_plot.png}
-    \caption{Time-varying case and death deviation model ($p_{c,t}$)}
-\end{subfigure} 
- \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S.
+% 
+%\begin{figure}
+%  \centering
+%     \begin{subfigure}{1\textwidth}
+%  \centering
+%    \includegraphics[scale=.15]{beta_t_plot.png}
+%    \caption{Time-varying beta ($\beta(t)$).}
+%\end{subfigure}
+%
+%\begin{subfigure}{1\textwidth}
+%  \centering
+%    \includegraphics[scale=.15]{detection_plot.png}
+%    \caption{Time-varying case and death deviation model ($p_{c,t}$)}
+%\end{subfigure} 
+% \caption{  A) Time varying beta parameter for all states with a day exceeding 50 incident deaths, including the U.S.
 by 2020-07-26.
@@
 The time varying case and death deviation model is able account for the increase in testing and other reporting anomolies.}
-\label{fig:model_details}
-\end{figure}
+%\label{fig:model_details}
+%\end{figure}
+\section{Results}
 
-\section{Real-time Forecasting Results}
+\subsection{Model Fitting and Inference}
 
-We can see from Figure \ref{fig:fit_and_forecast_results} that MechBayes is able to accurately model the observed data.
+MechBayes is able to accurately model the observed data, adapting to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves (Figure \ref{fig:fit_and_forecast_results} A).
 The model is able to adapt to highly variable incident death reporting, variable transmission rates, and overall heterogeneity of incidence curves.
@@
 Figure \ref{fig:fit_and_forecast_results}  shows prediction intervals at the 95\% level, with 92.3\% of observations falling within the bounds for each state.
-However, we can also see that in some states, such as California and Florida, the model is biased high, with all observations outside of the 95\% prediction interval falling below.
+However, we can also see that in some states, such as California and Florida, the model is biased high, with almost all observations outside of the 95\% prediction interval falling below.
 Figure \ref{fig:fit_and_forecast_results}  also shows 4 weeks of daily forecasts, along with the daily observed incidence for 1 week out.
-We can see that the predictions are tracking the data even under the weekly reporting cycles.
+We can see that the predictions are tracking the data even under the highly variable weekly reporting fluctuations.
 
-We can see from Figure \ref{fig:model_details} that MechBayes is able to learn to adapt to the evolving pandemic situation.
-Panel A shows the time-varying transmission $\beta(t)$ for 20 states with the highest total deaths.
-We can see that $\beta(0)$ is centered on our prior but as data comes in, the estimate increases.
-This is especially true in NY, where the epidemic took off quickly in March.
-However, the model is then able to adjust to the varying levels of interventions present in each state.
-This radically reduces the transmissibility parameter by the first week of April 2020 for almost all states.
-This is consistent with a peak in overall deaths two weeks later in mid-April 2020.
-There seems to be some estimation issues at the boundary of $\beta(t)$ where the number of cases does not match the number of deaths, since the cases have not yet converted to deaths.
-This results in the model underestimating transmissibility to reduce the flow through the compartments.
+MechBayes is able to adapt to changes in transmission through the $\beta(t)$ parameter (Figure \ref{fig:fit_and_forecast_results} C).
+Transmission was high across all four example states in March, with exponential growth rates shown in panel B.
+Transmission then slowed significantly in April across each states.
+While New York remained low through July, Texas, California and Florida saw increases in late June.
+This is reflected in the increase in incident deaths approximately two weeks later in mid-July.
 
-We can also see from Figure \ref{fig:model_details}  that MechBayes is able to account for the drastic increase in testing that has occurred across the U.S.
-since March 2020.
-Panel B also shows that our prior estimate of 30\% of cases being detected, may have been too high, as all regions show a dip in case and death deviation model before climbing again.
+MechBayes is able to adapt to changes in the case-fatality ratio through the $p_{c,t}$ parameter (Figure \ref{fig:fit_and_forecast_results} D).
+At the start of the pandemic, only those who were very sick were able to be tested.
+This lead to an unusually high case-fatality ratio (with respect to estimates from other countries).
+MechBayes was able to increase the number of latent cases (by a factor of $p_{c,t}$) to lower the estimated case-fatality ratio and avoid extreme over estimates of incident death predictions that would have accompanied such a drastic increase in cases in March.
+The model was able to adapt to this change across all four example states.
 
@@
 
-\caption{Scores from COVID-19 Forecast Hub broken down by target and forecast week.
+\caption{Scores from CDC COVID-19 forecast initiative broken down by target and forecast week.
+In Panels A and B, each observation represents mean absolute error (MAE) and empirical prediction interval coverage rate averaged across all targets and locations for the given week in which forecasts were made.
 MechBayes improves in average MAE in 9 out of 12 timepoints.
@@
 \end{subfigure}
-\caption{Scores from COVID-19 Forecast Hub broken down by region for the 20 regions with the highest incident deaths.
+\caption{Scores from CDC COVID-19 forecast initiative broken down by region for the 20 regions with the highest incident deaths.
+Each observation represents mean absolute error (MAE) or empirical prediction interval coverage rate averaged across all targets and forecast dates for the given region in which forecasts were made.
 Regions are sorted by total deaths left to right.
@@
 
-\section{Real-Time Model Results}
+\subsection{CDC Forecast Results}
 
-We next turn to the comparison of MechBayes against the COVID-19 Forecast Hub baseline model.
+We next turn to the comparison of MechBayes against the CDC baseline model.
 This baseline model uses the previous daily incident as the mean forecast for the current daily incidence, along with  bootstrapped prediction intervals from historical changes in daily incidence.
-See COVID-19 Forecast Hub for more details.
-\cite{covidhub}.
+See CDC COVID-19 Forecast Hub for more details \cite{covidhub}.
 
-We begin by breaking down the results by week the forecast was made (forecast week) and averaging over both region and target.
-As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes outperformed the COVID-19 Forecast Hub baseline on MAE and coverage probability when broken down by forecast week.
-However, the improvements broken down by time are small, and improves in 9 out of 12 evaluation weeks.
+Overall, MechBayes had an MAE of 34.1, when averaging over all regions, forecast dates, and targets.
+The CDC baseline model had an MAE of 53.2.
+Similarly, the empirical coverage probability at the 95\% level for MechBayes was 94.6\%, compared to 93.2\% for the baseline model.
+
+ We also break down the results by week the forecast was made (forecast week) and averaging over both region and target.
+As we can see from Figure \ref{fig:covidhub} (A,B), MechBayes outperformed the CDC baseline on MAE and coverage probability when broken down by forecast week.
+While the improvements broken down by time are small, there is an improvment in 9 out of 12 evaluation weeks in MAE.
 As we can see from \ref{fig:data}, in weeks with the largest increase in deaths (mostly during the month of May) MechBayes significantly outperformed the baseline model.
@@
 Out of the states with the 10 highest death counts, only in California did the baseline model outperform MechBayes on the average MAE.
-However, the results are more mixed for states with lower death counts.
-
 We break down the results by target by averaging over region and forecast week  \ref{fig:covidhub} (E,F) Here we can see uniform improvement over the baseline model by MechBayes in terms of MAE and coverage probability.
@@
 Finally, we include a formal test of the difference in MAE to demonstrate statistically significant advantages in using MechBayes over the baseline model.
+MechBayes has statistically significantly lower MAE for 1-4 weeks ahead at the 95\% level (coefficients in Appendix 2).
+The difference in skill seems to improve as the forecast horizon increases.
+For example, if the baseline model MAE for a particular 1 step ahead forecast is 30, one would expect the MechBayes forecast to have an MAE of 24.8.
+If the baseline model MAE for a particular 4 step ahead forecast is 30, one would expect the MechBayes forecast to have an MAE of 22.5.
+We demonstrate that the mixed model proposed above fits the MAE results reasonably in Appendix 2.
+
+ \subsection{Ablation Test Results}
+
+MechBayes Case/Death Time-Varying is consistently better than MechBayes Death Fixed or MechBayes Case/Death Fixed.
+The only exception seems to be June 22nd 2020 for the two week ahead target.
+The difference in MAE between MechBayes Case/Death Time-Varying and the competing models increases as forecast horizon increases.
+This suggest that MechBayes Case/Death Time-Varying is a better long term forecasting model.
+
+We can also see that the MechBayes Death Fixed is consistently better than MechBayes Case/Death Fixed.
+This may be evidence that naively including case data, without adjusting for discrepancies between cases and deaths due to testing etc., may be worse than not including it at all.
+
+\begin{figure}
+  \centering
+     \begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.15]{ablation_1.png}
+    \caption{1 week ahead.}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.15]{ablation_2.png}
+    \caption{2 week ahead.}
+\end{subfigure}
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.15]{ablation_3.png}
+    \caption{3 week ahead.}
+\end{subfigure}%
+\begin{subfigure}{.5\textwidth}
+  \centering
+    \includegraphics[scale=.15]{ablation_4.png}
+    \caption{4 week ahead. }
+\end{subfigure}
+
+\caption{Ablation test MAE broken down by forecast week and target.
+We can see that MechBayes Case/Death Time-Varying performs better than MechBayes Case/Death Fxied and MechBayes Death Fixed in almost all breakdowns.
+We can also see that the improvement becomes more pronounced at larger horizons, suggesting that MechBayes Case/Death Time-Varying is a better longer term forecasting model.
+}
+\label{fig:ablation}
+\end{figure}
+
+\section{Discussion}
+
+MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world data challenges during a pandemic.
+This model produced consistently accurate real-time forecasts over the course of 3 months, and was ranked as one of the top 3 of 11 models on an independent scoreboard of COVID-19 forecast models \cite{yyg}.
+Our experiments led us to the following conclusions about the performance of this model and the underlying methodology.
+
+\begin{itemize}
+
+\item \textbf{MechBayes is accurate when compared to a baseline model}.
+As Figure \ref{fig:covidhub} shows, MechBayes improved relative to the baseline model when broken down by timezero and target.
+The results are more pronounced when breaking down by target, where the boxplots are almost completely separated.
+These results are statistically significant for 1-4 weeks ahead, with larger improvements at later horizons, suggesting that the biggest improvements from MechByes over the baseline model comes in longer term forecasting.
+Additionally, as seen in Figure \ref{fig:covidhub_region_results} , the biggest gains in performance in MAE occur in regions with the largest total deaths counts.
+This is a desirable feature of a pandemic forecasting model.
+
+\item \textbf{MechBayes is probabilistically well-calibrated}.
+As we can see from Figure \ref{fig:covidhub} (B,D), the coverage probability of MechBayes is much closer to 95\%.
+When averaging over target, region, and timezero the coverage probability is 94.6\%.
+This suggests that MechBayes is very well calibrated at the 95\% level.
+The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise.
+The model is able to learn reasonable estimates of these variance parameters to produce well calibrated results.
+
+\item \textbf{Adding case data when predicting deaths is helpful but only when accounting for data quality issues}.
+Our ablation tests (Figure \ref{fig:ablation}) shows that allowing for a time-varying case and death deviation model is a key feature for reducing MAE of forecasts.
+MechBayes Case/Death time-varying both incorporates incident cases into the model likelihood and allows for a flexible deviation between cases and deaths is consistently more accurate than a model that does not account for cases at all ("Death Fixed" model) and a model that does account for cases but does not account for a time-varying deviation between cases and deaths.
+
+%As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the case and death deviation model at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
+There is remarkable similarity between the increase across geographical locations as well, suggesting (as testing data reflects) an overall increase in the number of detected cases across the U.S.
+Adjusting for case data quality also requires a QA process, where data dumps are redistributed across a specific time-window.
+This avoids the situation where MechBayes interprets data dumps as a drastic change in disease dynamics.
+This is visible in Figure \ref{fig:model_details}, where the time-varying case and death deviation model in New Jersey follows the typical trend of other states, until the data dump, where the case and death deviation model dips significantly down.
+
+%*\item \textbf{MechBayes is biased high}.
+We can see from Figure \ref{fig:results_discussion} that MechBayes was biased high with respect to the median forecast.
+We suspect this is due to the inherent exponential growth of an SEIR model when the number of susceptibles is large.
+Since overall prevalence in any state is well below the number needed to reach herd immunity, when the model sees an increase in cases it treats this as exponential growth, which translates into exponential growth in deaths.
+However, in the recent forecasts the bias is consistently lower than the COVID-19 Forecast Hub baseline model.
+Note the extreme low bias in the first submission is due to the omission of the U.S.
+in the submission.
+Late June 2020 is when many states in the U.S.
+began to see an uptick in cases again, with Texas, Florida, and California seeing their largest total case counts since the beginning of the pandemic.
+Since MechBayes is conditional on the current level of interventions, forecasts made from June 22nd assumed the same level of intervention present on June 22nd.
+During this time, each state was re-opening various establishments, while cases were rising.
+MechBayes Case/Death Time-Varying forecasted an exponential growth in cases, which was reflected in a large over prediction two weeks later on July 4th.
+*%
+
+\item \textbf{Most epidemiological parameters are unidentifiable from the data}.
+MechBayes requires relatively informative priors on $\sigma$, $\gamma$, $\rho$ and $\lambda$.
+These parameters reflect the biology of the disease, from latent incubation period to average time from symptom onset to death.
+Since the data used to fit MechBayes is only a time series of confirmed cases and deaths, these parameters are simply not identifiable from the data.
+MechBayes was built for forecasting, not inference on epidemiological parameters.
+However, using a Bayesian framework we can simultaneously set priors based on the literature and allow for small deviations from the prior due to variations across states.
+
+\item \textbf{Allowing for time-varying transmissibility is necessary to non-parametrically capture the effect of  interventions}.
+Our ablation test explicitly did not include a model that fixed $\beta$ across time.
+This is because the model would not converge without the flexibility to capture changes in transmission.
+While non-parametrically modeling interventions is appealing from a forecasting perspective, it does modify the philosophy behind compartmental modeling.
+By including such a flexible parameter, we may view MechBayes as simply a random-walk model, with a set of epidemiological parameters transforming that random-walk in an almost deterministic way to match both cases and deaths.
+For instance, if the variance of the random walk $\sigma_{\beta}^2$ was allowed to be arbitrarily large, then $\beta(t)$ could vary enough to match the data exactly.
+This would clearly attribute reporting issues as true changes in transmissibility.
+Bypassing the epidemiological interpretation that compartmental models provide.
+
+\end{itemize}
+
+%\begin{figure}
+%\centering
+%\includegraphics[scale=.15]{bias_by_timezero.png}
+%\caption{Bias of MechBayes and COVID-Baseline as a function of time.  }
+%\label{fig:results_discussion}
+%\end{figure}
+
+\section{Conclusion}
+
+We have seen that MechBayes is a powerful Bayesian compartmental model that can capture the real-world complexities of forecasting during a pandemic.
+Through real-time and retrospective evaluation, we have demonstrated the success of MechBayes in forecasting COVID-19.
+The model is able to improve over a naive baseline model as well as a naive compartmental model.
+Allowing for time-varying interventions and case and death deviation model is a necessary model component during a real-time pandemic forecasting effort.
+
+While we chose an exponential random walk on $\beta(t)$ there are many other choices for flexible non-parametric modeling of transmissibility.
+Further work might consider a spline model, or a Gaussian process, or semi-parametric models capable of taking intervention dates as covariates.
+
+However, using relatively simple non-parametric methods we were able to beat a baseline model in both point and probabilistic forecast evaluations.
+Our ablation tests show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.
+
+\newpage
+
+\section{Appendix}
+\subsection{Statistical Test for Significance}
 In order to do this, we just a random effects regression model of the form,
@@
 
-We can see that MechBayes is significantly better for 1-4 weeks ahead.
-The difference in skill seems to improve as the forecast horizon increases.
-The model estimates $\beta_{h_4mb}$ = -0.29, which translates to a multiplicative difference of $e^{-.29}=.75$ on the MAE scale.
-For example, if the basline model MAE for a particular 4 step ahead forecast is 30, one would expect the MechBayes forecast to have an MAE of 22.5.
-We demonstrate that the mixed model proposed above fits the MAE results reasonably in Appendix 2.
-
-   \section{Ablation Test Results}
-
-We can see from Figure \ref{fig:ablation} that MechBayes Full is consistently better than MechBayes Death Only or MechBayes Fixed case and death deviation model.
-The only exception seems to be June 22nd 2020 for the two week ahead target.
-Late June 2020 is when many states in the U.S.
-began to see an uptick in cases again, with Texas, Florida, and California seeing their largest total case counts since the beginning of the pandemic.
-Since MechBayes is conditional on the current level of interventions, forecasts made from June 22nd assumed the same level of intervention present on June 22nd.
-During this time, each state was re-opening various establishments, while cases were rising.
-MechBayes Full forecasted an exponential growth in cases, which was reflected in a large over prediction two weeks later on July 4th.
-Finally, we can also see that the difference in MAE between MechBayes Full and the competing models increases as forecast horizon increases.
-This suggest that MechBayes Full is a better long term forecasting model.
-
-We can also see that the MechBayes Death Only is consistently better than MechBayes Fixed case and death deviation model.
-This may be evidence that naively including case data, without adjusting for discrepancies between cases and deaths due to testing etc., may be worse than not including it at all.
-
-\begin{figure}
-  \centering
-     \begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.15]{ablation_1.png}
-    \caption{1 week ahead.}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.15]{ablation_2.png}
-    \caption{2 week ahead.}
-\end{subfigure}
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.15]{ablation_3.png}
-    \caption{3 week ahead.}
-\end{subfigure}%
-\begin{subfigure}{.5\textwidth}
-  \centering
-    \includegraphics[scale=.15]{ablation_4.png}
-    \caption{4 week ahead. }
-\end{subfigure}
-
-\caption{Ablation test MAE broken down by forecast week and target.
-We can see that MechBayes Full performs better than MechBayes Fixed Case/Death Deviation and MechBayes Only death in almost all breakdowns.
-We can also see that the improvement becomes more pronounced at larger horizons, suggesting that MechBayes Full is a better longer term forecasting model.
-}
-\label{fig:ablation}
-\end{figure}
-
-\section{Discussion}
-
-MechBayes is a fast, fully Bayesian compartmental model capable of accounting for real-world data challenges during a pandemic.
-This model produced consistently accurate real-time forecasts over the course of 3 months, and was ranked as one of the top 3 of 11 models on an independent scoreboard of COVID-19 forecast models \cite{yyg}.
-Our experiments led us to the following conclusions about the performance of this model and the underlying methodology.
-
-\begin{itemize}
-
-\item \textbf{MechBayes is accurate when compared to a baseline model}.
-As Figure \ref{fig:covidhub} shows, MechBayes improved relative to the baseline model when broken down by timezero and target.
-The results are more pronounced when breaking down by target, where the boxplots (95\%) are almost completely separated.
-These results are statistically significant for 1-4 weeks ahead, with larger improvements at later horizons, suggesting that the biggest improvements from MechByes over the baseline model comes in longer term forecasting.
-Additionally, as seen in Figure \ref{fig:covidhub_region_results} , the biggest gains in performance in MAE occur in regions with the largest total deaths counts.
-This is a desirable feature of a pandemic forecasting model.
-
-\item \textbf{MechBayes is probabilistically well-calibrated compared with the baseline model}.
-As we can see from Figure \ref{fig:covidhub} (B,D), the coverage probability of MechBayes is much closer to 95\%.
-When averaging over target, region, and timezero the coverage probability is 94.6\%.
-This suggests that MechBayes is very well calibrated at the 95\% level.
-The two sources of uncertainty in MechBayes come from the distribution over the differential equation parameters and the observation noise.
-The model is able to learn reasonable estimates of these variance parameters to produce well calibrated results.
-
-\item \textbf{Adding case data when predicting deaths is helpful but only when accounting for data quality issues}.
-Our ablation tests (Figure \ref{fig:ablation}) clearly shows that time-varying case and death deviation model is a key feature in the model for reducing MAE of forecasts.
-The "Full" model that both incorporates incident cases into the model likelihood and allows for a flexible deviation between cases and deaths is consistently more accurate than a model that does not account for cases at all ("Death Only" model) and a model that does account for cases but does not account for a time-varying deviation between cases and deaths.
-
-%As we can see in Figure \ref{fig:model_details}, the model relies on this parameter heavily, setting the case and death deviation model at around 15\% at the start of the pandemic in March, to nearly 80\% by August 1st 2020.
-There is remarkable similarity between the increase across geographical locations as well, suggesting (as testing data reflects) an overall increase in the number of detected cases across the U.S.
-Adjusting for case data quality also requires a QA process, where data dumps are redistributed across a specific time-window.
-This avoids the situation where MechBayes interprets data dumps as a drastic change in disease dynamics.
-This is visible in Figure \ref{fig:model_details}, where the time-varying case and death deviation model in New Jersey follows the typical trend of other states, until the data dump, where the case and death deviation model dips significantly down.
-
-\item \textbf{MechBayes is biased high}.
-We can see from Figure \ref{fig:results_discussion} that MechBayes was biased high with respect to the median forecast.
-We suspect this is due to the inherent exponential growth of an SEIR model when the number of susceptibles is small.
-Since overall prevalence in any state is well below the number needed to reach herd immunity, when the model sees an increase in cases it treats this as exponential growth, which translates into exponential growth in deaths.
-However, in the recent forecasts the bias is consistently lower than the COVID-19 Forecast Hub baseline model.
-Note the extreme low bias in the first submission is due to the omission of the U.S.
-in the submission.
-\textbf{is this true}
-
-\item \textbf{Most epidemiological parameters are unidentifiable from the data}.
-MechBayes requires relatively informative priors on $\sigma$, $\gamma$, $\rho$ and $\lambda$.
-These parameters reflect the biology of the disease, from latent incubation period to average time from symptom onset to death.
-Since the data used to fit MechBayes is only a time series of confirmed cases and deaths, these parameters are simply not identifiable from the data.
-However, using a Bayesian framework we can simultaneously set priors based on the literature and allow for small deviations from the prior due to variations across states.
-
-\item \textbf{Allowing for time-varying transmissibility is necessary to non-parametrically capture the effect of  interventions}.
-Our ablation test explicitly did not include a model that fixed $\beta$ across time.
-This is because the model would not converge without the flexibility to capture changes in transmission.
-While non-parametrically modeling interventions is appealing from a forecasting perspective, it does modify the philosophy behind compartmental modeling.
-By including such a flexible parameter, we may view MechBayes as simply a random-walk model, with a set of epidemiological parameters transforming that random-walk in an almost deterministic way to match both cases and deaths.
-For instance, if the variance of the random walk $\sigma_{\beta}^2$ was allowed to be arbitrarily large, then $\beta(t)$ could vary enough to match the data exactly.
-This would clearly attribute reporting issues as true changes in transmissibility.
-Bypassing the epidemiological interpretation that compartmental models provide.
-
-\end{itemize}
-
-\begin{figure}
-\centering
-\includegraphics[scale=.15]{bias_by_timezero.png}
-\caption{Bias of MechBayes and COVID-Baseline as a function of time.  }
-\label{fig:results_discussion}
-\end{figure}
-
-\section{Conclusion}
-
-We have seen that MechBayes is a powerful Bayesian compartmental model that can capture the real-world complexities of forecasting during a pandemic.
-Through real-time and retrospective evaluation, we have demonstrated the success of MechBayes in forecasting COVID-19.
-The model is able to improve over a naive baseline model as well as a naive compartmental model.
-Allowing for time-varying interventions and case and death deviation model is a necessary model component during a real-time pandemic forecasting effort.
-
-While we chose an exponential random walk on $\beta(t)$ there are many other choices for flexible non-parametric modeling of transmissibility.
-Further work might consider a spline model, or a Gaussian process, or semi-parametric models capable of taking intervention dates as covariates.
-
-However, using relatively simple non-parametric methods we were able to beat the COVID-19 Forecast Hub baseline model in both point and probabilistic forecast evaluations.
-Our ablation tests show that extending the basic SEIR compartmental to real-world pandemic challenges improves forecasting accuracy.
-
-\newpage
-
-\section{Appendix}
-\subsection{A1}
 \subsection{Seeding Epidemic}
@@
 Finally, our prior on the rate at which dying patients succumb  satisfies $\lambda$ $\E[\lambda] = 0.1$ with shape 10 corresponding to roughly 10 days in the $D_{1}$ compartment.
-\subsection{A2}
-\textbf{is this needed}
 
@@
 \newpage
+
 \bibliographystyle{unsrt}
-\bibliography{bib}
+\bibliography{mechbayes}
 

### 2020-10-20 ablation.R

@@ -6,7 +6,7 @@ library(ggthemes)
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(6),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
+levels(ablation_res_subset$model) <- c("Case/Death Time-Varying","Case/Death Fixed","Death Fixed")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
   
 ablation_1 <- ggplot(ablation_res_subset %>% group_by(forecast_date,model) %>% 
@@ -16,12 +16,12 @@ ggsave("/Users/gcgibson/mech_bayes_paper/ablation_1.png",ablation_1,device="png"
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(13),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
+levels(ablation_res_subset$model) <- c("Case/Death Time-Varying","Case/Death Fixed","Death Fixed")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
+levels(ablation_res_subset$model) <- c("Case/Death Time-Varying","Case/Death Fixed","Death Fixed")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
@@ -35,7 +35,7 @@ ggsave("/Users/gcgibson/mech_bayes_paper/ablation_2.png",ablation_2,device="png"
 
 ablation_res_subset <- ablation_res[ablation_res$horizon %in% c(20),]
 ablation_res_subset$model <- as.factor(ablation_res_subset$model)
-levels(ablation_res_subset$model) <- c("Full","Fixed Case/Death Deviation","Death Only")
+levels(ablation_res_subset$model) <- c("Case/Death Time-Varying","Case/Death Fixed","Death Fixed")
 ablation_res_subset$forecast_date <- as.Date(ablation_res_subset$forecast_date)
 
 
[…]

### 2020-10-20 analysis.R

@@ -4,29 +4,33 @@ library(cowplot) #for manuscript ready figures
 library(lme4) #for lmer & glmer models
 library(sjPlot) #for plotting lmer and glmer mods
 library(sjmisc) 
+library(latex2exp)
+
 #library(effects)
 library(sjstats) #use for r2 functions
 
 np <- import("numpy")
-files = list.files("/Users/gcgibson/covid-intervention-analysis/samples/")
+files = list.files("/Users/gcgibson/mech_bayes_paper/samples/")
+#files <- unlist(lapply(files,function(x){return (substr(x,1,2))}))
 files <- state_deaths_subset_idx
 files <- files[files !="US"]
 files <- state.abb[match(files,state.name)]
 
 
-beta <- matrix(NA,ncol=length(files),nrow=137 )
-beta_upper <- matrix(NA,ncol=length(files),nrow=137 )
-beta_lower <- matrix(NA,ncol=length(files),nrow=137 )
+beta <- matrix(NA,ncol=length(files),nrow=137+28)
+
+beta_upper <- matrix(NA,ncol=length(files),nrow=137+28)
+beta_lower <- matrix(NA,ncol=length(files),nrow=137+28 )
 library(matrixStats)
 col_idx <- 1
 states = c()
 for (file in files){
   if (file !="DC.npz"){
     print (file)
-    npz1 <- np$load(paste0("/Users/gcgibson/covid-intervention-analysis/samples/",file,".npz"),allow_pickle = T)
-    beta[,col_idx] = colMeans(npz1$f$mcmc_samples[[1]]$beta)
-    beta_upper[,col_idx] = t(rowQuantiles(t(npz1$f$mcmc_samples[[1]]$beta),probs = .975))
-    beta_lower[,col_idx] = t(rowQuantiles(t(npz1$f$mcmc_samples[[1]]$beta),probs = .025))
+    npz1 <- np$load(paste0("/Users/gcgibson/mech_bayes_paper/samples/",file,".npz"),allow_pickle = T)
+    beta[,col_idx] = colMeans(cbind(npz1$f$mcmc_samples[[1]]$beta,npz1$f$forecast_samples[[1]]$beta))
+    beta_upper[,col_idx] = t(rowQuantiles(t(cbind(npz1$f$mcmc_samples[[1]]$beta,npz1$f$forecast_samples[[1]]$beta)),probs = .975))
[…]

### 2020-10-20 binomial_sir.R

@@ -1,0 +1,132 @@
+library(ggplot2)
+
+pop <- 1000
+
+
+
+a <- matrix(1,nrow=pop,ncol=pop)
+diag(a) <- 0
+
+## random graph
+for (i in 1:pop){
+  for (j in 1:pop){
+    delete <- rbinom(1,1,0 )
+    if (delete == 1 & i !=j){
+      a[i,j] <- 0
+      a[j,i] <- 0 
+    }
+  }
+}
+
+
+#a <- matrix(0,nrow=pop,ncol=pop)
+
+#for (j in 110:210){
+ # a[j,] <- 1
+#  a[,j] <-1
+#}
+diag(a) <- 0
+### SIR MODEL
+
+
+sir <- function(time, pop, I_init, beta,gamma,delta_t,on_graph) {
+  infected_ids <-c(1:I_init)
+  recovered_ids <- c()
+  susceptible_ids <- c((I_init+1):pop)
+ sir_matrix <- matrix(NA,nrow=time,ncol=4)
+ sir_matrix[1,] <- c(pop-I_init,I_init,0,I_init)
+ for (t in 2:time){
+   if ( !on_graph){
+    new_infections <- mean(rpois(n=10000, sir_matrix[t-1,1]*beta*sir_matrix[t-1,2]/pop))
+    new_recoveries <- mean(rpois(n=10000,sir_matrix[t-1,2]*gamma))
+   } else{
+     new_infections <- 0
+     infected_ids_frozen <- infected_ids
+     for (i in infected_ids_frozen){
+      # print (infected_ids)
+       connected <- a[,i]
+       connected_ids <- which(a[,i]==1)
+       connected_sus_ids <- susceptible_ids[susceptible_ids %in% connected_ids]
+       
+       if (length(connected_sus_ids) >0 ){
+         succesful_infections <- rbinom(1,size=length(connected_sus_ids),prob=delta_t*beta/pop)
+         if (length(connected_sus_ids) > succesful_infections ){
+          sus_to_be_infected <- sample(connected_sus_ids,succesful_infections)
+         } else{
+           sus_to_be_infected <- connected_sus_ids
+         }
+       } else{
+         sus_to_be_infected <- c()
+       }
+       
+       if (length(sus_to_be_infected) > 0){
+         infected_ids <- c(infected_ids,sus_to_be_infected)
[…]

### 2020-10-20 data.R

@@ -26,6 +26,8 @@ state_deaths_subset$total_deaths <- factor(state_deaths_subset$total_deaths,leve
 state_deaths_subset$location_name <- as.factor(state_deaths_subset$location_name)
 state_deaths_subset <- state_deaths_subset[order(state_deaths_subset$total_deaths),]
 state_deaths_subset$location_name <- factor(state_deaths_subset$location_name,levels = unique(state_deaths_subset$location_name))
+
+state_deaths_subset[state_deaths_subset$location_name == "New Jersey" & state_deaths_subset$date == "2020-07-26",]$incident <- 1887
 data_plot <- ggplot(state_deaths_subset,aes(x=date,y=incident)) + geom_point(size=.5) + facet_wrap(~location_name,scales="free") + theme_bw() +ylab("Incident Reported COVID-19 Deaths (per day)") + xlab("")
 #data_plot <- data_plot + ylab("Incident Deaths") + theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
 ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",width=10,height=10)
@@ -34,8 +36,8 @@ ggsave("/Users/gcgibson/mech_bayes_paper/data_plot.png",data_plot,device="png",w
 
 library(reticulate)
 np <- import("numpy")
-files = list.files("/Users/gcgibson/mech_bayes_paper/samples/samples")
-fit_plot_df <- array(NA,dim=c(length(files),3,172) )
+files = list.files("/Users/gcgibson/mech_bayes_paper/samples/samples/")
+fit_plot_df <- array(NA,dim=c(length(files),3,165) )
 col_idx <- 1
 states = c()
 for (file in files){
@@ -62,10 +64,10 @@ for (file in files){
   
 }
 
-plot_df <- data.frame(time=rep(seq(as.Date("2020-03-05"),as.Date("2020-08-23"),by="day"),length(states)),
+plot_df <- data.frame(time=rep(seq(as.Date("2020-03-05"),as.Date("2020-08-16"),by="day"),length(states)),
                       median=c(t(fit_plot_df[,1,])),upper=c(t(fit_plot_df[,2,])),
                       lower=c(t(fit_plot_df[,3,])))
-plot_df$location <- rep(states,each=172)
+plot_df$location <- rep(states,each=165)
[…]
